---
title: 
- Statistical Methods in Cancer Epidemiology \newline using R
author: 
- \textbf{Karri Seppä}  \newline \newline
- Finnish Cancer Registry  \newline  \newline
- Lecture 4
date: "Mar 2, 2020"
institute: "karri.seppa@cancer.fi"
output:
  beamer_presentation:
    includes:  
         in_header: beamer_header.tex
editor_options: 
  chunk_output_type: console
---

## Time splitting

Contents:

1. Basic concepts of time to event analysis
\medskip
2. Piecewise constant hazards model and \newline
age-specific incidence rates
\medskip
3. Splitting follow-up times by age
\medskip
4. Accounting for current age in rate ratio estimation 
\bigskip

Main R functions to be covered, all in \texttt{Epi} package 

- \texttt{Lexis()} 
- \texttt{splitLexis()}
- \texttt{timeBand()} 


## Time to event analysis

Analysis of incidences = analysis of \textit{times to event} or \textit{failure times} or \textit{survival times} (censored).

Mathematical concepts:
\begin{align*}
    T & =  \mbox{ time to outcome event --  random variable,} \\
 S(t) & =  P(T>t) = \mbox{ {\bf survival} function of $T$,} \\
     & =  \mbox{ probability of avoiding 
the event up to given time }t,\\
\lambda(t)  = & -S'(t)/S(t) = 
\mbox{ {\bf intensity} or {\bf hazard} function}, \\
\Lambda(t) & =  \int_0^t \lambda(u)du = - \log S(t)
   = \mbox{\bf cumulative hazard}, \\
  F(t) & =  1-S(t) = 1-\exp\{ -\Lambda(t) \} 
  = \mbox{{\bf risk} function} \\
    & =  \mbox{ probability of the outcome to occur by }t   \\
    & =  \mbox{ cumulative distribution function of }T.
\end{align*}


## Hazard rate or intensity function

Can be viewed as \textit{theoretical incidence rate}. Formally
$$ \lambda(t) = \lim_{\Delta \rightarrow 0}
 \frac{P(t < T \leq t+\Delta \mid T > t)}{\Delta} $$
\begin{enumerate}[$\approx$]
\item Probability of failure occurring 
in a short interval \\
 $]t, t+ \Delta]$, given ``survival'' or
avoidance of event 
 up to its start $t$,  divided by the interval length.
\end{enumerate}
This is equivalent to saying that over this short interval
$$ 
\mbox{risk } \approx  \mbox{ rate }\times 
\mbox{ length of interval} 
$$
or $\qquad P(t < T \leq t+\Delta \mid T > t) \ 
 \approx \ \lambda(t)\times \Delta$.


## Exponential or constant hazard model

Simplest probability model for time to event: \newline
\textbf{Exponential distribution}, Exp($\lambda$), in which
$$
 \lambda(t) = \lambda \mbox{  (constant)}\quad 
\Rightarrow\quad 
 \Lambda(t) = - \log S(t) = \lambda t
$$
Analysis of failure data of $n$ individuals. 
For subject $i$ let
\begin{align*}
y_i & = \mbox{ time to event or 
time to censoring,}\quad
  Y = \sum y_i \\
d_i & = \mbox{ indicator for 
observing the event,}\quad
  D = \sum d_i
\end{align*}
Exp($\lambda$) model $\Rightarrow$ 
\textbf{Likelihood function} of $\lambda$ is
$$ L(\lambda) = \prod_{i=1}^n \lambda(y_i)^{d_i} S(y_i)
     =  \prod_{i=1}^n \lambda^{d_i} e^{-\lambda y_i}
             \ = \ \exp( D \log \lambda -\lambda Y) $$

## Constant rate -- Poisson model

This is actually equivalent to the \textit{Poisson-likelihood},
\textit{i.e.} likelihood of $\lambda$
assuming that the number of cases 
$D$ is distributed according to the \textbf{Poisson
distribution} with expected value $\lambda Y$.

\bigskip
With randomly censored exponential times 
$D$ is only \newline
approximately Poisson. This is sufficient, though,
for likelihood-based (& asymptotic frequentist) inference.
 
 \bigskip
Solving the \textit{score equation}: \ $d \log L(\lambda)/d\lambda = 0$ \newline
$\rightarrow$ \textbf{maximum likelihood estimator} (MLE) of $\lambda$ is
$$
 \widehat\lambda = \frac{D}{Y}  = 
  \frac{\mbox{number of cases}}
  {\mbox{total person-time}} 
  = \mbox{ empirical incidence rate!}
$$


## Time to event -- when to start the clock?

Incidence can be studied on various time scales, \textit{e.g.}
\begin{itemize}
\item age (starting point = birth),
\item exposure time (first exposure),
\item follow-up time (entry to study),
\item duration of disease (diagnosis).
\end{itemize}
Age is usully the strongest time-dependent 
determinant of health 
outcomes.

\medskip
 Age is also often correlated with
duration of ``chronic'' exposure (\textit{e.g.} years of smoking).

\medskip
Therefore, adjustment for \textit{current age} is needed rather than
for \textit{age at entry} to follow-up (like in clinical survival studies).


## Age to event split into agebands

Let $T$ = age at which outcome event occurs. \newline
Parametric form of $\lambda(t)$, hazard by age -- usually unknown.

\medskip
\textbf{Piecewise exponential model} or \textbf{piecewise constant hazards' model} -- an approximation for $\lambda(t)$:
$$ \lambda(t) = \lambda_k, \qquad t \in \ ]a_{k-1}, a_k],\quad \Delta_k = a_k - a_{k-1}, $$
where cutpoints $0 = a_0 < a_1 < \dots < a_K$ divide the age range into
disjoint \textbf{agebands}, each with constant rate.

\medskip
In chronic disease epidemiology agebands with 
$\Delta_k = 5$ years (0-4, 5-9, $\dots$, 80-84)
or 10 years are commonly used.


## Age-specific incidence rates

For empirical estimation of rates we calculate 
in each ageband
\begin{align*}
 D_k & = \mbox{ number of cases occurring in ageband } k, \\
 Y_k & = \sum_{i=1}^n y_{ik} = \mbox{ total person-time in ageband } k,
\end{align*}
where $y_{ik}$ is the time slot that subject $i$ spends
in ageband $k$ out of his/her whole \textbf{follow-up
time} (from \textbf{entry} to \textbf{exit}).

ML estimators of  $\lambda_1, \dots, \lambda_K$: 
\textbf{age-specific incidence rates}
$$ \widehat\lambda_k = I_k = {D_k}/{Y_k}, \quad k = 1, \dots, K $$
based on log-likelihood 
$\ \ \log L = \sum_{k} (D_k \log \lambda_k - \lambda_k Y_k)$.


## Cumulative rates \& risks

In this model, the cumulative hazard and risk functions are
\begin{align*}
 \Lambda (t) & = \sum_{a_j < t} \lambda_j \Delta_j + \lambda_k(t - a_{k-1}),\qquad t \in \ ]a_{k-1}, a_k] \\
 F(t) & = 1 - S(t) = 1 - \exp\{ - \Lambda(t) \}, 
\end{align*}
the latter assuming that no \textbf{competing risks} are present.

\medskip
Estimation: Plug in empirical rates $\widehat\lambda_j = D_j/Y_j$ to get
the cumulative rate $C$ and incidence proportion $R$ by $t$:
\begin{align*}
  C & = \widehat{\Lambda} (t) =  \sum_{a_j < t} \widehat\lambda_j \Delta_j + \widehat\lambda_k (t - a_{k-1}),\qquad t \in \ ]a_{k-1}, a_k] \\
  R & = \widehat{F}(t)  = 1 - \widehat{S}(t) = 1 - \exp\{ - \widehat{\Lambda}(t) \}
\end{align*}

## Example: Follow-up of a small geriatric cohort

\setlength{\unitlength}{1.2pt}
\begin{center}
\includegraphics[width=10cm,height=6cm ]{lifelines}
\end{center}
No's of cases/p-years \& rates (/100 y) in 5-y agebands: 
$$ {1}/{21} = 4.8, \quad {1}/{16}= 6.2, \quad {2}/{16.5} = 12.1$$

<!---\bes
D_1 & = & 1, \quad D_2 =  1, \quad D_3  =  2 \\
Y_1 & = & 14, \ \ Y_2  = 17, \ \  Y_3  =  14 \mbox{ years}\\
I_1 & = & 7.1, \ \ I_2  = 5.9, \ \ I_3  =  14.3 \mbox{ per 100 y} 
\ees -->

## Splitting follow-up by \texttt{Lexis()} in package \texttt{Epi}

Individual ages at entry and at exit, as well as outcomes
are assigned into vectors and stored in a data frame \texttt{coh}:  
\small
```{r, comment=NA,echo=T,warning=FALSE}
ag.entry <- c(69, 70, 70.5, 71, 72, 76.9, 81, 81.9)
ag.exit <- c(83.5, 74.5, 76, 77, 85, 82, 84, 85)
event <- c(0,1,0,1,0,1,1,0) ; ind <- 1:8 
coh <- data.frame( ind, ag.entry, ag.exit, event)
```
\normalsize
Function \texttt{Lexis()} specifies the
time scale(s) to be considered. It 
 creates an enriched data frame
belonging to class \texttt{Lexis}.
\small
```{r, comment=NA,echo=T,warning=FALSE,results='hide'}
library(Epi)
coh.L <- Lexis(entry = list(age = ag.entry), 
               exit = list(age = ag.exit), 
               exit.status = event, data = coh, id = ind)
```
\normalsize


## Data frame of class \texttt{Lexis}

\scriptsize
```{r, comment=NA,echo=T,warning=FALSE}
coh.L
```
\normalsize
Interpretation of new columns
\begin{align*}
 \texttt{age} & =  \text{age at entry to follow-up},\\
 \texttt{lex.dur} & =  \text{duration of follow-up},\\
 \texttt{lex.Cst} & =  \text{current status at entry}, \\
  \texttt{lex.Xst} & =  \text{status at exit from follow-up}.
\end{align*}


## Splitting follow-up times by agebands

\medskip
Function \texttt{splitLexis()} splits individual follow-up times
into given agebands and expands the data frame.
\small
```{r, comment=NA,echo=T,warning=FALSE}
coh.A <- splitLexis(coh.L, 
       br = c(70,75,80,85), time.scale="age")
```
\scriptsize
```{r, comment=NA,echo=T,warning=FALSE}
coh.A[-c(8:12),]
```
\normalsize


## Splitted \texttt{Lexis} object

- Function \texttt{splitLexis()} expanded the original data frame such that
for all cohort members one or more rows were created, one for
each ageband into which a subject contributes person time.
\medskip
- Ex: Subject 1 has been under follow-up in all agebands considered, but subjects 7 and 8 only in 80$-<85$ y.
\medskip
- Function \texttt{timeBand()} 
converts variable \texttt{age} into 
factor \texttt{ageband}. Also, shorthand
names for person-time slots and occurrence of outcome event are given.
\small
```{r, comment=NA,echo=T,warning=FALSE}
coh.A$ageband <- timeBand(coh.A, "age", "factor") 
coh.A$y_ik <- coh.A$lex.dur # person-time slot
coh.A$d_ik <- coh.A$lex.Xst # occurrence of outcome
```
\normalsize


## Splitted \texttt{Lexis} object (cont'd)

:::::: {.columns}

::: {.column width="40%"}
\scriptsize
```{r, comment=NA,echo=T}
coh.A[, c(1,10:12)] 
```
\normalsize
:::

::: {.column width="60%"}
<!---Interpretation of variable names in \texttt{lexis}:-->
\begin{enumerate}[ ]
\item
\item
\item
\item \verb|lex.id| = subject \\ $\quad$index 
in original \\ $\quad$data frame,
\item \verb|ageband| = ageband \\
            $\quad$ and  its limits, 
\item \verb|y_ik| = person-time slot 
        \\ $\quad$ 
       spent in ageband
\item \verb|d_ik| = indicator for \\
   $\quad$ event occurring \\
   $\quad$ in ageband.
\end{enumerate}
:::
::::::

 \bigskip
Subject 1's follow-up time ($14.5\mbox{ y} = 1+ 5 + 5 + 3.5\mbox{ y}$)
is split into 4 agebands,
$\dots,$ subject 8 contributes only to 1 ageband.


## Tabulation of cases, rates etc. by ageband

Event indicators \& person-time slots
are summed over the rows of the split-expanded data
frame in categories of \texttt{ageband}: 
\small
```{r, comment=NA,echo=T}
D <- with(coh.A, tapply(d_ik, ageband, sum))
Y <- with(coh.A, tapply(y_ik, ageband, sum))
```
\normalsize
Incidence rates ($I$), cumulative rates ($C$) and incidence proportions ($R$), 
the latter two by the end of each \texttt{ageband}:
\small 
```{r, comment=NA,echo=T}
I <- 100*D/Y; C <- cumsum((D/Y)*5); R <- 1-exp(-C)
tab <- round(cbind(D,Y,I,C,R),3)[2:4, ]
```
\normalsize


## Age standardised incidence rate

Direct age standardisation

- e.g. to the World Standard population

\scriptsize
```{r, comment=NA,echo=T,eval=F}
library(popEpi)
stdpop18
```
\vspace{-0.5cm}
```{r, comment=NA,echo=F}
library(popEpi)
stdpop18
```


## Age standardised incidence rate (cont'd)

Weighted average of age-specific incidence rates $I_k$
$$
\sum_k w_kI_k, \ \text{ where }\  \sum_k w_k = 1
$$
\scriptsize
```{r, comment=NA,echo=T}
tab <- cbind(tab,stdpop18[15:17,1:2])
tab$w <- tab$world/sum(tab$world)
tab
#age-standardised incidence per 100 person years
sum(tab$w*tab$I)
#non-standardised incidence per 100 person years
100*sum(tab$D)/sum(tab$Y)
```


## Age standardised incidence rate (cont'd)

$$
\text{SE}\left[\log\left(\sum_k w_kI_k\right)\right]\approx \frac{1}{\sum_k w_kI_k} \sqrt{\sum_k  \frac{w_k^2 d_k}{y_k^2}}
$$

\scriptsize
```{r, comment=NA,echo=T}
std.rate <- sum(tab$w*tab$D/tab$Y)
SE.std.rate <- sqrt( sum(tab$w^2*tab$D/tab$Y^2) )
exp( log(std.rate)+1.96*(1/std.rate)*SE.std.rate*c(-1,1) )
```


```{r, comment=NA,echo=T}
rate(tab,obs="D",pyrs="Y",adjust="agegroup",weights=tab$w)[,c(3,5:6)]
```

\normalsize
- Age-standardised incidence in age group 70-84 is:
    - 6.2 per 100 person years, 95% CI (2.1, 18.6)




## Example: The Diet Study (see C\&H)

A cohort of 337 men in three occupational groups in England, \newline
aged 30 to 67 y at entry, recruited in '50s and '60s, \newline 
followed-up until mid '70s for incidence of CHD events. 

\bigskip
Risk factors of interest, measured by dietary survey at entry.
\begin{align*}
\mbox{\texttt{energy}}  = & \mbox{ total energy intake (kcal/d)}, \\
\mbox{\texttt{energy.grp}}  = & \mbox{\texttt{ energy} dichotomized:} \\ 
  { }  & \mbox{ 1: ``$<=$2750 KCals'', 2: ``$>$2750 KCals''},\\ 
\mbox{\texttt{fat}}  = & \mbox{ fat intake (g/d)}, \\
\mbox{\texttt{fibre}}  = & \mbox{ dietary fibre intake (g/d),} \\
       {  } & \mbox{\texttt{ height, weight, bmi, }}etc.
\end{align*}


## Important dates and outcome event

The data set \texttt{diet} in \texttt{Epi} contains three dates: 
\begin{align*}
\mbox{\texttt{dob}} = & \mbox{ date of \textbf{birth}}, \\
\mbox{\texttt{doe}} = & \mbox{ date of \textbf{entry} into follow-up}, \\
\mbox{\texttt{dox}} = & \mbox{ date of \textbf{exit}, end of follow-up}.
\end{align*}
These are given in format \texttt{yyyy-mm-dd} but implicitly stored as 
\textit{number of days since 1.1.1970}. 

In addition, the outcome event is represented by
\begin{align*}
\mbox{\texttt{chd}} = & \mbox{ indicator for \textbf{status} at exit:} \\ 
{ } & \quad \mbox{ 1 = CHD event occurred, 0 = censored} . 
\end{align*}


## Data \texttt{diet}: creating a \texttt{Lexis} object

First convert all dates into fractional calendar years
using function \texttt{cal.yr()} in \texttt{Epi} 
\small
```{r, comment=NA,echo=T}
data(diet)
diet <- transform(diet, doe = cal.yr(doe), 
                  dox = cal.yr(dox), dob = cal.yr(dob) )
```
\normalsize
Convert the data frame into a \texttt{Lexis} object.  
\small
```{r, comment=NA,echo=T,results='hide'}
dietL <- Lexis( entry = list(age = doe-dob),
                exit = list(age = dox-dob),
                exit.status = chd, data = diet )
```
\normalsize
In the nexty step the \texttt{Lexis} object is splitted
according to 3 agebands (y): $30-<50$, $50-<60$, $60-<70$


## Splitting the \texttt{Lexis} object into agebands

\small
```{r, comment=NA,echo=T,results='hide'}
dietA <- splitLexis(dietL, br = c(30,50,60,70), 
                   time.scale = "age")
dietA$ageband <- timeBand(dietA, "age", "factor")
dietA$y_ik <- dietA$lex.dur ; dietA$d_ik <- dietA$lex.Xst
```
\scriptsize
```{r, comment=NA,echo=F}
dietA$dob <- round(dietA$dob,1)
dietA$doe <- round(dietA$doe,1)
dietA$dox <- round(dietA$dox,1)
dietA$y <- round(dietA$y,1)
dietA$age <- round(dietA$age,1)
dietA$y_ik <- round(dietA$y_ik,1)
```

```{r, comment=NA,echo=T}
dietA[1:8,c("id","dob","doe","dox","y","chd","energy.grp","ageband",
            "age","y_ik","d_ik")]
```
\small
Properties of the original data frame and the expanded object:
\scriptsize
```{r, comment=NA,echo=T,eval=F}
> str(diet) 
'data.frame':   337 obs. of  17 variables:
> str(dietA)
Classes Lexis and data.frame  729 obs. of 25 variables
```
\normalsize


## Relevelling of \texttt{energy.grp} and some tabulations

The \texttt{energy.grp} variable is relevelled such that
"high energy" is taken as the reference or
"unexposed" category and "low energy" as the "exposed" one.
 \small
```{r, comment=NA,echo=T,results='hide'}
dietA$eg2 <- Relevel( dietA$energy.grp, 
           ref = ">2750 KCals" )
```
\normalsize
Tabulation of cases, person-years and rates: \small
```{r, comment=NA,echo=T,results='hide'}
tab.ae <- stat.table( list( ageband, eg2),
      list( D = sum(d_ik), Y = sum(y_ik),
            I = ratio(d_ik, y_ik, 1000) ),
       margin = T, data = dietA )
print(tab.ae, digits= c(sum=0, ratio=1))
```
\normalsize


## Rates by ageband and energy intake

:::::: {.columns}

::: {.column width="50%"}
\scriptsize
```latex
 ------------------------------------ 
            -----------eg2----------- 
 ageband       >2750  <=2750   Total  
               KCals   KCals          
 ------------------------------------ 
 (-Inf,30]        NA      NA      NA  
  ...                                     
 (30,50]           4       2       6  
                 622     381    1003  
                 6.4     5.2     6.0                                      
 (50,60]           6      12      18  
                1128     979    2107  
                 5.3    12.3     8.5                                      
 (60,70]           8      14      22  
                 794     699    1493  
                10.1    20.0    14.7                                      
 (70,Inf]         NA      NA      NA
 ...                                      
 Total            18      28      46  
                2544    2059    4604  
                 7.1    13.6    10.0  
 ------------------------------------ 
```
\normalsize
:::

::: {.column width="50%"}

\small
Crude rate ratio\vspace{-0.3cm}
```{r, comment=NA,echo=T}
tab.ae[3, 6, 2]/
  tab.ae[ 3, 6, 1]
```

\medskip
Rate ratios by ageband:\vspace{-0.3cm} 
```{r, comment=NA,echo=T}
IRs <- tab.ae[3, 2:4, 2]/
  tab.ae[3, 2:4, 1]       
round(IRs,2)
```
\begin{itemize}
\item Low intake risky?
\item No effect in young?
\end{itemize}
\normalsize
:::
::::::

## Poisson model on age and exposure

Let $D_{kj}, Y_{kj}$, and $I_{kj}$ be cases, p-years \& rate
in ageband $k$ \& exposure category $j$ 
(1="unexposed", 2="exposed"). \newline
Piecewise Exp-model in both exposure categories assumed:
$$ 
\lambda_{kj} = \mbox{ theoretical rate in cell }kj . 
$$
Theoretical rate ratio $\rho_k = \lambda_{k2}/\lambda_{k1}$, \newline comparing 
exposed \textit{vs.} unexposed.

(a) What are the ``true'' values of $\rho_k$?
(b) Can we assume $\rho_k = \rho$, same rate ratio in all agebands?
(c) What is the value of the common rate ratio $\rho$?


## Poisson model (cont'd)

Assuming common rate ratio the true rates are modelled
$$ 
\log \lambda_{kj} = \alpha_k + \beta_j 
= \sum_{k=1}^K \alpha_k A_k + \sum_{j=1}^2 \beta_j X_j , 
$$
where $A_k$ and $X_j$ are indicator (1/0) variables for level $k$
of ageband and level $j$ of exposure.
In exponential form
$$ 
\lambda_{kj} = \exp( \alpha_k + \beta_j ) 
= e^{\alpha_k} e^{\beta_j} . 
$$
Set $\beta_1 = 0$ ("unexposed" as reference)
 $\Rightarrow$ Interpretation:
\begin{align*}
&\alpha_k =  \log(\lambda_{k1})
 = \mbox{ log-rate of unexposed in ageband }k\\
&\beta_2  =   \log(\lambda_{k2}/\lambda_{k1}) = \log(\rho)  
 = \mbox{ log-common rate ratio}
\end{align*}


## Fitting the Poisson model

Use function \texttt{glm()} on the expanded data frame:
```{r, comment=NA,echo=F}
dietA <- splitLexis(dietL, br = c(30,50,60,70), 
                   time.scale = "age")
dietA$ageband <- timeBand(dietA, "age", "factor")
dietA$y_ik <- dietA$lex.dur ; dietA$d_ik <- dietA$lex.Xst
dietA$eg2 <- Relevel( dietA$energy.grp, 
           ref = ">2750 KCals" )
```

```{r, comment=NA,echo=T,warning=F}
m.ea <- glm( d_ik/y_ik ~ ageband + eg2, 
             fam = poisson, w = y_ik, data = dietA )
```
\small
```{r, comment=NA,echo=T}
round(ci.lin(m.ea, Exp=T)[ , -(3:4)], 4 )
```
\normalsize
The estimated rate ratio for "low" vs. "high"
energy consumption, adjusted for age, is thus
1.87 [1.03 to 3.38], only slightly lower than the 
unadjusted one 1.92 [1.06 to 3.47].


## Concluding remarks

- Modelling could continue from this to include other confounders,
  continuous covariates, interactions, \textit{etc.} 
 \medskip
- Agebands may well be much narrower than in our example. With infinitely
narrow bands Poisson regression equals the famous Cox model.
\medskip
- Splitting by many time scales (\textit{e.g.} age, calendar time, time since first exposure, \textit{etc.}) simultaneously and the corresponding data frame expansion is straightforward using these tools. More about this 
in the next lecture. 


