---
title: 
- Statistical Methods in Cancer Epidemiology \newline using R
author: 
- \textbf{Karri Seppä}  \newline \newline
- Finnish Cancer Registry  \newline  \newline
- Lecture 5
date: "Mar 2, 2020"
institute: "karri.seppa@cancer.fi"
output:
  beamer_presentation:
    includes:  
         in_header: beamer_header.tex
editor_options: 
  chunk_output_type: console
---

## More than one time scale and SIR & SMR

Contents

1. Special exposure cohorts 
\medskip
2. Observed and expected cases, SIR \& SMR
\medskip
3. Lexis diagram and life lines
\medskip
4. Splitting follow-up time simultaneously by age and period
\medskip
5. Merging reference rates with the cohort data
and performing SIR/SMR computations
\bigskip

Main R functions to be covered

- \texttt{Lexis.diagram()} and other \texttt{Lexis} tools in \texttt{Epi}  
- \texttt{merge()}
- \texttt{sir()} and \texttt{sirspline()} in \texttt{popEpi}


## Special cohorts of exposed subjects

- Occupational cohorts, exposed to potentially \\
hazardous agents
\medskip
- Cohorts of patients on chronic medication, which may have harmful long-term side-effects

No internal comparison group of unexposed subjects.

\medskip
\textit{Question}:
Do incidence or mortality rates in the \newline \textit{exposed} target cohort differ from those of a
roughly comparable \textit{reference} population?

Reference rates obtained from: 

- population statistics (mortality rates)
- disease \& hospital discharge registers (incidence)


## Accounting for age distribution

- Compare rates in a study cohort with a standard set of
  age-specific rates from the reference population.
  \medskip
- Reference rates normally based on large numbers of cases, so
  they are assumed to be "known" without error.
  \medskip
- Calculate \textbf{expected} number of cases, $E$, if the standard
  age-specific rates had applied in our study cohort. 
 \medskip
-  Compare this with the
  \textbf{observed} number of cases, $D$, by the \textbf{standardized incidence ratio} SIR \newline
  (or standardized mortality ratio SMR)
  
$$ \text{SIR} = D/E, \qquad \text{SE}[\log(\text{SIR})] = 1/\sqrt{D} $$


## Example: HT and breast ca.

- A cohort of 974 women treated with hormone (replacement) therapy
 were followed up. 
- $D$ = 15 incident cases of breast cancer were
observed. 
- Person-years ($Y_a$) and reference rates 
($\lambda_a^*$, per 100000 y)
by age group ($a$) were:

\small
\begin{center}
\begin{tabular}{lrrr}
Age & $Y_a$   & $\lambda_a^*$ & $E_a$ \\ 
\hline
40--44 \rule{0em}{1em} &  975 & 113 & 1.10 \\
45--49 & 1079 & 162 & 1.75 \\
50--54 & 2161 & 151 & 3.26 \\ 
55--59 & 2793 & 183 & 5.11 \\
60--64 & 3096 & 179 & 5.54 \\
\hline 
$\sum$ \rule{0em}{1em} & & & 16.77
\end{tabular}
\end{center}

## Example: HT use and breast ca. (cont'd)

- "Expected" number of cases at ages 40--44:
\[ 975 \times \frac{113}{100\,000} = 1.10 \]
\medskip
- Total "expected" cases is $E = 16.77$
 \medskip 
- The SIR is $15/16.77 = 0.89$.
 \medskip 
- Error factor: $\exp(1.96\times \sqrt{1/15}) = 1.66$
 \medskip 
- 95\% confidence interval is:
 \[ 0.89 \mpydiv 1.66 = (0.54, 1.48) \]


## A statistical model for SIR

- The theoretical rates $\lambda_{ap}$ by age ($a$)
and calendar period ($p$) in the cohort are assumed to be proportional to the rates $\lambda^*_{ap}$ in
the reference population:
\[   \lambda_{ap} = \rho\times \lambda^*_{ap}  \]
$\rho$ = hazard ratio btw the cohort and the reference
pop'n. 
\medskip
- The population rates 
$\lambda^*_{ap}$ are assumed to be known.
\medskip
- Cohort data: numbers of cases $D_{ap}$ and p-years $Y_{ap}$ by $a$ge and $p$eriod are computed.
\medskip
- It can be shown that the likelihood of $\rho$ is of Poisson type, and the maximum likelihood estimator of $\rho$ is:
\[
  \widehat\rho = \frac{D}{\sum \lambda^*_{ap} Y_{ap} } =
  \frac{\mbox{Observed}}{\mbox{Expected}}
  = \text{SIR}
\]


## Example: The Welsh Nickel Workers' Study

- A cohort of 679 men working in nickel smelters in South Wales first employed 1903-25 (for details see \textbf{B\&D}). 
\medskip
- Outcomes of interest: deaths from nasal (ICD code 160) 
and lung cancer (ICD 162 and 163) during follow-up 1934-76. 
\medskip
- Outcome event indicator and basic time variables:
\begin{align*}
\texttt{icd}  = & \mbox{ code for cause of death, 0 if not yet dead}\\
\texttt{date.bth}  = & \mbox{ date of birth}\\
\texttt{date.in}  = & \mbox{ date of starting follow-up}\\
\texttt{date.out}  = & \mbox{ date when follow-up ended}
\end{align*}


## Example (cont'd)

- Interesting risk factors in the original data frame:
\begin{align*}
\mbox{\tt expos}  = & \mbox{ exposure index based on years employed in}\\
   { } & \mbox{ high-risk areas in the smelter by 1925}\\
   { } & \mbox{ }\rightarrow \mbox{ categorized version EXP} \\
%  & { } & 1\ : \mbox{ 0 y}\\
%  & { } & 2\ : \mbox{ 0.5 to 4.0 y}\\
%  & { } & 3\ : \mbox{ 4.5 to 8.0 y}\\
%  & { } & 4\ : \mbox{ 8.5 to 12.0 y}\\
%  & { } & 5\ : \mbox{ $\geq$ 12.5 y}\\
\mbox{\tt date.1st}  = & \mbox{ date when first employed }
     \rightarrow\mbox{ AFE}
\end{align*}
- Risk factors to be formed from original variables:
\begin{align*}
\mbox{\tt age.1st}  = & \mbox{ age when first employed }
     \rightarrow\mbox{ AFE} \\
\mbox{\tt year.1st}  = & \mbox{ year of first employment }
   \rightarrow \mbox{ YFE} \\
\mbox{\tt time.1st}  = & 
\mbox{ time since first exposure }\rightarrow\mbox{ TFE}
\end{align*}


## Lexis diagram \& 4 lifelines from the nickel cohort

Diagram invented by \textit{Wilhelm Lexis} (1837-1914), 
German mathematician and demographer, professor in
Tartu 1874-76.

\vspace{0.5cm}
:::::: {.columns}

::: {.column width="50%"}
\includegraphics[height=2in]{Lexis1}
:::

::: {.column width="50%"}

Individual lifelines run dia-\newline
gonally from 
a given\newline
 (age, time) starting point\newline
to an endpoint. 

\bigskip
Here the lines go from \newline
start of exposure till \newline
 the age and time of exit.
 
 \bigskip
 Mortality follow-up started in 1934.
:::
::::::


## Nickel cohort: All lifelines in the Lexis diagram



\vspace{0.5cm}
:::::: {.columns}

::: {.column width="50%"}
\includegraphics[height=2in]{Lexis2}
:::

::: {.column width="50%"}

\bigskip
\bigskip
Follow-up starts not until 1934 for all subjects.

- dot (red) \newline = lung ca. death, 
- circle (green)\newline  = censoring
:::
::::::

Function \texttt{splitLexis()} splits individual follow-up times into
rectangles defined by agebands and calendar periods. 



## Splitting follow-up by age \& calendar time

:::::: {.columns}

::: {.column width="50%"}
\textbf{from} the registration of:

- Entry,
- Exit,
- Failure status

 of the individuals in the cohort,
 and the definition of the scale by:
 
- \textbf{O}rigin
- \textbf{S}cale
- \textbf{C}utpoints
:::

::: {.column width="50%"}

\textbf{to} the table of:

- $D$ = events,
  \medskip
- $Y$ = person time,

 by age and period.
:::
::::::


## Expected numbers in practice

- From the records of age-period split \& expanded 
  cohort data:
  \begin{enumerate}[ ]
\item $\quad y_{i,ap}$ = person-time slot in a record defined by
\item $\quad a$ = ageband of the record
\item $\quad p$ = period of the record
\end{enumerate}
<!---\item sex = The sex of the record-->

\bigskip

- From the file containing the reference rates:
  \begin{enumerate}[ ]
\item $\quad \lambda_{ap}^*$ = age \& period specific rate
\item $\quad a$ = ageband of the population rate
\item $\quad p$ = period of the population rate
\end{enumerate}
<!---\item sex --- The sex of the population rate-->


## Expected numbers in practice (cont'd)

Population rates are matched up 
to the expanded cohort data, and
  expected numbers individually are computed as:
$$ 
e_{i,ap} = \lambda_{ap}^* \times y_{i,ap}  
$$
and these are eventually summed: $E = \sum e_{i,ap}$

\bigskip
Always two datasets are needed for SIR:

1. the \textit{cohort} data with follow-up information on its
  individual members. This must be split \& expanded to match with
2. the \textit{reference rate} data with age \& period specific rates in the chosen reference population.


## SMR-calculations in R using Lexis tools:
 
\large  
\color{blue} \mbox{1.} Read in the cohort data (Welsh Nickel Workers)
\color{black}
\normalsize

  \medskip
 and  convert the dates \texttt{dd/mm/yyyy} into ``decimal years'' 


\small
```{r, comment=NA,echo=T}
library(Epi)
nick <- read.table( "nickel.txt", header=T, as.is=T )
for (j in 4:7) nick[ , j] <-
   cal.yr( nick[ , j], format = "%d/%m/%Y" )
```
\normalsize

List the records for the 4 men in a previous Lexis diagram
\small
```{r, comment=NA,echo=T}
round(nick[11:14, ],2)
```
\normalsize


## 2. Reference rates in E \& W read in
\small
```{r, comment=NA,echo=T}
ewrates <- read.table("ewrates.txt",header=T)
```

8 first and last rows checked
\scriptsize
```{r, comment=NA,echo=T}
ewrates[c(1:8, 143:150), ]
```


## E \& W lung ca. death rates by age and period}
\small
```{r, comment=NA,echo=T,results='hide'}
tapply(ewrates$lung, list("age" = ewrates$age,
         "year" = ewrates$year),sum)
```
\scriptsize
```{r, comment=NA,echo=F}
tapply(ewrates$lung, list("age" = ewrates$age,
         "year" = ewrates$year),sum)
```



## 3. Creating and expanding the \texttt{Lexis} object}

The data frame converted to a \texttt{Lexis} object in \newline
two time scales: \texttt{year} (calendar time) and \texttt{age}: 

\small 
```{r, comment=NA,echo=T} 
nickL <- Lexis( entry = list( year = date.in ),
                 exit = list( year = date.out, 
                              age = date.out - date.bth ),
    exit.status = as.numeric( nick$icd %in% c(162, 163) ),
           data = nick )
```

The \texttt{Lexis} object jointly split by age and period. Agebands and period bands are named like in the \texttt{ewrates} file --
\texttt{"left"} means the lower cutpoint (1st year) of a band.
\small
```{r, comment=NA,echo=T} 
nickL.a <- splitLexis(nickL, "age", br=seq(10,85,5) )
nickL.ap<- splitLexis(nickL.a,"year",br=seq(1931,1986,5))
```


## Splitting follow up time by age and calendar year

\scriptsize
```{r, comment=NA,echo=T,eval=F} 
plot( nickL.ap, c("year","age"), col=c("black"), xlim=c(1931,1986) )
points( nickL.ap, c("year","age"), pch=c(NA,1)[nickL.ap$lex.Xst+1], col="red" )
```
\vspace{-1cm}
```{r, comment=NA,echo=F, dev.args=list(pointsize=15)} 
plot( nickL.ap, c("year","age"), col=c("black"), xlim=c(1931,1986) )
points( nickL.ap, c("year","age"), pch=c(NA,1)[nickL.ap$lex.Xst+1], col="red" )
```
\vspace{-0.7cm}
```{r, comment=NA,echo=T} 
nickL.ap$year <- timeBand(nickL.ap, "year", "left")
nickL.ap$age <- timeBand(nickL.ap, "age", "left")
```


## The expanded data frame viewed

\small
```{r, comment=NA,echo=T,results='hide'} 
dim(nickL.ap)  # 10-fold expansion!
```
\scriptsize
\vspace{-0.3cm} 
```{r, comment=NA,echo=F} 
dim(nickL.ap)
```

```{r, comment=NA,echo=T,results='hide'} 
round( subset( nickL.ap, lex.id %in% 13:14)
       [ , c(1:4,6,8,10,12,13)] ,2)
```
\scriptsize
\vspace{-0.3cm} 
```{r, comment=NA,echo=F} 
round( subset( nickL.ap, lex.id %in% 13:14)
       [ , c(1:4,6,8,10,12,13)] ,2)
```


## 4. Merging the cohort data with E\&W rates

\small
```{r, comment=NA,echo=T,results='hide'} 
nickLew.ap <- merge(nickL.ap, ewrates, 
        by = c("age", "year"))  # key columns
round(nickLew.ap[1:20, c(1:4,6:8,10,12,13,14) ],1) 
```
\scriptsize
\vspace{-0.5cm} 
```{r, comment=NA,echo=F} 
round(nickLew.ap[1:20, c(1:4,6:8,10,12,13,14) ],1) 
```



## 5. Calculation of observed and expected

Cases & person-time slots renamed, expectations $\lambda_{ap}^* y_{i,ap}$ of becoming a case computed,
and tables by $a$ and $p$ produced.

\small
```{r, comment=NA,echo=T} 
nickLew.ap <- transform( nickLew.ap, 
         d_iap = lex.Xst, y_iap = lex.dur, 
         e_iap = lex.dur * lung/1.0E6 )

Obs.lung <- with(nickLew.ap, tapply(d_iap,
  list("age" = age, "year" = year), sum))

Exp.lung <- with(nickLew.ap, tapply(e_iap,
  list("age" = age, "year" = year), sum))
```

```{r, comment=NA,echo=T,results='hide'} 
Obs.lung ; round(Exp.lung,3)
```



## Observed and expected numbers printed
\vspace{-0.3cm} 
\scriptsize
```{r, comment=NA,echo=F} 
Obs.lung[-c(1:2),]
```
\vspace{-0.4cm}
```{r, comment=NA,echo=F} 
round(Exp.lung,3)[-c(1:2),]
```

## 6. Calculation of SMR

We can sum either over individual time slots:
```{r, comment=NA,echo=T} 
D <- sum(nickLew.ap$d_iap) 
E <- sum(nickLew.ap$e_iap) 
```

or over the newly formed tables:
```{r, comment=NA,echo=T} 
D <- sum(Obs.lung, na.rm=T)
E <- sum(Exp.lung, na.rm=T)
```

\medskip
Either way, the calculation proceeds:

\small
```{r, comment=NA,echo=T} 
SMR <- D/E; SE <- 1/sqrt(D); EF <- exp(1.96*SE)
round(c(D, E, SMR, SMR/EF, SMR*EF), 2) 
```


## 6. Fit model for SMR

```{r, comment=NA,echo=T} 
m <- glm( d_iap ~ 1+offset(log(e_iap)),
 family=poisson, data=nickLew.ap )
round( ci.exp( m ), 2 ) 
```


## 6. Fit model for SMR by age group

\scriptsize
```{r, comment=NA,echo=T} 
nickLew.ap$agegrp <- cut(nickLew.ap$age,c(20,50,65,100),right=F)
m <- glm( d_iap ~ -1+agegrp+offset(log(e_iap)),
 family=poisson, data=nickLew.ap )
round( ci.exp( m ), 2 ) 
```



## 6. Fit model for SMR using \texttt{popEpi} package

\scriptsize
```{r, comment=NA,echo=T} 
library(popEpi)
nickLew.ap$ref.rate <- nickLew.ap$e_iap/nickLew.ap$y_iap
m <- sir( coh.data = nickLew.ap, coh.obs = 'd_iap', coh.pyrs = 'y_iap', 
           ref.data = nickLew.ap, ref.rate = 'ref.rate', 
           adjust = c('age', 'year'), print = 'agegrp', conf.type = "wald")
m
```

## 6. Fit model for SMR using \texttt{popEpi} package

\scriptsize
```{r, comment=NA,echo=T, fig.height=6,eval=F} 
m <- sirspline( 
  coh.data = nickLew.ap, coh.obs = 'd_iap', coh.pyrs = 'y_iap', 
  ref.data = nickLew.ap, ref.rate = 'ref.rate', 
  adjust = c('age', 'year'), spline = 'age', knots=5, subset = age>30)
plot(m, conf.int = F)
lines(m, conf.int = T, lty=2)
```
\vspace{-1.3cm}
```{r, comment=NA,echo=F, fig.height=7, dev.args=list(pointsize=15)} 
m <- sirspline( 
  coh.data = nickLew.ap, coh.obs = 'd_iap', coh.pyrs = 'y_iap', 
  ref.data = nickLew.ap, ref.rate = 'ref.rate', 
  adjust = c('age', 'year'), spline = 'age', knots=5, subset = age>30)
plot(m, ylab="SMR", conf.int = F, las=1)
lines(m, conf.int = T, lty=2)
```

## Concluding remarks


- If specific exposure factors exist that have variable values within
the target cohort, the estimation of rate ratios associated with them
 may be efficiently adjusted for age and calendar period by taking the 
 age- and period-specific expected number as the baseline in Poisson-modelling.
 \medskip
- Follow-up time could be split yet by another relevant time axis, 
like time passed since start of exposure, which may be taken as an explanatory variable when modelling the effects of exposure within a cohort.
\medskip
- The main challenge is to identify a sufficiently 
comparable reference population. The ``general'' population is rarely an ideal one. 



