---
title: 'Epidemiologic data analysis using R '
author:
- Faculty of Social Sciences, University of Tampere
- ' -- '
- Janne Pitkaniemi
date: "`r format(Sys.time(), '%d.%m.%Y')`"
output:
  word_document: default
  pdf_document: default
subtitle: Practicals 9
---


```{r p9_0, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, comment = NA)
setwd("C:/Users/janne.pitkaniemi/Projects/TRE2018")
```

```{r p9_1,eval=TRUE, echo=TRUE,comment=NA}
library(Epi)
library(mstate)
library(survival)
library(lubridate)
```


#Survival analysis: Oral cancer patients


## Description of the data


File \verb|oralca2.txt|, that you may
access from a url address to be given in the practical,
contains data from 338
patients having an oral squamous cell carcinoma diagnosed and treated
in one tertiary level oncological clinic in Finland since 1985, followed-up
for mortality until 31 December 2008.
The dataset contains the following variables:

\medskip
\begin{tabular}{lcl}
{\tt sex}         &=& sex, a factor with categories
    \verb|1 = "Female", 2 = "Male"|,    \\
{\tt age}         &=& age (years) at the date of diagnosing the cancer,  \\
% loc  &-& location of the tumour (factor):
% \verb|1 = " Other", 2 = " Tung"|, \\
{\tt stage}       &=& TNM stage of the tumour (factor): \verb|1 = "I", ..., 4 = "IV", 5 = "unkn"|, \\
{\tt time}       &=& follow-up time (in years) since diagnosis until death or censoring,\\
{\tt event}      &=& event ending the follow-up (numeric): \\
{ }         & & 0 = censoring alive, 1 = death from oral cancer, 2 = death from other causes. \\
\end{tabular}


## Loading the packages and the data


Load the R packages \texttt{Epi, mstate}, and \texttt{survival} needed in this exercise.

Read the datafile {\tt oralca2.txt} from a website, whose precise address  will be given in the practical, into an R data frame named \texttt{orca}. 

Look at the head, structure and the summary of the data frame. Using function \texttt{table()} count the numbers of censorings as well as deaths from oral cancer and other causes, respectively, from the \texttt{event} variable.

```{r p9_2, eval=TRUE,echo=TRUE,warnings=FALSE}
orca <- 
  read.csv("C:/Users/janne.pitkaniemi/Projects/TRE2018/oralca2.txt", 
           sep="")
head(orca)

```

## Estimate cumulative incidence using competing risks

1. Use KM-estimator for oral cancer and other deaths separately and estimate
5 year mortality 

```{r p9_3a, eval=TRUE,echo=TRUE,warnings=FALSE}
orca$suob <- Surv(orca$time, 1*(orca$event > 0) )
km1 <- survfit( suob ~ 1, data = orca)
par(mfcol=c(3,1))
plot(km1,main="Survival (S,KM)",xlab="time",ylab="Proportion")
plot(c(0,25),c(0,1),main="1-S (F,KM)",xlab="time",ylab="Proportion",pch="")
lines(km1$time,1-km1$surv,type="s")
lines(km1$time,1-km1$upper,type="s",lty=2)
lines(km1$time,1-km1$lower,type="s",lty=2)
plot(km1,main="Cumulative hazard",xlab="time",ylab="Cumulative hazard",fun="cumhaz")
```


```{r p9_3b, eval=TRUE,echo=TRUE,warnings=FALSE}
km1<- survfit( Surv( time, 1*(event==1)) ~ 1,
                 data = orca)
res<-summary(km1)
index<-which(floor(res$time)==5)[1]-1

cat("Probablity of dying to oral cancer before 5th year and 95%CI",
    1-res$surv[index],"(",
    1-res$lower[index],";",
    1-res$upper[index],") \n" )

```

```{r p9_4, eval=TRUE,echo=TRUE,warnings=FALSE}
km1<- survfit( Surv( time, 1*(event==2)) ~ 1,
                 data = orca)
res<-summary(km1)
index<-which(floor(res$time)==5)[1]-1

cat("Probablity of dying to oral cancer before 5th year and 95%CI",
    1-res$surv[index],"(",
    1-res$lower[index],";",
    1-res$upper[index],") \n" )

```


## Estimate cumulative incidence using competing risks

1. Type help(plotCIF) and look at the help of plotting AJ-estimator for CIF. Then plot CIF for oral cancer data for deaths due to cancer and other causes.

```{r p9_5, eval=TRUE,echo=TRUE,warnings=FALSE}
par(mfrow=c(1,2))
cif<- survfit( Surv( time, event, type="mstate") ~ 1,
                 data = orca)
 plotCIF(cif, 1, main = "Oral cancer death",
    col=1, ylim = c(0, 0.7) )
 plotCIF(cif, 2, main= "Other death",
    col=1, ylim = c(0, 0.7) )
```

Print summary of cif and find 5-year CIF for oral and other cause death

```{r p9_6, eval=TRUE,echo=TRUE,warnings=FALSE}

res<-summary(cif)
index<-which(floor(res$time)==5)[1]-1
probs<-res$pstate[index,]

cat("Probablity of dying to oral cancer before 5th year and 95%CI",
    probs[1],"(",
    res$lower[index,1],";",
    res$upper[index,1],") \n" )
cat("Probablity of dying to other causes before 5th year and 95%CI",
    probs[2],"(",
    res$lower[index,2],";",
    res$upper[index,2],")" )

```





2. Type help(stackedCIF) for plotting stacked version of the CIF for both causes of death.

```{r p9_7, eval=TRUE,echo=TRUE,warnings=FALSE}
par(mfrow=c(1,1))
stackedCIF(cif, colour = c("gray80", "gray90"),
           main = "", xlab="Time (years)",xlim=c(0,25) )	
```

Add text indicating areas under the curve 

```{r p9_8, eval=TRUE,echo=TRUE,warnings=FALSE}
par(mfrow=c(1,1))
stackedCIF(cif, colour = c("gray80", "gray90"),
           main = "", xlab="Time (years)",xlim=c(0,25) )
text( 15, 0.15, "Oral cancer", pos = 2)
text( 15, 0.5, "Other causes", pos = 2)
```

Add reference lines for 5-year mortalities.

```{r p9_9, eval=TRUE,echo=TRUE,warnings=FALSE}

par(mfrow=c(1,1))
stackedCIF(cif, colour = c("gray80", "gray90"),
           main = "", xlab="Time (years)",xlim=c(0,25) )
text( 15, 0.15, "Oral cancer", pos = 2)
text( 15, 0.5, "Other causes", pos = 2)

res<-summary(cif)
index<-which(floor(res$time)==5)[1]-1
probs<-res$pstate[index,]
segments(5, 0, 5, probs[1], col= 'blue',lwd=2)
segments(5, probs[1], 5, probs[1]+probs[2], col= 'red',lwd=2)

segments(-1, probs[1], 5, probs[1], col= 'blue',lwd=2,lty=2)
segments(-1, probs[1]+probs[2], 5, probs[1]+probs[2], col= 'red',lwd=2,lty=2)

```


3. Plot stacked CIF for males and females for oral and all cause mortality. 


```{r p9_10, eval=TRUE,echo=TRUE,warnings=FALSE}
orca$nsex<-1*(orca$sex=="Male")

par(mfrow=c(1,2))
cif<- survfit( Surv( time, event, type="mstate") ~ nsex,
                 data = orca)
stackedCIF(cif, group=1, colour = c("gray80", "gray90"),
           main = "Male", xlab="Time (years)" )	
text( 15, 0.15, "Oral cancer", pos = 2)
text( 15, 0.5, "Other", pos = 2)

stackedCIF(cif, group=2, colour = c("gray80", "gray90"),
           main = "Female", xlab="Time (years)" )	
text( 15, 0.15, "Oral cancer", pos = 2)
text( 15, 0.5, "Other", pos = 2)

```


4. Explore the effect of gender to death causes using cox model and report hazard ratio for males vs females.

```{r, eval=TRUE,echo=TRUE,warnings=FALSE}
cat("Oral cancer mortality")
ci.exp(coxph(Surv( time, event==1) ~ nsex,  data = orca))
cat("Other cause mortality")
ci.exp(coxph(Surv( time, event==2) ~ nsex,  data = orca))
```

5. a. Look at oral cancer and other cause mortality by stage

```{r, eval=TRUE,echo=TRUE,warnings=FALSE}
par(mfrow=c(2,2))
cif<- survfit( Surv( time, event, type="mstate") ~ stage,
               data = orca)

stackedCIF(cif, group=1, colour = c("gray80", "gray90"),
           main = "stage I", xlab="Time (years)" ) 
text( 15, 0.15, "Oral cancer", pos = 2)
text( 15, 0.5, "Other", pos = 2)

stackedCIF(cif, group=2, colour = c("gray80", "gray90"),
           main = "stage II", xlab="Time (years)" )   
text( 15, 0.15, "Oral cancer", pos = 2)
text( 15, 0.5, "Other", pos = 2)

stackedCIF(cif, group=3, colour = c("gray80", "gray90"),
           main = "stage III", xlab="Time (years)" )   
text( 15, 0.15, "Oral cancer", pos = 2)
text( 15, 0.5, "Other", pos = 2)

stackedCIF(cif, group=4, colour = c("gray80", "gray90"),
           main = "stage IV", xlab="Time (years)" )   
text( 15, 0.15, "Oral cancer", pos = 2)
text( 15, 0.5, "Other", pos = 2)

cat("Oral cancer mortality")
ci.exp(coxph(Surv( time, event==1) ~ stage,  data = orca))
cat("Other cause mortality")
ci.exp(coxph(Surv( time, event==2) ~ stage,  data = orca))
```




b. Look at oral cancer and other cause mortality by the following age groups: (0,65] , (65,80] , (80,120]


```{r, eval=TRUE,echo=TRUE,warnings=FALSE}
orca$ageg<-cut(orca$age,breaks = c(0,65,80,120))

par(mfrow=c(2,2))
cif<- survfit( Surv( time, event, type="mstate") ~ ageg,
               data = orca)

stackedCIF(cif, group=1, colour = c("gray80", "gray90"),
           main = levels(orca$ageg)[1], xlab="Time (years)" ) 
text( 15, 0.15, "Oral cancer", pos = 2)
text( 15, 0.5, "Other", pos = 2)

stackedCIF(cif, group=2, colour = c("gray80", "gray90"),
           main = levels(orca$ageg)[2], xlab="Time (years)" )   
text( 15, 0.15, "Oral cancer", pos = 2)
text( 15, 0.5, "Other", pos = 2)

stackedCIF(cif, group=3, colour = c("gray80", "gray90"),
           main = levels(orca$ageg)[3], xlab="Time (years)" )   
text( 15, 0.15, "Oral cancer", pos = 2)
text( 15, 0.5, "Other", pos = 2)
```


## Time dependent effects- Heart transplant data

Look at the mortality related to the heart transplant using pre transplant time as refere and split the follow-up as in the lectures.
Essentially we are redoing the analysis in the lecture for practice. In this case we have predefined time intervals of mortality risk.


First a naive Cox analysis with age, year and surgery as covarites and treatment as the factor of interest.

```{r,eval=TRUE, echo=TRUE,comment=NA}
ci.lin(coxph(Surv(start, stop, event) ~
    age + year + surgery + transplant,
data=heart),Exp=TRUE)[,c(5:7)]
```

Use the data called jasa as the heart data, because this version of the data
consists also dates for accepting into the program (accept.dt), transplant date (tx.date),
and follow-up time (futime).


```{r,eval=TRUE,echo=TRUE,comment=NA}
heart.Lx <-
Lexis(entry=list(age.time=age, 
                 fu.time=0,
tx.time=decimal_date(tx.date)-decimal_date(accept.dt)),
duration=futime/365, 
exit.status=as.character(fustat),
data=subset(jasa, is.na(tx.date) | fu.date != tx.date))
```


```{r,eval=TRUE,echo=TRUE,comment=NA}
head(heart.Lx)
```

Next, in order to make the cut points for subject separately we create
three new variables for line 

```{r,eval=TRUE,echo=TRUE,comment=NA}

heart.Lx <- within(heart.Lx, {cut.0 <- tx.time; 
                              cut.1 <- tx.time+0.2;
                              cut.2 <- tx.time+0.4} );
```


```{r,eval=TRUE,echo=TRUE,comment=NA}
head(heart.Lx)
```


```{r, eval=TRUE,echo=TRUE,comment=NA}
heart.Lx.cut <- mcutLexis(heart.Lx, timescale="fu.time",
wh=c("cut.0","cut.1","cut.2"),
new.states=c("a","b","c"))
```

Look at the variable lex.Cst whic indicates the state at the entry to follow-up period.

```{r,eval=TRUE,echo=TRUE,comment=NA}
table(heart.Lx.cut$lex.Cst)
```

Do the cox regression using lex.Cst as factor covariate in the analysis.


```{r,eval=TRUE,echo=TRUE,comment=NA}
print(format(as.data.frame(ci.exp(coxph( 
      Surv(fu.time, fu.time + lex.dur, lex.Xst==1)
      ~ factor(lex.Cst),data=heart.Lx.cut))),digits=3)) 
```

If we want to adjust for the age at the biginning of the follow-up period we
need the add age.time in the analysis.Let look at subject with lex.id =66.

```{r,eval=TRUE,echo=TRUE,comment=NA}
 heart.Lx.cut[heart.Lx.cut$lex.id==66,]
```


```{r,eval=TRUE,echo=TRUE,comment=NA}
print(format(as.data.frame(ci.exp(coxph( 
      Surv(fu.time, fu.time + lex.dur, lex.Xst==1)
      ~ factor(lex.Cst)+age.time,data=heart.Lx.cut))),digits=5)) 
```

Use some other variable that you find interesting and try to add it as a covariate of stratify 
the analysis.
