---
title: 
- Statistical Methods in Cancer Epidemiology \newline using R
author: 
- \textbf{Karri Seppä}  \newline \newline
- Finnish Cancer Registry  \newline  \newline
- Lecture 9
date: "Mar 23, 2020"
institute: "karri.seppa@cancer.fi"
output:
  beamer_presentation:
    includes:  
         in_header: beamer_header.tex
tables: true
editor_options: 
  chunk_output_type: console
---

## Points to be covered

1. Relative survival estimators
 \medskip
2. Age-standardised relative survival
 \medskip
3. Period approach
 \medskip
4. Regression modelling of relative survival.


- Packages \texttt{popEpi} and \texttt{rstpm2}.
\medskip
- Functions \texttt{survtab()}, \texttt{glm()}, \texttt{relpois\_ag()}, \texttt{stpm2()}.


## Population-based cancer survival

- Relative (net) survival is the standard measure in population-based cancer survival

- Relative survival summarizes patients' excess mortality, i.e. additional mortality patients have compared with mortality in a relevant reference population.

- Does not rely on cause of death information
    - The **quality of cause-of-death information varies** over time, between types of cancer and between regions and countries.
    - Many cancer registries **do not record cause of death**.
    - Cause of death is rarely a simple dichotomy.



## Excess mortality and relative survival

- Total mortality among patients $\lambda(t)$ is the sum of population mortality $\lambda_\text{P}(t)$ and excess mortality $\lambda_\text{E}(t)$ due to cancer:
\[\lambda(t)=\lambda_\text{P}(t) + \lambda_\text{E}(t)\]


- Relative survival $S_E(t)$ is the survival function related to the excess mortality:
\[S_\text{E}(t)=\exp\left[ - \int_0^t \lambda_\text{E}(u)\ du \right] \]
    - Often interpreted as the survival **probability** in a hypothetical situation where the cancer would be the only possible cause of death with all other
causes being eliminated.
    - **N.B.** Independence of competing risks is required but this assumption cannot be assessed based on the data! 

## Non-parametric relative survival estimators (hazard-based)

- Ederer II (Ederer and Heise 1958)
\[S_\text{E,EII}(t_k)=\exp\left(-\Delta_j\frac{d_j-d_j^*}{y_j} \right)\]
    - $\Delta_j$ = length of follow-up time interval $j$
    - $d_j$ = number of deaths 
    - $d_j^*$ = expected number of deaths (same as in SMR estimation)
    - $y_j$ = number person-years


- Pohar Perme estimator (Pohar Perme et al. 2012)
\[S_\text{E,PP}(t_k)\exp\left(-\Delta_j\frac{d_j^w-d_j^{*w}}{y_j^w} \right)\]
    - $d_j^w$, $d_j^{*w}$ and $y_j^w$ are based on individual-level observations
of $d_j$, $d_j^*$ and $y_j$ weighted by the patient-specific cumulative expected survival probability (see Seppä et al. 2016).



## Properties of the estimators

- Non-standardised Ederer II estimator overestimates relative survival because of informative censoring of old patients with high other-cause mortality
    - use age-standardised Ederer II

- Pohar Perme estimator corrects for this bias by giving larger weights for  observations of older patients

- Both the Pohar Perme and the age-standardised
Ederer II estimator are valid for 5-year net survival of cancer patients.


## Age-standardised relative survival

\begin{itemize}
\item  Weighted average of age-specific relative survival estimates
\begin{enumerate}[]
\item $S_\text{E}(t)=\sum_{a=1}^K w_aS_{\text{E},a}(t)\quad$ where $\sum_{a=1}^K w_a =1$
\end{enumerate}
\begin{itemize}
\item weight $w_a$ is a standard for the proportion of patients in age group $a$ \underline{at the beginning of follow-up} 
\end{itemize}  
\end{itemize}  

- E.g. International Cancer Survival Standards 
    - ICSS1: elderly (most cancers)
    - ICSS2: little age dependence (eg. melanoma, brain, cervix)
    - ICSS3: young adults (eg. testis, hodgkin lymphoma); starify 15-44 into narrower groups
    

Age group ($a$) | $w_\text{ICSS1}$ | $w_\text{ICSS2}$ | $w_\text{ICSS3}$
--------------- | ---------------- | ---------------- | -----------------
15-44           | 0.07             | 0.28             | 0.60
45-54           |	0.12             | 0.17             |	0.10
55-64           | 0.23             | 0.21	            | 0.10
65-74           | 0.29             | 0.20             |	0.10
75+             | 0.29	           | 0.14             |	0.10


## Ex. Oral cancer data

Relative survival analysis

- expected mortality is derived from the  mortality rates in the population of Finland

- assume the date on diagnosis would a random date between 1985 and 2005.


\footnotesize
```{r, comment=NA,echo=T}
orca <- read.table(file="oralca2.txt")
library(Epi)
library(popEpi)
set.seed(23032020)
orca$dg_date <- runif(nrow(orca), min=1985, max=2005)
orca$sex <- ifelse(orca$sex=="Male",0,1)
names(orca)[2] <- "dg_age"
Lex <- Lexis(entry = list(FUT = 0, AGE = dg_age, CAL = dg_date), 
           exit = list(FUT = time), 
           data = orca,
           exit.status = 
             factor(event, levels = 0:2, 
                    labels = c("alive", "canD", "othD")), 
           merge = TRUE)
```

## Ex. Oral cancer data, Ederer II and Pohar Perme

\footnotesize
```{r, comment=NA,echo=T,message=FALSE}
pm <- popmort
names(pm) <- c("sex", "CAL", "AGE", "haz")
cutpoints <- c(0,55,65,75,Inf)
Lex$agegr <- cut(Lex$dg_age, cutpoints, right=F)

#Age-standardised Ederer II estimator
surv.EII <- survtab(FUT ~ 1 + adjust(agegr), data = Lex, 
              surv.type = "surv.rel",
              pophaz = pm, weights = "internal",
              breaks = list(FUT = seq(0, 5, 1/12)))
#Pohar Perme estimator
surv.PP <- survtab(FUT ~ 1, data = Lex, 
              surv.type = "surv.rel", relsurv.method = "pp",
              pophaz = pm,
              breaks = list(FUT = seq(0, 5, 1/12)))
```

## Ex. Oral cancer data, Ederer II and Pohar Perme

\footnotesize
```{r, comment=NA,echo=T,message=F,fig.height=5.5, dev.args=list(pointsize=23)}
par(mar=c(4,4,0.5,1))
plot(surv.EII, lwd=2, ylim=c(0,1))
lines(surv.PP, lwd=2, col="red")
legend("bottomleft",c("Age-standardised Ederer II (internal weights)",
                    "Pohar Perme"),
       lty=1, lwd=2 ,col=c("black","red"))
```


## Ex. Oral cancer data, ICSS weights

Use external weights (ICSS1) in age standardisation:

\footnotesize
```{r, comment=NA,echo=T,message=F,fig.height=5.5, dev.args=list(pointsize=23)}
ICSS$agegr <- cut(ICSS$age, cutpoints, right = FALSE)
w <- aggregate(ICSS1~agegr, data = ICSS, FUN = sum)
w
```

## Ex. Oral cancer data, ICSS weights

\footnotesize
```{r, comment=NA,echo=T,message=F,fig.height=5.5, dev.args=list(pointsize=23)}
Lex$agegr <- cut(Lex$dg_age, cutpoints, right=F)
surv.EII.ICSS <- survtab(FUT ~ 1+adjust(agegr), data = Lex, 
              surv.type = "surv.rel",
              pophaz = pm, weights = w$ICSS1,
              breaks = list(FUT = seq(0, 5, 1/12)))
surv.PP.ICSS <- survtab(FUT ~ 1+adjust(agegr), data = Lex, 
              surv.type = "surv.rel",relsurv.method = "pp",
              pophaz = pm, weights = w$ICSS1,
              breaks = list(FUT = seq(0, 5, 1/12)))
```



## Ex. Oral cancer data, internal vs ICSS weights

\footnotesize
```{r, comment=NA,echo=T,fig.show='hide',message=F}
par(mar=c(4,4,0.5,1))
plot(surv.EII, lwd=2, conf.int=F,ylim=c(0,1)) 
lines(surv.PP, lwd=2,col="red", conf.int=F)
lines(surv.EII.ICSS, lwd=2, lty=2, conf.int=F)
lines(surv.PP.ICSS, lwd=2,col="red",lty=2, conf.int=F)

legend("bottomleft",c("Age-standardised Ederer II (internal weights)",
                    "Pohar Perme",
                    "Age-standardised Ederer (ICSS weights)",
                    "Age-standardised Pohar Perme (ICSS weights)"),
       lty=c(1,1,2,2), lwd=2,col=c("black","red","black","red"))
```

## Ex. Oral cancer data, ICSS weights

\footnotesize
```{r, comment=NA,echo=F,message=F,fig.height=5.5, dev.args=list(pointsize=20)}
par(mar=c(4,4,0.5,1))
plot(surv.EII, lwd=2, conf.int=F,ylim=c(0,1)) 
lines(surv.PP, lwd=2,col="red", conf.int=F)
lines(surv.EII.ICSS, lwd=2, lty=2, conf.int=F)
lines(surv.PP.ICSS, lwd=2,col="red",lty=2, conf.int=F)

legend("bottomleft",c("Age-standardised Ederer II (internal weights)",
                    "Pohar Perme",
                    "Age-standardised Ederer (ICSS weights)",
                    "Age-standardised Pohar Perme (ICSS weights)"),
       lty=c(1,1,2,2), lwd=2,col=c("black","red","black","red"))
```

## Period approach

- Survival estimates in a cohort of patients tend to be outdated
    - 5-year survival estimate requires follow-up data on patients diagnosed $\geq 5$ years ago.
    
- Use period approach (Brenner et al. 2004) to produce maximally
updated estimates of patient survival
    - follow-up of patients is restricted to a calendar time window.
    - e.g. patients diagnosed in 2011-2018 and followed-up in 2016-2018.
    - patients diagnosed earlier than 2016 do not contribute to survival estimate before coming to the follow-up window 2016-2018 (late entry).
    - interpretation: survival in patients, who would have the same follow-up time specific excess mortality rates as those observed in 2016-2018. (A prediction for patients diagnosed in 2016-2018.) 


## Period approach

Example: 5-year survival based on patients diagnosed in 2011-2018 and followed in calendar time window 2016-2018. 

\includegraphics[height=7cm]{period_approach}

## Ex. Oral cancer data, period approach

\footnotesize
```{r, comment=NA,echo=T,message=F,fig.height=5.5, dev.args=list(pointsize=20)}
#Age-standardised Ederer II estimator
surv.EII.period <- survtab(FUT ~ 1 + adjust(agegr), data = Lex, 
              surv.type = "surv.rel",
              pophaz = pm, weights = "internal",
              breaks = list(FUT = seq(0, 5, 1/12), CAL=c(2002,2005)))
```

## Ex. Oral cancer data, period approach

\footnotesize
```{r, comment=NA,echo=T,message=F,fig.height=5.5, dev.args=list(pointsize=20)}
#Age-standardised Ederer II estimator
par(mar=c(4,4,0.5,1))
plot(surv.EII, lwd=2, conf.int=F,ylim=c(0,1)) 
lines(surv.EII.period, lwd=2,col="red", conf.int=F)
legend("bottomleft",c("Diagnosed in 1985-2004",
         "Diagnosed in 1985-2004 and followed-up in 2002-2004"),
       lty=c(1,1), lwd=2,col=c("black","red"))
```



## Regression models for relative survival

1. Logarithm of excess mortality
\[\log[\lambda_\text{E}(t)]=\alpha+\beta X\]
    
- can be modelled by using Poisson regression with a user-specified link function (Dickman et al. 2004).

\medspace

2. Logarithm of cumulative excess mortality

\[\log[\Lambda_\text{E}(t)]=\text{spline}[log(t)\mid \alpha]+\beta X\]
    
- can be modelled using spline functions for the baseline cumulative excess hazard (\texttt{rstpm2} package)


## Ex. Oral cancer data, piecewise constant excess hazard

\footnotesize
```{r, comment=NA,echo=T,message=F,fig.height=8, dev.args=list(pointsize=20)}
#cutpoints for piecewise constant excess hazard
fb <- c(0,1,2,3,5)
pm2 <- pm
names(pm2) <- c("sex","year","agegroup","haz")
orca$agegr <- cut(orca$dg_age, cutpoints, right=F)
#combines orca and population mortality rates
Lex <- lexpand(orca, birth = dg_date-dg_age, entry = dg_date,
             exit = dg_date+time, status=event %in% 1:2,
             breaks = list(fot=fb), 
             pophaz=pm2, pp = FALSE,
             aggre = list(agegr, fot))
```

## Ex. Oral cancer data, piecewise constant excess hazard

Aggregated data

\footnotesize
```{r, comment=NA,echo=T,message=F,fig.height=8, dev.args=list(pointsize=20)}
Lex
```


## Ex. Oral cancer data, piecewise constant excess hazard

Proportional hazards between age groups?

\footnotesize
```{r, comment=NA,echo=T,message=F,fig.height=8, dev.args=list(pointsize=20)}
## fit model using aggregated data
rpm <- relpois_ag(formula = from0to1 ~ fot + agegr,  
                  data = Lex, d.exp = d.exp, offset = log(pyrs))
ci.exp(rpm)
 
## non-proportional effect of age (=follow-up time age interaction)
rpm2 <- update(rpm, . ~ fot*agegr)
anova(rpm, rpm2, test="LRT")$`Pr(>Chi)`[2]
```

## Ex. Oral cancer data, spline functions

Mortality from any cause and from cancer, and excess mortality

\footnotesize
```{r, comment=NA,echo=T,message=F,fig.height=5.5, dev.args=list(pointsize=20),warning=FALSE}
library(rstpm2)
orca$agegr <- cut(orca$dg_age, cutpoints, right=F)

#mortality from any cause
fit <- stpm2(Surv(time,event%in%c(1,2))~1,
             data=orca,df=3)

#cause (cancer)-specific mortality
fit2 <- stpm2(Surv(time,event%in%c(1))~1,
              data=orca,df=3)

#excess mortality
orca$ex_age <- floor(orca$dg_age+orca$time)
orca$ex_year <- floor(orca$dg_date+orca$time)
orca <- merge(orca,pm,
              by.x=c("sex","ex_age","ex_year"),
              by.y=c("sex","AGE","CAL"))
fit3 <- stpm2(Surv(time,event%in%c(1,2))~1,
              data=orca,df=3,
              bhazard=orca$haz)
```

## Ex. Oral cancer data, hazard

\footnotesize
```{r, comment=NA,echo=T,message=F,fig.height=5.5, dev.args=list(pointsize=20),fig.show='hide'}
plot(fit,newdata=data.frame(time=seq(0.1,5,0.1)),
     type="hazard",add=FALSE,ci=FALSE,line.col=1,
     ylim=c(0.01,0.3),rug=F)
plot(fit2,newdata=data.frame(time=seq(0.1,5,0.1)),
     type="hazard",add=TRUE,ci=FALSE,line.col=2)
plot(fit3,newdata=data.frame(time=seq(0.1,5,0.1)),
     type="hazard",add=TRUE,ci=FALSE,line.col=3)
legend("topright",
       legend=c("Total mortality",
                "Oral cancer mortality",
                "Excess mortality"),
       col=c("black","red","green"),lty=1)
```

## Ex. Oral cancer data, hazard

\footnotesize
```{r, comment=NA,echo=F,message=F,fig.height=8, dev.args=list(pointsize=20)}
plot(fit,newdata=data.frame(time=seq(0.1,5,0.1)),
     type="hazard",add=FALSE,ci=FALSE,line.col=1,
     ylim=c(0.01,0.3),rug=F)
plot(fit2,newdata=data.frame(time=seq(0.1,5,0.1)),
     type="hazard",add=TRUE,ci=FALSE,line.col=2)
plot(fit3,newdata=data.frame(time=seq(0.1,5,0.1)),
     type="hazard",add=TRUE,ci=FALSE,line.col=3)
legend("topright",
       legend=c("Total mortality",
                "Oral cancer mortality",
                "Excess mortality"),
       col=c("black","red","green"),lty=1)
```

<!---

## Ex. Oral cancer data, spline by age 

Assuming proportional hazards

\footnotesize
```{r, comment=NA,echo=T,message=F,fig.height=8, dev.args=list(pointsize=20),fig.show='hide',eval=F}
#proportional hazards between age groups
fit4 <- stpm2(Surv(time,event%in%c(1,2))~agegr,
              data=orca,df=3,
              bhazard=orca$haz)

# allowing for non-proportional hazards
fit5 <- stpm2(Surv(time,event%in%c(1,2))~agegr,
              data=orca,df=3,tvc=list(agegr=3),
              bhazard=orca$haz)

plot(fit4,newdata=data.frame(time=seq(0.1,5,0.1),agegr="[0,55)"),
     type="hazard",add=FALSE,ci=FALSE,line.col=1,lwd=2,
     ylim=c(0.01,0.45),rug=F)
plot(fit4,newdata=data.frame(time=seq(0.1,5,0.1),agegr="[0,55)"),
     type="hazard",add=TRUE,ci=FALSE,line.col=2,lwd=2)
plot(fit4,newdata=data.frame(time=seq(0.1,5,0.1),agegr="[55,65)"),
     type="hazard",add=TRUE,ci=FALSE,line.col=3,lwd=2)
plot(fit4,newdata=data.frame(time=seq(0.1,5,0.1),agegr="[65,75)"),
     type="hazard",add=TRUE,ci=FALSE,line.col=4,lwd=2)
plot(fit4,newdata=data.frame(time=seq(0.1,5,0.1),agegr="[75,Inf)"),
     type="hazard",add=TRUE,ci=FALSE,line.col=5,lwd=2)
legend("topright",
       legend=c("0-44","45-54","55-64","65-74","75+"),
       col=1:5,lty=1,lwd=2)
```


## Ex. Oral cancer data, spline by age 

Assuming proportional hazards

\footnotesize
```{r, comment=NA,echo=F,message=F,fig.height=8, dev.args=list(pointsize=20),eval=F}
par(mar=c(4,4,1,1))
plot(fit4,newdata=data.frame(time=seq(0.1,5,0.1),agegr="[0,55)"),
     type="hazard",add=FALSE,ci=FALSE,line.col=1,lwd=2,
     ylim=c(0.01,0.45),rug=F)
plot(fit4,newdata=data.frame(time=seq(0.1,5,0.1),agegr="[0,55)"),
     type="hazard",add=TRUE,ci=FALSE,line.col=2,lwd=2)
plot(fit4,newdata=data.frame(time=seq(0.1,5,0.1),agegr="[55,65)"),
     type="hazard",add=TRUE,ci=FALSE,line.col=3,lwd=2)
plot(fit4,newdata=data.frame(time=seq(0.1,5,0.1),agegr="[65,75)"),
     type="hazard",add=TRUE,ci=FALSE,line.col=4,lwd=2)
plot(fit4,newdata=data.frame(time=seq(0.1,5,0.1),agegr="[75,Inf)"),
     type="hazard",add=TRUE,ci=FALSE,line.col=5,lwd=2)
legend("topright",
       legend=c("0-44","45-54","55-64","65-74","75+"),
       col=1:5,lty=1,lwd=2)
```

## Ex. Oral cancer data, spline by age

Allowing for non-proportional hazards

\footnotesize
```{r, comment=NA,echo=T,message=F,fig.height=8, dev.args=list(pointsize=20),fig.show='hide',eval=F}
plot(fit5,newdata=data.frame(time=seq(0.1,5,0.1),agegr="[0,55)"),
     type="hazard",add=FALSE,ci=FALSE,line.col=1,lwd=2,
     ylim=c(0.01,0.45),rug=F)
plot(fit5,newdata=data.frame(time=seq(0.1,5,0.1),agegr="[0,55)"),
     type="hazard",add=TRUE,ci=FALSE,line.col=2,lwd=2)
plot(fit5,newdata=data.frame(time=seq(0.1,5,0.1),agegr="[55,65)"),
     type="hazard",add=TRUE,ci=FALSE,line.col=3,lwd=2)
plot(fit5,newdata=data.frame(time=seq(0.1,5,0.1),agegr="[65,75)"),
     type="hazard",add=TRUE,ci=FALSE,line.col=4,lwd=2)
plot(fit5,newdata=data.frame(time=seq(0.1,5,0.1),agegr="[75,Inf)"),
     type="hazard",add=TRUE,ci=FALSE,line.col=5,lwd=2)
legend("topright",
       legend=c("0-44","45-54","55-64","65-74","75+"),
       col=1:5,lty=1,lwd=2)
```

## Ex. Oral cancer data, spline by age

Allowing for non-proportional hazards

\footnotesize
```{r, comment=NA,echo=F,message=F,fig.height=8, dev.args=list(pointsize=20),eval=F}
par(mar=c(4,4,1,1))
plot(fit5,newdata=data.frame(time=seq(0.1,5,0.1),agegr="[0,55)"),
     type="hazard",add=FALSE,ci=FALSE,line.col=1,lwd=2,
     ylim=c(0.01,0.45),rug=F)
plot(fit5,newdata=data.frame(time=seq(0.1,5,0.1),agegr="[0,55)"),
     type="hazard",add=TRUE,ci=FALSE,line.col=2,lwd=2)
plot(fit5,newdata=data.frame(time=seq(0.1,5,0.1),agegr="[55,65)"),
     type="hazard",add=TRUE,ci=FALSE,line.col=3,lwd=2)
plot(fit5,newdata=data.frame(time=seq(0.1,5,0.1),agegr="[65,75)"),
     type="hazard",add=TRUE,ci=FALSE,line.col=4,lwd=2)
plot(fit5,newdata=data.frame(time=seq(0.1,5,0.1),agegr="[75,Inf)"),
     type="hazard",add=TRUE,ci=FALSE,line.col=5,lwd=2)
legend("topright",
       legend=c("0-44","45-54","55-64","65-74","75+"),
       col=1:5,lty=1,lwd=2)
```

--->