---
title: 'Epidemiologic data analysis using R '
author:
- Faculty of Social Sciences, University of Tampere
- ' -- '
- Janne Pitkaniemi / Karri Seppä
date: 23.3.2020
output:
  word_document: default
  pdf_document: default
subtitle: Practicals 9
editor_options: 
  chunk_output_type: console
---


```{r p7_0, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, comment = NA)
options(warn=-1)
setwd("C:/Users/karri.seppa/Documents/TRE")
```


# Topics of practical 9



Learning objectives of this practical

*   Compare estimates and interpretations of different survival measures (overall, cause-specific and relative survival, and 1-cumulative incidence)
*   Age standardisation of relative survival
*   Piecewise constant excess hazard model for relative survival


# Survival analysis: Rectal cancer patients


## Description of the data


Data frame ``sire`` includes observations of a simulated cohort of 8243 female Finnish rectal cancer patients diagnosed between 1993-2012.
The dataset contains the following variables:

variable    | description
----------- | --------------------
sex         | 0 = Male, 1 = Female   
bi_date     |date of birth  
dg_date     |date of cancer diagnosis 
ex_date     |date of exit from follow-up (death or censoring)
status      |status at the end of follow-up (numeric): \newline 0 = alive, 1 = death from rectal cancer, 2 = death from other causes. 
dg_age      | age at diagnosis (in fractional years)

            

## Loading the packages and the data


Load the R packages ``Epi`` and ``popEpi`` needed in this exercise.
```{r p9_1, eval=TRUE,echo=TRUE,warnings=FALSE, message=FALSE, results='hide'}
library(Epi)
library(popEpi)
#mortality rates of the population of Finland (included in popEpi package)
pm <- popmort
```




First, let's estimate the different measures of survival by using ``survtab`` function in ``popEpi`` package.

## Lexis object with multi-state set-up


```{r p9_2, eval=TRUE,echo=TRUE,warnings=FALSE}
sire$agegr <- cut(sire$dg_age,c(0,45,55,65,75,Inf),right=F)

Lex <- Lexis(entry = list(FUT = 0, AGE = dg_age, 
                          CAL = cal.yr(dg_date)),
             exit = list(CAL = cal.yr(ex_date)),
             data = sire,
             exit.status = factor(status, levels = 0:2,
                                  labels = c("alive", "canD", "othD")),
             merge = TRUE)
```


## Overall survival

In overall survival, outcome event is death from any cause.

```{r p9_3, echo=TRUE, message=F}
#piecewise constant hazard using monthly intervals
BL <- list(FUT=seq(0, 15, by = 1/12))
#overall survival
st.obs <- survtab(
  Surv(time=FUT, event=lex.Xst%in%c("canD", "othD")) ~ 1, 
  data = Lex,surv.type = "surv.obs",breaks = BL)
#summary(st.obs,t=c(1,5,10,15))
```





## 1 - Cumulative incidence of rectal cancer death

One minus cumulative incidence = Proportion of patients who avoid death from rectal cancer in the presence of other causes of death. Rectal cancer death is avoided if a patient died from other causes than cancer.

```{r p9_4, echo=TRUE, message=F}
st.cif <- survtab(Surv(time = FUT, event = lex.Xst) ~ 1, 
                  data = Lex, 
                  surv.type = "cif.obs",breaks = BL)
#summary(st.cif,t=c(1,5,10,15))
```





## Relative survival

Relative survival gives an estimate of net survival = the proportion of patients alive in the absence of other causes of death than cancer

```{r p9_5, echo=TRUE, message=F}
names(pm) <- c("sex","CAL","AGE","haz")
w <- table(Lex$agegr)
w <- list(agegr = as.numeric(w))
st.rel <- survtab(
  Surv(time = FUT, event = lex.Xst%in%c("canD", "othD")) 
    ~ adjust(agegr), data = Lex, pophaz = pm,
    surv.type = "surv.rel",breaks = BL, weights=w)
#summary(st.rel,t=c(1,5,10,15))
```




## Cause-specific survival (event = death from rectal cancer)

In cause-specific survival (normal Kaplan-Meier), outcome event is death from rectal cancer. It gives an estimate of net survival = the proportion of patients alive in the absence of other causes of death than cancer

```{r p9_6, echo=TRUE, message=F}
st.cause<- survtab(
  Surv(time = FUT, event = lex.Xst) ~ adjust(agegr), 
  data = Lex, pophaz = pm,
  surv.type = "surv.cause",breaks = BL, weights=w)
#summary(st.cause,t=c(1,5,10,15))
```


```{r p9_7, echo=FALSE, message=FALSE, results='hide',fig.width=8,fig.height=8}
plot(st.obs,conf.int=F,ylim=c(0,1),ylab="")
lines(st.cif$Tstop,1-st.cif$CIF_canD,col="magenta")
lines(st.rel,y="r.e2.as",conf.int=F,col="dodgerblue")
lines(st.cause,y="surv.obs.canD.as",conf.int=F,col="green")
legend("bottomleft",c("Overall survival",
                      "Proportion of patients who avoid rectal cancer death",
                      "Relative survival (proportion who avoid rectal cancer death\n in the absence of other causes of death)",
                      "Cause-specific survival (proportion who avoid rectal cancer death\n in the absence of other causes of death)"),
       col=c("black","magenta","blue","green"),lty=1)
```

Compare the estimates and their interpretation.
Remember that the probability interpretation of relative survival and cause-specific survival relies on the assumption that competing risks are independent, and you never know whether it is true or not. 

Relative and cause-specific survival are, however, useful summary measures of excess mortality and cause-specific mortality, respectively, when the aim is to compare cancer survival between calendar periods (or e.g. countries) between which mortality due to other causes of death may also differ. 


# Age-standardised relative survival

Has rectal cancer survival in improved from 1993-2002 to 2003-2012. Estimate the age-standardised relative survival using ICSS weights and the Pohar Perme estimator.  

```{r p9_8, echo=TRUE, message=F}
Lex$period <- cut( cal.yr(Lex$dg_date), c(1993,2003,2013), right=F,dig.lab = 4)
ICSS1 <- aggregate(ICSS1~cut(ICSS$age, 
                   c(0,45,55,65,75,Inf), right = FALSE), 
                   data = ICSS, FUN = sum)
w <- list(agegr = ICSS1$ICSS1)
st.rel <- survtab(
  Surv(time = FUT, event = lex.Xst%in%c("canD", "othD")) 
  ~ sex + period + adjust(agegr), 
  data = Lex,pophaz = pm,
  surv.type = "surv.rel",breaks = BL, weights=w)
summary(st.rel,t=c(1,5,10,15))
```


```{r p9_9, echo=FALSE, message=FALSE, results='hide'}
plot(st.rel[st.rel$period=="[1993,2003)",])
lines(st.rel[st.rel$period=="[2003,2013)",],col="red")
legend("topright",c("1993-2002","2003-2012"),col=c("black","red"),lty=1)
```



# Regression model for relative survival

Estimate the relative excess risk (ratio of excess hazards) for the difference in rectal cancer survival between 1993-2002 and 2003-2012. First, tabulate the observations by follow-up time interval, age group, sex and period.

```{r p9_10, echo=TRUE, message=F}
BL <- list(fot=c(0,3/12,1:5))
names(pm) <- c("sex","year","agegroup","haz")
sire$period <- cut( cal.yr(sire$dg_date), c(1993,2003,2013), right=F,dig.lab = 4)
Lex <- lexpand(sire, birth = bi_date,
               entry = dg_date, exit = ex_date,
               status = status%in%1:2,
               breaks = BL, pophaz = pm,pp=F,
               aggre = list(fot,agegr,sex,period)
)

```

The piecewise constant excess hazard model can be fitted using ``relpois_ag`` function in ``popEpi`` package:

```{r p9_11, echo=TRUE, message=F}
m1 <- relpois_ag(formula = from0to1 ~ period+fot*agegr, 
                 data = Lex, d.exp = d.exp, offset = log(pyrs))
ci.exp(m1)
```

Age-adjusted excess mortality among patients diagnosed in 2003-2012 is 0.78-fold compared to patients diagnosed in 1993-2002. 

In cancer survival, it is almost always necessary to include interactions between follow-up time and age into the model (i.e. excess hazards are not proportional between age groups):

```{r p9_12, echo=TRUE, message=F}
m0 <- relpois_ag(formula = from0to1 ~ period+fot + agegr, 
                 data = Lex, d.exp = d.exp, offset = log(pyrs))
anova(m0,m1,test="LRT")
```

The same model can be fitted using ``glm`` function. The link function of relative survival Poisson regression can be specified as follows:

```{r p9_13, echo=TRUE, message=F}
Lex$FUT <- cut( Lex$fot, c(0,3/12,1:5), right=F)
d_star <- Lex$d.exp

#define link function for relative survival regression
rspois <- poisson()
rspois$link <- "glm relative survival model with Poisson error"
rspois$linkfun <- function(mu) log(mu - d_star)
rspois$linkinv <- function(eta) d_star + exp(eta)
assign(".d_star", Lex$d.exp, env = .GlobalEnv)
rspois$initialize <- expression({
  if (any(y < 0)) stop(paste("Negative values not allowed for", 
                             "the Poisson family"))
  n <- rep.int(1, nobs)
  mustart <- pmax(y, .d_star) + 0.1
})

#glm model
m2 <- glm(formula = from0to1 ~ period+FUT*agegr, 
          data = Lex, family=rspois, offset = log(pyrs))
ci.exp(m2)
```



