---
title: |
  | Epidemiologic Data Analysis using R 
  | Part 9 
  | Time dependent covariates
author: "Janne Pitkäniemi"
date: |
  | Tampere University, Faculty of Social Sciences 
  | University of Tampere Postgraduate training 
  | Feb 26 - Apr 9, 2018
output:
  beamer_presentation:
    includes:
      in_header: mystyle.tex
---



```{r setup, include=FALSE,comment=NA}
knitr::opts_chunk$set(echo = FALSE,warn=-1)
library(Epi)
```

## Time-Dependent covariates in TTE

- A key underlying condition (a martingale) is that present actions depend only on the past. 
- The decision of whether one is in the risk set or not 
and covariates can depend in any way on prior covariates and
risk set patterns
- *but cannot look into the future.*

## Analysis by treatment response - NO ! (T. Thernau)

- An example based on the advanced lung cancer data set, 
where assignment of the responder was made by looking 
- The problem arises because any early deaths, those that occur before
response can be assessed, will all be assigned to the non-responder group, even deaths that have nothing to do with the condition under study.
- Assume that subjects came in monthly for 12 cycles of treatment, and
randomly declare a **response** for 5% of the subjects at each visit.

```{r,warnings=FALSE,comment=NA,message=FALSE,fig.height=6, fig.width=15}
suppressWarnings(library(survival))
suppressWarnings(library(Epi))
suppressWarnings(library(lubridate))
suppressWarnings(library(data.table))
suppressWarnings(library(rstpm2))

set.seed(1953) # a good year
nvisit <- floor(pmin(lung$time/30.5, 12))
 response <- rbinom(nrow(lung), nvisit, .05) > 0
 badfit <- survfit(Surv(time/365.25, status) ~ response, data=lung)

 plot(badfit, mark.time=FALSE, lty=1:2,
xlab="Years post diagnosis", ylab="Survival")
 legend(1.5, .85, c("Responders", "Non-responders"),
lty=2:1, bty='n')
```
 

## Classification of time dependent covariates

- Values change continuosly (change in small during a small time interval) blod pressure, cholesterol **Diffusion process**
- Values change a lot during a small time interval (start and stop smoking 0->1, number of visits before a given time point 0,1,2,...) **Counting process**

- The longer the follow-up time the more likely are changes

- lognitudinal data: registry based follow-up, repeated measures, e-equipments (smart tech.)

## Heart transplant data

Mortality of patients on the waiting list for the Stanford heart transplant program.

- start, stop:	Entry and exit time 
- Event:  and status for this interval of time (0=Alive, 1=Death)
- age:	 age-48 years
- year:	 year of acceptance (in years after 1 Nov 1967)
- surgery:	 prior bypass surgery 1=yes
- transplant:	 received transplant 1=yes
- id:	 patient id


```{r, ,comment=NA,size="tiny"}
print(format(head(heart,n=3),digits=2),row.names=FALSE)
```

## Heart transplant data

Cox model with constant treatment (transplant) effect

```{r,eval=TRUE, echo=TRUE,comment=NA}
ci.lin(coxph(Surv(start, stop, event) ~
    age + year + surgery + transplant,
data=heart),Exp=TRUE)[,c(5:7)]
```

Transplant has no effect on mortality RR=0.99 !!!!!

## How the risk is modified by time


- Change in the covariate value changes the risk instantly
- time lag of t > 0 months,
- risk increases gradually until reaches some higher level or
- risk increases at first but decreases later.
In this case the risk depends on the time when the covariate value
changed.

- With diffuse processes change times can be defined in many ways!
covariate value exceeds a certain level or
covariate value increases by a certain amount during a (short) time interval.

## HT data - the effect of transplantation

- Divide the follow-up time
into e.g. at most four parts
based on
- transplantation + 0 days
- transplantation + 30 days
- transplantation + 60 days.

Model a RR parameter for
each interval.

Natural reference category
can be interval before
transplantation (set RR=1).

## Time dependent effect of transplant -Lexis object

```{r,eval=TRUE,echo=TRUE,comment=NA}
heart.Lx <-
Lexis(entry=list(age.time=age, 
                 fu.time=0,
tx.time=decimal_date(tx.date)-decimal_date(accept.dt)),
duration=futime/365, 
exit.status=as.character(fustat),
data=subset(jasa, is.na(tx.date) | fu.date != tx.date))
```

## Time dependent effect of transplant -Lexis object

```{r,eval=TRUE,echo=TRUE,comment=NA}

heart.Lx <- within(heart.Lx, {cut.0 <- tx.time; 
                              cut.1 <- tx.time+0.2;
                              cut.2 <- tx.time+0.4});     
```
```{r,eval=TRUE,echo=FALSE,comment=NA}

subset(heart.Lx[,1:6], heart.Lx$lex.id==66)
subset(heart.Lx[,7:13], heart.Lx$lex.id==66)
subset(heart.Lx[,14:20], heart.Lx$lex.id==66)
subset(heart.Lx[,20:24], heart.Lx$lex.id==66)
```

## Time dependent effect of transplant

Package Epi contains function mcutLexis, which splits the follow-up
interval w.r.t. variables cut.0, cut.1 and cut.2:
accept.dt acceptance == 0
tx.time transplant date (cut.0)
tx.time +0.2  (cut.1)
tx.time+0.4 (cut.2)


```{r, comment=NA}
heart.Lx.cut <- mcutLexis(heart.Lx, timescale="fu.time",
wh=c("cut.0","cut.1","cut.2"),
new.states=c("a","b","c"))
```
```{r,eval=TRUE,echo=FALSE,comment=NA}

subset(heart.Lx.cut[,1:6], heart.Lx.cut$lex.id==66)
subset(heart.Lx.cut[,20:24], heart.Lx.cut$lex.id==66)
```

## Time dependent effect of transplant

Reference group is the time between acceptance and transplant operation, with RR=1

- a [transplant+0.2]
- a-b [transplant+0.2,transplant+0.4]
- a-b-c [transplant+0.4,]


```{r,eval=TRUE,echo=TRUE,comment=NA}
print(format(as.data.frame(ci.exp(coxph( 
      Surv(fu.time, fu.time + lex.dur, lex.Xst==1)
      ~ factor(lex.Cst),data=heart.Lx.cut))),digits=3)) 
```


## What should I consider when interpreting results

Main assumption has been that changes in the covariate value
does not depend on outcome process. 
E.g. an intermediator:

- Treatment for HIV begins when CD4 has decreased below 350.
- Risk to contract AIDS (CD4<200) is higher for patients who have
started treatment than patients who have not (CD4>350).
- A simple analysis where timedependent variable is defined by the
time when CD4 reaches 350 may result in conclusion that the HIV
treatment increases the risk of AIDS!
- In reality the risk would be higher without the treatment.
- More elaborate methods and adjustments are needed () consult a
statistician).

## How to proceed?

- Timedependent covariates can provide valuable information on
treatment or risk factor effects.
- But: application of timedependent covariates in predicting e.g.
survival can be challenging.
- Association of a timedependent covariate and the outcome can be
instant, lagged or cumulative ) selection of appropriate functional
form?
- What causes the change of the covariate value?
- If the changetime of the covariate is completely random, the
methods presented here are appropriate (e.g. RCT, heart
transplantation example).


