---
title:
- Statistical Methods in Cancer Epidemiology \newline using R
author: 
- \textbf{Karri Seppä}  \newline \newline
- Finnish Cancer Registry  \newline  \newline
- Lecture 8
date: "Mar 23, 2020"
institute: "karri.seppa@cancer.fi"
output:
  beamer_presentation:
    includes:  
         in_header: beamer_header.tex
tables: true
editor_options: 
  chunk_output_type: console
---

## Timetable today

- Lecture 8 (competing risks): 
    - 9.15-10.00

- Practical 7 (survival without competing risks): 
    - 10.15 ->

- Lecture 9 (population-based cancer survival): 
    - 13.15-14.00

- Practical 8 (competing risks): 
    - 14.15 ->
 


## Points to be covered

1. Competing risks: event-specific cumulative incidences & hazards.
 \medskip
2. Kaplan--Meier and Aalen--Johansen estimators.
 \medskip
3. Regression modelling of competing hazards:
 \medskip
4. Packages \texttt{survival, mstate, cmprisk}.
\medskip
5. Functions \texttt{Surv(), survfit(), plot.survfit(), coxph(), Cuminc()}.




## Ex. Survival of 338 oral cancer patients


Important variables: 

\begin{itemize}
\item \texttt{time} = duration of patientship from \\ 
 diagnosis (\textbf{entry}) till death or censoring, 
\medskip
\item
\texttt{event} = indicator for the outcome and its \\
 observation at the end of follow-up (\textbf{exit}): \\
  0 = censoring,  \\
  1 = death from oral cancer, \\
  2 = death from some other cause.
\end{itemize}


\medskip
Special features:
\begin{itemize}
\item
   Several possible endpoints, \textit{i.e.} alternative causes of death, of which
   only one is realized.
   \medskip
\item
   Censoring -- incomplete observation of the survival time.   
\end{itemize}



## Ex. Oral cancer data

```{r, comment=NA,echo=T}
orca <- read.table(file="oralca2.txt")
head(orca)
```



## Ex. Oral cancer data

Analysis of overall mortality

\scriptsize
```{r, comment=NA,echo=T}
library(survival)
orca$suob <- Surv(orca$time, 1*(orca$event > 0) )
km1 <- survfit( suob ~ 1, data = orca)
km1              #  brief  summary
summary(km1)     #  detailed KM-estimate


```


## Oral cancer: Kaplan-Meier estimates

```{r, comment=NA,echo=T, warning=F, message=FALSE, fig.height=4, dev.args=list(pointsize=23)}
par(mfrow=c(1,3),mar=c(3,2.5,2,0.5),las=1)
plot(km1,main="Survival")
plot(km1,fun="F",main="CDF=1-Survival")
plot(km1,fun="cumhaz",main="Cumulative hazard")
```



## Competing risks model: causes of death
   
\begin{itemize}
\item
Often the interest is focused on the risk or hazard of dying 
from one specific cause.
\medskip
\item
That cause may eventually not be realized, because
a \textbf{competing cause} of death hits first.

\bigskip

\begin{center}
\setlength{\unitlength}{0.65pt}
\begin{picture}(400,190)
  \thicklines
  \put(  0, 80){\framebox(120,50){Alive}}
  \put(230,130){\framebox(180,50){Dead from cancer}}
  \put(230, 30){\framebox(180,50){Dead, other causes}}
  \put(135,125){\vector(3, 1){90}}
  \put(135, 85){\vector(3,-1){90}}
  \put(165,157){\makebox(0,0)[b]{$\lambda_1(t)$}}
  \put(165, 53){\makebox(0,0)[t]{$\lambda_2(t)$}}
\end{picture}
\end{center}
\item
 Generalizes to several competing causes.
 \end{itemize}



## Competing events & competing risks

In many epidemiological and clinical contexts there are
competing events that may 
occur before the target event and remove the person from 
 the population at risk for the event, \textit{e.g.}

\begin{itemize}
\item \textit{target event}: occurrence of endometrial cancer,
 \textit{competing events}: hysterectomy or death.
\medskip
\item \textit{target event}: relapse of a disease \\
(ending the state of remission), \\
 \textit{competing event}: death while still in remission.
 
\item \textit{target event}: divorce, \\ 
 \textit{competing event}: death of either spouse.
 
\end{itemize}



## Event-specific quantities

\textbf{Cumulative incidence function} (CIF) or \newline
 \textbf{subdistribution function} for event $c$:
\[  F_c(t) = P(T \leq t \text{ and } C = c ), \quad c = 1,2,  \]
subdensity function $f_c(t) = dF_c(t)/dt$ 

\medskip
From these one can recover
\begin{itemize}
\item
$F(t) = \sum_{c} F_c(t)$, CDF of event-free survival time $T$, \textit{i.e.} 
cumulative risk of any event by $t$.
\medskip
\item
$S(t) = 1 - F(t)$, \textbf{event-free survival function}, \textit{i.e.} probability of avoiding all events by $t$
\end{itemize}



## Event-specific quantities (cont'd)


\textbf{Event-} or \textbf{cause-specific hazard function}
\begin{align*}
 \lambda_c(t) & =  \underset{\Delta\to 0}{\lim} 
    \frac{P(t < T \le t+\Delta \text{ and } C = c \mid T > t)}{\Delta}  \\
        & =  \frac{f_c(t)}{1-F(t)}
\end{align*}
\begin{enumerate}[$\approx$]
\item Risk of \textit{event} $c$ 
in a short interval $(t, t+ \Delta]$,  given 
\textit{avoidance of all events} up to $t$, 
per interval length.
\end{enumerate}

\textbf{Event-} or \textbf{cause-specific cumulative hazard} 
 $$ \Lambda_c(t) = \int_0^t \lambda_c(v)dv $$ 



## Event-specific quantities (cont'd)

\begin{itemize}
\item CIF = risk of event $c$ over risk period $[0,t]$ in the presence of competing risks, also obtained 
$$ F_c(t) = \int_0^t \lambda_c(v) S(v) dv, \quad c = 1,2, $$
\item
Depends on the hazard of the competing event, too, 
via 
\begin{align*} 
S(t) & = \exp\left\{ - \int_0^t [\lambda_1(v) + \lambda_2(v)]dv 
 \right\} \\
 & = \exp\left\{ - \Lambda_1(t) \right\} \times        \exp\left\{ - \Lambda_2(t) \right\} .
\end{align*}
\end{itemize}
\textbf{Hazard of the subdistribution} 
\begin{center}
  $\gamma_c(t) = {f_c(t)}/[1 - F_c(t)]$
\end{center}	
\begin{itemize}
\item Is not the same as 
$\lambda_c(t) = f_c(t)/[1-F(t)]$,
\item Interpretation tricky!
\end{itemize}


## Warning of "net risk" and "cause-specific survival"

\begin{itemize}                              
\item
The ``\textbf{\textit{net risk}}'' of outcome $c$ by time $t$, 
assuming hypothetical elimination of competing risks,
is often defined as
\begin{center}
$ F_c^*(t) = 1 - S_c^*(t) = 1- \exp\{ - \Lambda_c(t) \} $
\end{center}
\medskip
\item
In clinical survival studies, function 
$S_c^*(t)$ is often called ``\textbf{\textit{cause-specific survival}}'',
and estimated by KM, but treating competing deaths
as censorings.
\medskip
\item
Yet, these *-functions, $ F_c^*(t)$ and $S_c^*(t)$, lack proper probability interpretation when
 competing risks exist.
\medskip
\item
Hence, their use and naive KM estimation should be viewed critically
(Andersen \& Keiding, \textit{Stat Med}, 2012)
\end{itemize}



## Example: Risk of lung cancer by age $a$?

\begin{itemize}
\item 
 Empirical \textbf{cumulative rate} 
CR$(a) = \sum_{k < a} I_k \Delta_k$, i.e.
 ageband-width ($\Delta_k$) weighted 
 sum of empirical 
 age-specific incidence rates $I_k$
 up to a given age $a$ \\
 = estimate of cumulative hazard $\Lambda_c(a)$.
 \medskip
 \item 
 Nordcan \& Globocan give 
``\textbf{\textit{cumulative risk}}'' by 75 y of age, computed from 
$1 - \exp\{-\text{CR}(75)\}$, as an estimate  of the probability of 
getting cancer before age $75$ y, 
assuming that death were avoided by that age. This is based on 
deriving ``net risk'' from cumulative hazard:
\begin{center} 
$ F_c^*(a) = 1 - \exp\{ - \Lambda_c(a) \}. $
\end{center}
\medskip
\item
Yet, cancer occurs in a mortal population.
\medskip
\item
As such CR$({75})$ is a sound age-standardized summary 
 measure for comparing cancer incidence across populations 
 based on a neutral standard population.
\end{itemize}


## Example. Male lung cancer in Denmark

Event-specific hazards $\lambda_c(a)$ by age estimated by age-spec. rates of death and lung ca., resp.

\includegraphics[height=6.5cm]{lung-ca-rates}




## Cumulative incidence of lung cancer by age

\includegraphics[height=6.5cm]{lung-ca-prob}

Both \text{CR} and $1 - \exp(- \text{CR})$ tend to \newline
 overestimate the real cumulative incidence CI after 60 y.


## Non-parametric estimation of CIF 

\begin{itemize}
\item
Let $t_1 < t_2 < \dots < t_K$ be the $K$ distinct 
time points at which any outcome event was observed, \\ Let also
 $\widetilde{S}(t)$ be KM estimator for overall $S(t)$. 
\medskip
\item
\textbf{Aalen-Johansen estimator} (AJ) for the cumulative incidence function $F(t)$
is obtained as
$$ 
\widetilde{F}_c(t) 
   = \sum_{t_k \leq t} \frac{D_{kc}}{n_k} \times \widetilde{S}(t_{k-1}),
   \quad\text{where}
$$
$n_k$ = size of the risk set at $t_k$ $(k=1, \dots, K)$,
\\
$D_{kc}$ = no. of cases of event $c$ observed  at $t_k$.
\medskip
\item
Naive KM estimator $\widetilde{F}^*_c(t)$ of ``net survival'' treats  
competing events occuring first as censorings:  
$$  
\widetilde{F}^*_c(t) = 1 - \widetilde{S}^*_c(t)
  = 1 - \prod_{t_k \leq t} \frac{n_k - D_{kc}}{n_k} 
$$
\end{itemize}



## R tools for competing risks analysis

Package \texttt{mstate}
\begin{itemize}
\item \texttt{Cuminc(time, status, ...)}:  \\
 AJ-estimates (and SEs) for each event type 
  (\texttt{status}, value 0 indicating censoring)
\end{itemize}
Package \texttt{cmprsk}
\begin{itemize}
\item \texttt{cuminc(ftime, fstatus, ...)} 
  computes CIF-estimates,  
   \texttt{plot.cuminc()} plots them. 
  \medskip
\item \texttt{crr()}
 fits Fine--Gray models for 
the hazard $\gamma_c(t)$ of the subdistribution
\end{itemize}   
Package \texttt{Epi} -- \texttt{Lexis} tools for multistate analyses





## Ex. Survival from oral cancer

\begin{itemize}
\item
Creating a {\tt Lexis} object with two outcome events and \\ 
obtaining a summary of transitions
\end{itemize}
\scriptsize
```{r, comment=NA,echo=T,message=FALSE,warning=FALSE,results='hide'}
library(Epi)
orca.lex <- Lexis(exit = list(stime = time), 
           exit.status = factor(event, 
    labels = c("Alive", "Oral ca. death", "Other death") ),
                  data = orca)
summary(orca.lex)  
```
```{r, comment=NA,echo=F,message=FALSE,warning=FALSE}
summary(orca.lex)  
```





## Box diagram for transitions

Interactive use of function \tt boxes().
\small
```{r, comment=NA,echo=T,message=FALSE,warning=FALSE,eval=FALSE}
boxes(orca.lex)
```
\begin{center}
\includegraphics[height=6cm]{orca-boxes}
\end{center}
\normalsize



## Ex. Survival from oral cancer

\begin{itemize}
\item
AJ-estimates of CIFs (solid) for both causes.
\item
Naive KM-estimates of CIF (dashed) $>$ AJ-estimates 
\item
CIF curves may also be stacked (right).  
\end{itemize}

\includegraphics[width=10cm]{orcaCI1}

\textbf{NB.} The sum of the naive KM-estimates of CIF exceeds 100\% at 13 years! 




## Ex. CIFs by cause in men and women

\includegraphics[width=10cm]{orcaCI2}

CIF for cancer higher in women (chance?) but for other causes
higher in men (no surprise).


## Modelling with competing risks


Main options, providing answers to different questions:

\begin{enumerate}[(a)]
\item
  Cox model for event-specific hazards $\lambda_c(t) = f_c(t)/[1-F(t)]$, when \textit{e.g.} the interest is in the biological effect of the prognostic factors on the fatality of the very disease that often leads to the relevant outcome. 
  
  \vspace{0.1in}
  If $\lambda_1(y_i)$ and $\lambda_2(y_i)$ have no common parameters,
  % event-specific hazards 
  they may be fitted separately treating competing events
  as censorings. \\ -- Still, avoid estimating ``net risks'' from 
  $F_c^* = 1 - \exp(-\Lambda_c)$!  
  \bigskip 
\item
 \textbf{Fine--Gray model} for the hazard  of the subdistribution $\gamma_c(t) = f_c(t)/[1-F_c(t)]$ 
  when we want to assess the impact of the factors on the overall cumulative incidence of event $c$.  \\
  -- Function \texttt{crr()} in package \texttt{cmprsk}. 
\end{enumerate}




## Fine-Gray model
   
\begin{itemize}
\item
Regression model for subdistribution hazards
\item
CIF quantifies cumulative disease incidence
\item 
Risk ratio of subdistribution hazards can be used to assess the effects of covariates on CIF 
\item
The same subdistribution hazard model allow one to
make inferences about the relative magnitudes of the effects of the covariates on the incidence of the given type of event
\item 
The direction of the subdistribution hazard ratio denotes the direction but does not directly provide the magnitude of the
effect of the covariate on the CIF.
\end{itemize}


## Estimation: Fine-Gray model vs Cox model

\scriptsize
```{r, comment=NA,echo=T,results='hide'}
orca <- read.table("oralca2.txt", header=T)
orca$agegr <- cut(orca$age,c(0,55,75,Inf),right=F)
table(orca$agegr)
```

\vspace{-1.2cm}

```{r, comment=NA,echo=F}
orca <- read.table("oralca2.txt", header=T)
orca$agegr <- cut(orca$age,c(0,55,75,Inf),right=F)
table(orca$agegr)
```

\normalsize
- Estimation of cause-specific hazards

\scriptsize
```{r, comment=NA,echo=T}
cox1 <- coxph(Surv(time, event==1) ~ sex + agegr + stage, data=orca)
cox2 <- coxph(Surv(time, event==2) ~ sex + agegr + stage, data=orca)
```

\normalsize
- Estimation of subdistribution hazards

\scriptsize
```{r, comment=NA,echo=T}
library(cmprsk)
fg1 <- crr(orca$time, orca$event, cov1 = model.matrix(cox1), failcode=1)
fg2 <- crr(orca$time, orca$event, cov1 = model.matrix(cox2), failcode=2)
```
\normalsize

## Estimation: RRs for death from cancer

\footnotesize
```{r, comment=NA,echo=T}
round( ci.exp(cox1), 2)
round(summary(fg1, Exp=T)$conf.int[,-2], 2)
```


## Estimation: RRs for death from other causes

\footnotesize
```{r, comment=NA,echo=T}
round( ci.exp(cox2), 2)
round(summary(fg2, Exp=T)$conf.int[,-2], 2)
```


## Recommendations for analysing competing risk data
   
\begin{itemize}
\item
Cumulative incidence functions (CIFs) should be used to estimate the
incidence of each of the different types of competing risks. {\bf Do not use the
Kaplan-Meier estimate} of the survival function for this purpose.
\item
Researchers need to decide whether the research objective:
\begin{itemize}
\item
Use the {\bf Fine-Gray} subdistribution hazard model when the focus is on
 estimating {\bf incidence} or predicting {\bf prognosis} in the presence of competing
risks.
\item
Use the {\bf cause-specific hazard} model (e.g. Cox model) when the focus is on addressing
{\bf etiologic} questions.
\end{itemize}
\item
In some settings, both types of regression models should be estimated for
each of the competing risks to permit a full understanding of the effect of
covariates on the incidence and the rate of occurrence of each outcome.
\end{itemize}

