---
title: 
- "Statistical Methods in Cancer Epidemiology using R"
subtitle: Survival model with long term survivors
author: 
- \textbf{Janne Pitkäniemi}  \newline \newline
- Faculty of Social Sciences, University of Tampere  \newline
- Finnish Cancer Registry  \newline  \newline
date: "Feb,17 2020 "
institute: "janne.pitkaniemi@cancer.fi"
output:
  beamer_presentation:
    includes:  
         in_header: beamer_header.tex
---

# Survival model with long term survivors

- In standard survival analysis sufficient follow-up assumed
- Fraction of the study subjects will never experice the event of interest
   - fraction of patients treated will be cured 
   - fraction of population non-susceptible (immune) to event
   
# Survival model with long term survivors

- Let $D$ be partially latent variable indicating
if subject is susceptible, cured $D=1$ and $D=0$ otherwise

- Then the probability of an event for a subject is the product
of probability of beeing susceptible and event at time $t$
$$ P(D=1\mid X_i) f(t \mid D==1,X_i)$$

- It is convienient to specify survivor function
$$ S(t \mid D==1,X_i)=P(T >t \mid D==1,X_i)$$

# Survival model with long term survivors

- Susceptibility can be modelled with any parametric function for binary rv. f.ex logistic
$$ P(D=1\mid X_i) = \frac{exp(\alpha+\beta x_i)}{1+exp(\alpha+\beta x_i)}$$

- Time-to-event with any parametric function exponential, weibull
as well as proportional hazards

- Problems: identifiability between susceptiblity intercept and time-to-event parameters, need more censored observations

- Maller and Zhou presented a testing procedure for susceptiblity fraction 0.

- restrict to problems we consensus is that there is group of non-susceptibles in the population

- separate modelling more informative of the problem if the groups exist

# Survival model with long term survivors


```{r,echo=T,warning=FALSE,message=FALSE,comment=NA, fig.height=4.5}
library(smcure); library(survival) 
data("e1684"); attach(e1684); head(e1684)

#Kaplan Meier estimate of S,CDF
fit <- survfit(Surv(FAILTIME,FAILCENS)~TRT,data = e1684)

#LTS model
pd <- smcure(Surv(FAILTIME,FAILCENS)~TRT,cureform=~TRT,
             data=e1684,model="ph",Var = FALSE)
res<-printsmcure(pd,Var = FALSE)


```
# Survival model with long term survivors
\scriptsize
```{r,echo=FALSE,warning=FALSE,message=FALSE,comment=NA, fig.height=4.5}
printsmcure(pd,Var = FALSE)
```
\normalsize

# Survival model with long term survivors

```{r,echo=FALSE,warning=FALSE,message=FALSE,comment=NA, fig.height=4.5}
# extract suscceptible proportions
lp1<-coef(res$logistfit)[1]
lp2<-sum(coef(res$logistfit))
p1<-exp(lp1)/(1+exp(lp1))
p2<-exp(lp2)/(1+exp(lp2))

plot(fit,fun="event")
abline(h=p1,lty=2)
abline(h=p2,lty=2)
```

- Proportion of immunes in TRT==0 is `r 1-p1`
- Proportion of immunes in TRT==1 is `r 1-p2`
- HR for TRT==1 vs TRT==0 for non-immunes is `r exp(res$beta)`


