---
title: 
- Statistical Methods in Cancer Epidemiology \newline using R
author: 
- \textbf{Karri Seppä}  \newline \newline
- Finnish Cancer Registry  \newline  \newline
- Lecture 7
date: "Mar 9, 2020"
institute: "karri.seppa@cancer.fi"
output:
  beamer_presentation:
    includes:  
         in_header: beamer_header.tex
tables: true
editor_options: 
  chunk_output_type: console
---



## Topics somewhat covered

1. Survival or time to event data & censoring.
\medskip
2. Probability concepts for times to event: \newline 
 survival, hazard and cumulative hazard functions
\medskip
3. Kaplan--Meier and Nelson--Aalen estimators.
\medskip
4. Regression modelling of hazards: Cox model.
\bigskip

Main R functions to be covered (\texttt{survival} package)

- \texttt{Surv(), survfit(), survminer(), coxph()}


## Survival time -- time to event

Let $T$ be the \textbf{time} spent in a given \textbf{state} from its 
beginning till a certain \textit{endpoint} or \textit{outcome} \textbf{event} or \textit{transition}
 occurs, changing the state to another. \newline
 (\texttt{lex.Cst - lex.dur - lex.Xst})

\bigskip
Examples of such times and outcome events:

- lifetime: birth $\rightarrow$ death,
\medskip
- duration of marriage: wedding $\to$ divorce, 
\medskip
- healthy exposure time: \newline start of exposure $\rightarrow$ onset of disease,
  \medskip
\item clinical survival time: \newline
 diagnosis of a disease  $\rightarrow$ death.



## Set-up of classical survival analysis

- \textbf{Two-state model}: only one type of event changes the initial state.
\medskip
- Major applications: analysis of lifetimes
 since birth and of survival times since diagnosis of a disease 
 until death from any cause.

\setlength{\unitlength}{0.7pt}
\begin{picture}(400,80)(-40,70)
  \thicklines
  \put(  0, 80){\framebox(110,50){Alive}}
  \put(240, 80){\framebox(110,50){Dead}}
  \put(125,105){\vector(1, 0){100}}
  \put(170,110){\makebox(0,0)[b]{Transition}}
\end{picture}

 \textbf{Censoring}: Death and final lifetime not observed
  for some subjects 
  due to emigration or closing the follow-up while they are still
 alive. 


## Distribution concepts: survival function

Cumulative distribution function (CDF)
$F(t)$ and density function $f(t) =F'(t)$ of survival time $T$:
\[
F(t) =  P( T \le t) = \int_0^t f(u) du 
\]
= \textbf{risk} or probability that the event occurs by $t$. 

\bigskip
\textbf{Survival function} 
\[ 
S(t) = 1- F(t) =  P( T  >  t) = \int_t^{\infty} f(u)du, 
\]
= probability of avoiding the event at least up to $t$ \newline
(the event occurs
only after $t$).  



## Distribution concepts: hazard function

The \textbf{hazard rate} or \textbf{intensity} function $\lambda(t)$
\begin{align*}
\lambda(t) & = \underset{\Delta \rightarrow 0}{\lim} 
 {P(t < T \le t+\Delta | T > t)}/{\Delta} \\
   & = \underset{\Delta \rightarrow 0}{\lim} 
      \frac{P(t < T \le t+\Delta)/\Delta }{P(T > t)}
     = \frac{f(t)}{S(t)}  
\end{align*}
\begin{enumerate}[$\approx$]
\item  the conditional probability that
the event occurs in a short
 interval $(t, t+\Delta]$, given that it does not
occur before $t$, per interval length. 
\end{enumerate}

In other words, during a short interval
 \begin{center}
 risk of event $\approx$ hazard $\times$ interval length 
 \end{center}


## Distribution: cumulative hazard etc.

The \textbf{cumulative hazard} (or integrated intensity):
\[
\Lambda(t) = \int_0^t \lambda(v)dv
\]


## Observed data on survival times

For individuals $i = 1, \dots, n$ let 
\vspace{-0.3cm}
\begin{enumerate}[ ]
\item $\qquad T_i$ = true time to event,
\item $\qquad U_i$ = true time to censoring.
\end{enumerate}
Censoring is assumed \textbf{noninformative}, \textit{i.e.} independent 
 from occurrence of events.
 
We observe 
\vspace{-0.3cm}
\begin{enumerate}[ ]
\item
$y_i = \text{min}\{ T_i, U_i \}$, \textit{i.e.}
the exit time, and
\item
 $ \delta_{i} = 1_{ \{ T_i < U_i  \} }$, 
  indicator (1/0) for the event occurring first, before censoring. 
\end{enumerate}

Censoring must properly be taken into account in the statistical analysis.



## Approaches for analysing survival time

\begin{itemize}
\item 
\textbf{Parametric models} on hazard rate $h(t)$ \\
(like Weibull, gamma, etc.) -- Likelihood:
\begin{align*} L & = \prod_{i=1}^n \lambda(y_i)^{\delta_i} S(y_i) \\
  & = 
   \exp\left\{ \sum_{i=1}^n 
     [ \delta_i \: \log \: \lambda(y_i) - \Lambda(y_i) ] 
       \right\} 
\end{align*}   
\item 
\textbf{Piecewise constant rate} model on $\lambda(t)$ \\ 
-- estimation of $\hat\lambda(t)$ with poisson regression 
\item 
\textbf{Non-parametric} methods, 
like \\ Kaplan--Meier (KM) 
estimator of survival curve $S(t)$ and \newline Cox 
proportional hazards model.
\end{itemize}



## R package \texttt{survival}

Tools for analysis with one outcome event.

- \texttt{Surv(time,event) -> sobj} \newline 
creates a \textbf{survival object} \texttt{sobj},

- \texttt{survfit(sobj) -> sfo} \newline 
non-parametric survival curve estimates, like KM 
(also estimated baseline in a Cox model), 

- \texttt{plot(sfo)} \newline 
survival curves and related graphs, 

- \texttt{coxph(sobj \~{} x1 + x2 +...)} \newline  
fits the Cox model
with covariates \texttt{x1} and \texttt{x2}. 

- \texttt{survreg()} -- parametric survival models.


## KM esimate for survival function $S(t)$

\begin{itemize}
\item  Order event times (possibly separately in groups)
\item $\widehat{S(t_j)}=\widehat{S(t_{j-1})} \left(1-\frac{d_j}{n_j}\right) $, where $t_0=0$ and
${\widehat{S(0)}=1}$
\item ${\widehat{S(t)}}$ is constant between event -- step function 
\item Each observations contributes as long as at risk for the event and confidence intervals can be introduced using classical inference framework.
\end{itemize}  



## Age-standardised esimate for survival function $S(t)$

\begin{itemize}
\item  Weighted average of age-specific survival estimates
\begin{enumerate}[]
\item $S(t)=\sum_{a=1}^K w_aS_a(t)\quad$ where $\sum_{a=1}^K w_a =1$
\end{enumerate}
\item weight $w_a$ is a standard for the proportion of patients in age group $a$ \underline{at the beginning of follow-up} 
\begin{itemize}
\item e.g. the international cancer survival standards (ICSS; Corazziari et al. 2004) 
\end{itemize}
\item Can be estimated using \texttt{survtab()} function in \texttt{popEpi} package
\end{itemize}  


## Veterans' Administration Lung Cancer study

In this trial, males with advanced inoperable lung cancer were randomized to a standard therapy and a test chemotherapy. The primary endpoint for the therapy comparison was time to death in days, represented by the variable Time.

Variables 

\begin{itemize}
\item trt:	 1=standard 2=test
\item celltype:1=squamous, 2=smallcell, 3=adeno, 4=large
\item time:	 survival time (days)
\item status:	 status 1= death, 0=censored 
\item karno:	 Karnofsky performance score (100=good)
\item diagtime:months from diagnosis to randomisation
\item age:	 in years
\item prior:	 prior therapy 0=no, 1=yes
\end{itemize}

\scriptsize Reference: D Kalbfleisch and RL Prentice (1980), The Statistical Analysis of Failure Time Data. Wiley, New York.



## Veteran data

\scriptsize
```{r, comment=NA,echo=T}
library(survival)
head(veteran)
head(veteran[veteran$trt==2,])
```



## Estimate for hazard function $\lambda(t)$ in R \newline -- splitting time scale

\scriptsize 
```{r, comment=NA,echo=T, warning=F, message=FALSE}
#===============================
# get packages needed for the analysis
#===============================
library(dplyr)
library(survival)
library(Epi)
#===============================
# define stating time at 0 and create subject id
#===============================
veteran$start<-0
veteran$id<-1:nrow(veteran)
#===============================
# Split follow-up time into interval
#===============================
nvet<-survSplit(veteran,
                cut=c(100,200,300,400),
                event="status",
                start="start",
                end="time",episode="period")
```


## Estimate for hazard function $\lambda(t)$ in R \newline -- splitting time scale

\scriptsize 
```{r, comment=NA,echo=T, warning=F, message=FALSE}
#===============================
# dplyr -package,rather than stat.table() -function 
#===============================
gsvet<-group_by(nvet,period)
speriod<-summarize(gsvet, 
               n=length(trt),
               events=sum(status),
               pyrs=round(sum(time-start)),
               rate1000=(events/pyrs)*1000,
               lograte=log(events/pyrs))
speriod<-data.frame(speriod)
speriod[,1:6]
```


## Estimate for hazard function $\lambda(t)$ in R \newline -- Poisson regression

\scriptsize 
```{r, comment=NA,echo=T, warning=F, message=FALSE}
#===============================
# Poisson model for incidence in each period of time
#===============================
m<-glm(events ~ -1+factor(period),
       family=poisson,offset=log(pyrs),data=speriod)
round(ci.lin(m,Exp=TRUE)[,-(3:4)],6)
#1-year cumulative hazard 
Lambda_1yr<- sum(exp(m$coefficient[1:4])*c(100,100,100,65))
#1-year survival in percentages
exp(-Lambda_1yr)*100
```


## Estimation of the survival function $S(t)$


\scriptsize 
```{r, comment=NA,echo=T, warning=F, message=FALSE}
m <- survfit(formula = Surv(time, status) ~ trt, data = veteran)
m2 <- summary(m)
m2 <- data.frame(lapply(c(10,2:6) , function(x) m2[x]))
m2[m2$time<10,]
```

:::::: {.columns}

::: {.column width="50%"}
- trt=1

\begin{enumerate}[ ]
\item $S(0)=1$\\
\item $S(3)=S(0)\times(1-1/69)=0.986$\\
\item $S(4)=S(3)\times(1-1/68)=0.971$
    
\end{enumerate}

:::

::: {.column width="50%"}
- trt=2

\begin{enumerate}[ ]
\item $S(0)=1$\\
\item $S(1)=S(0)\times(1-2/68)=0.971$\\
\item $S(2)=S(1)\times(1-1/66)=0.956$

\end{enumerate}
:::
::::::

## Plot of Survival curve for Veteran data

\scriptsize
```{r, comment=NA,echo=T, warning=F, message=FALSE, fig.height=6.5}
library(survminer)
ggsurvplot(m, pval = TRUE, conf.int = TRUE, risk.table = TRUE,
           risk.table.y.text.col = TRUE)
```


## Estimate cumulative hazard function -- $\hat \Lambda(t)$

\begin{itemize}
\item KM for survival function $P(T>t)$ is often presented \newline
$P(T>t) = S(t)=\exp[-\Lambda(t)]=\exp\left[ -\int_0^t \lambda(u) du \right]$
\item Cumulative risk from $0$ to $t$: $P(T \leq t) = 1-S(t)=1-\exp[- \Lambda(t)]$
\item If low incidence rate $\lambda$ or short risk period $\Delta$: \newline  $1-\exp[- \Lambda(t)] \approx \Lambda(t) = \lambda \Delta $ \quad i.e. rate $\times$ period at risk
\item Cumulative hazard can be estimated from KM, but Nelson-Aalen should be preferred
\end{itemize}


## Estimate cumulative hazard function -- $\hat \Lambda(t)$

\scriptsize
```{r, comment=NA,echo=T, warning=F, message=FALSE, fig.height=3.5, dev.args=list(pointsize=23)}
m <- survfit(formula = Surv(time, status) ~ trt, data = veteran)
par(mfrow=c(1,3), mar=c(3,4,0,0.5))
plot(m, fun="F",ylab="1-S", col=c("blue","red"))
plot(m, fun="cumhaz", ylab="Cumulative hazard", col=c("blue","red"))
plot(m, fun="cloglog", ylab="Logarithm of cumulative hazard", col=c("blue","red"))
```

\vspace{-0.2cm}
\footnotesize

- KM curve of survival S(t) is the most popular.
- Informative are also graphs for estimates of
    - $F(t) = 1 - S(t)$ , i.e. CDF
    - $\Lambda(t) = -\log[1-F(t)]$, cumulative hazard,
    - $\log[\Lambda(t)]$, cloglog transform of CDF.


## Regression models for time-to-event data

Consider only one outcome & no competing events
\begin{itemize} 
\item
Subject $i$ $(i=1, \dots, n)$ has an own vector $x_i$ that contains
values $(x_{i1}, \dots, x_{ip})$ of a set of $p$ 
continuous and/or binary covariate terms.
\medskip
\item
In the spirit of generalized linear models we let 
$\beta = (\beta_1, \dots, \beta_p)$ be regression coefficients
and build  a \textbf{linear predictor} 
$$ 
\eta_i = x_i^{\small\text{T}} \beta =  \beta_1 x_{i1} + \dots + \beta_p x_{ip} 
$$ 
\item
Specification of outcome variable? \newline
Distribution (family)? Expectation? Link?  
\end{itemize}


## Regression models (cont'd)

Survival regression models can be defined \textit{e.g.} for 

(a) survival times directly 
$$ \log(T_i) = \eta_i + \epsilon_i, \quad\text{s.t. } 
\epsilon_i \sim F_0(t; \alpha)$$ 
where $F_0(t; \alpha)$ is some baseline model, 
\medskip
(b) hazards, multiplicatively: $$ 
\lambda_i(t) = \lambda_0(t; \alpha) r(\eta_i), \quad\text{where}$$
$\lambda_0(t; \alpha)$ = baseline hazard and \newline
$r(\eta_i)$ = relative rate function, typically $\exp(\eta_i)$
\medskip
(c) hazards, additively: 
$$ \lambda_i(t) = \lambda_0(t; \alpha) + \eta_i. $$



## Relative hazards model or Cox model

In model (b), the baseline hazard $\lambda_0(t,\alpha)$ may be given a parametric form (\textit{e.g.} Weibull) or
a piecewise constant rate (exponential) structure.

\bigskip
Often a parameter-free form $\lambda_0(t)$ is assumed. Then
\[
  \lambda_i(t) = \lambda_0(t) \exp(\eta_1),
\]
specifies the \textbf{Cox model} or the \textbf{semiparametric proportional hazards model}.

\bigskip
$\eta_i = \beta_1 x_{i1} + \dots + \beta_p x_{ip}$ not depending on time.  

\bigskip
Generalizations: \textbf{time-dependent} \newline covariates $x_{ij}(t)$, and/or 
effects $\beta_j(t)$.



## PH model: interpretation of parameters

Present the model explicitly in terms of $x$'s and $\beta$'s.
\[
\lambda_i(t) = \lambda_0(t)  \exp({\beta_1 x_{i1} + \dots +
\beta_p x_{ip}})
\]
Consider two individuals, $i$ and $i'$, having the same values of all
other covariates except the $j^{\text{th}}$ one.

\bigskip
The ratio of hazards is constant:
$$  \frac{\lambda_i(t)}{\lambda_{i'}(t)} = \frac{\exp( \eta_{i}) }{\exp(\eta_{i'})}
= \exp \{ \beta_j(x_{ij}-x_{i'j}) \} . $$
Thus $e^{\beta_j} = \text{HR}_j$ = \textbf{hazard ratio} or relative rate
 associated with
 a unit change in covariate $X_j$.


## Ex. Veteran data and treatment effect

Fitting Cox models with trt effect.
\scriptsize
```{r, comment=NA,echo=T, warning=F, message=FALSE, fig.height=4.5}
m <- coxph(formula = Surv(time, status) ~ trt, data = veteran)
summary(m)
```

\normalsize
HR for treatment 2 vs. 1 is 1.02 (95% CI 0.71;1.45) \newline
Not statistically significant (p=0.92)



## Proportionalilty of hazards?
\begin{itemize}
\item
Consider two groups $g$ and $h$ defined by one categorical covariate, and let $\rho > 0$. 
\medskip
If $\lambda_g(t) = \rho \lambda_h(t)$  then 
$\Lambda_g (t) = \rho \Lambda_h(t)$ and
$$ \log\: \Lambda_g  (t)  = \log (\rho) + \log\: \Lambda_h (t), $$
thus log-cumulative hazards should be parallel!
\end{itemize}
\begin{enumerate}[$\Rightarrow$]
\item
\textit{Plot the estimated log-cumulative hazards and see
whether they are sufficiently parallel}.
\end{enumerate}
\begin{itemize}
\item
\texttt{plot(coxobj, ..., fun = 'cloglog')}
\medskip
\item
Testing the proportionality assumptions: \texttt{cox.zph(coxobj)}.
\end{itemize}




## Ex. Veteran data - test PH

\begin{itemize}
\item With $>1$ covariates,  \texttt{cox.zph()} tests the assumption by checking, whether the corresponding parameters (\& hazard ratios)
may vary in time.
\item Suppose that I want to include information on patient baseline general disease status -- Karnofsky performance score (0=dead, 100=good)
\item Dichotomize the Karnofsky score 0 if Score $[$0,50 $]$ and 1 if $($50,100$]$
\end{itemize}




## Ex. Veteran data - test PH

\scriptsize 
```{r, comment=NA,echo=T, warning=F, message=FALSE, fig.height=4.5}
veteran$karnod<-as.numeric(veteran$karno>50)
m <- coxph(formula = Surv(time, status) ~ trt+ karnod, 
           data = veteran)
m
#testing proportionality
cox.zph(m)
```

Test for proportionality are significant (p<0.05) -- assumptions of proportionalty of hazards is rejected for both tretment and score variable
-- stratify accoding to the score 



## Ex. Veteran data - test PH

\scriptsize 
```{r, comment=NA,echo=T, warning=F, message=FALSE, fig.height=4.5}
m <- coxph(formula = Surv(time, status) ~ trt+strata(karnod), 
           data = veteran)
m
cox.zph(m)
```

Test for proportionality is not significant (p>0.05) \newline
HR for treatment effect is 1.01 95% CI (0.71 ;1.45)


## Homogeneity of HRs

Question: Are the HRs for different celltypes similar or not?

Testing hypothesis of regression coefficients equal (Not the hypothesis that they are zero)

More formally, the model: 
\[
\lambda_i(t) = \lambda_0(t)  \exp({\beta_1 x_{i1} + \beta_2 x_{i2} + \beta_3 x_{i3} }),
\]
where $x_{i1}=1$ if celltype smallcell 0 otherwise \newline
and $x_{i2}=1$ if celltype adeno 0 otherwise \newline
and $x_{i3}=1$ if celltype large 0 otherwise \newline
and squamous celltype has been chosen as the reference category
($x_{i1}=0$ and $x_{i2}=0$ and $x_{i3}=0$)





## Homogeneity of HRs

\scriptsize 
```{r, comment=NA,echo=T, warning=F, message=FALSE, fig.height=4.5}
m<-coxph(formula = Surv(time, status) ~ celltype, 
         data = veteran)
summary(m)
```

## Homogeneity of HRs

Test for homogeneity across HRs of three cell types: small cell, adeno and large ($\beta_1=\beta_2=\beta_3$?)

\scriptsize 
```{r, comment=NA,echo=T, warning=F, message=FALSE, fig.height=4.5}
veteran$celltype2 <- Relevel(veteran$celltype,           
                             list(1,"small_or_adeno_or_large"=2:4))
m2<-coxph(formula = Surv(time, status) ~ celltype2, 
         data = veteran)
anova(m,m2)
```

HR’s are not the same between celltypes

## Homogeneity of HRs

Test for homogeneity between HRs of smallcell and adeno cancer ($\beta_2=\beta_3?$).

\scriptsize 
```{r, comment=NA,echo=T, warning=F, message=FALSE, fig.height=4.5}
veteran$celltype3 <- Relevel(veteran$celltype,           
                             list(1,"small_or_adeno"=2:3,4))
m3<-coxph(formula = Surv(time, status) ~ celltype3, 
         data = veteran)
anova(m,m3)
```

The HRs do not differ significantly between small cell and adeno cancers (p=0.56).  
