---
title: 
- Statistical Methods in Cancer Epidemiology using R
author: 
- \textbf{Janne Pitkäniemi}  \newline \newline
- Faculty of Social Sciences, University of Tampere  \newline
- Finnish Cancer Registry  \newline  \newline
- Lecture 2b
date: "Feb,17 2020 "
institute: "janne.pitkaniemi@cancer.fi"
output:
  beamer_presentation:
    includes:  
         in_header: beamer_header.tex
---

# Confounding and effect modification

Consider another **factor** $Z$ which is

- also a risk factor of the outcome,
- possibly associated with exposure $X$ in study population,
- not a causal consequence of $X$.


$\Rightarrow$ Adjustment for possible *confounding* and evaluation of *effect modification* needed.


# Confounder

- **Confounder**s (Z) are variables that 
   - is both an ancestor of the exposure (X) 
   - and an ancestor of the outcome (Y)
   (along a path that does not include the exposure). 


```{r,echo=FALSE,warning=FALSE,message=FALSE,comment=NA, fig.height=4.5}
library(ggdag)
theme_set(theme_dag())
dagify(y ~ x,
       y ~ z,
       x ~ z) %>% 
  ggdag()

```

# Effect modification 

- Unfortunately, there's not an agreed upon way to show effect modification/moderation/interaction. 
- effects are assumed to be heterogeneous by default in causal models. (JP)

```{r,echo=FALSE,warning=FALSE,message=FALSE,comment=NA, fig.height=4.5}

library(tidyverse)

coords <- tribble(
  ~name,    ~x,  ~y,
  "x",      1,   1,
  "y",      2,   1,
  "m",      1.5, 0,
  "holder", 1.5, 1.05
)

dagify(
  y ~ x,
  holder ~ m,
  coords = coords
) %>% 
  ggplot(aes(x = x, y = y, xend = xend, yend = yend)) + 
    geom_dag_point(data = function(x) filter(x, name != "holder")) +
    geom_dag_edges() + 
    geom_dag_text(data = function(x) filter(x, name != "holder")) +
    theme_dag()
```


# Example: OS vs. PN (cont'd)

Failure of treatment depends also on initial condition
of patient, like extent and severity of disease.

Results stratified by initial diameter size of the stone:

\begin{center}
\begin{tabular}{lcrrcc}  \hline
Size   &   Trt        & Fails  & Npats & Fail-\% & {\bf RD(\%)}\\ \hline
$<$ 2 cm  &  OS                &         6        &  87        &        {\em 6.9}  & { } \\
 & PN                &        35        & 270        &        {\em 13.0} & {\bf $-$6.1} \\
$\geq$ 2 cm  &  OS         &          71 &        263        & {\em 27.0} & { }  \\
  &  PN         &          25 &         80        & {\em 31.3} & 
  {\bf $-$4.3} \\ \hline
\end{tabular}
\end{center}

OS seems more successful in both subgroups, even though overall PN appeared better.

*IS THERE A PARADOX?*

# Confounding

Solution:
Treatment groups are not *comparable*
w.r.t. initial size.
Size of the stone is a *confounder*
of the association between operation type
and failure risk,because it is

- an *independent determinant* of outcome (failure), based on
external knowledge,
-  *statistically associated* with operation type in the
   study population,
- *not causally affected* by operation type.


This is an instance of ``confounding by indication'':

- patient status
affects choice of treatment 
  $\to$ *bias* in comparing treatments.


This bias would be best to avoid in planning: \newline
$\rightarrow$  *randomized allocation* of treatments!

# Stratified analysis

Stratification of cohort data with proportions 

- at each level $k$ of factor $Z$ results are summarized:

\begin{center}
\begin{tabular}{cccc} \hline
Exposure to &  Number of &  Number of &  Group\\
risk factor &  cases &  non-cases  &  size\\ \hline
yes        &                $D_{1k}$ &    $C_{1k}$  &    $N_{1k}$ \\
no         &                $D_{0k}$ &    $C_{0k}$  &    $N_{0k}$ \\ \hline
Total        &                $D_{\+ k}$  &  $C_{\+ k}$  &   $N_{\+ k}$ \\ \hline
\end{tabular}
\end{center}

Stratum-specific incidence proportions by exposure group: 
$$ R_{1k}  = \frac{ D_{1k} }{ N_{1k} } , \qquad
   R_{0k}  = \frac{ D_{0k} }{ N_{0k} } $$

# Adjusted estimation of risk difference


- Let $\pi_{jk}$ be true risk in exposure group $j$ ($j=0,1)$ as to $X$
and stratum $k$ ($k=0, \dots, K$) of $Z$. Let also
$$ \theta_k = { \pi_{1k} } - { \pi_{0k} } $$
be the risk difference in stratum $k$.

- Many approaches to combine stratum specific results 
into one summary estimator that adjusts for confounding.

These are all somehow `weighted averages` of stratum-specific
estimators.

- Different weighting principles:
  
  - Maximum likelihood (ML), 
  - Mantel-Haenszel (MH) weights,
  - Standardization either by external standard population 
  or by ``indirect'' standardization.
  


# Model-based adjustment of risk difference

Define generalized linear model for binary outcome with

- one binary exposure variable $X$ and 
- one binary stratifying factor or covariate $Z$ (easily generalized to polytomous factors).


*Random part*: Number of cases $D_{jk}$ in exposure group $j$ ($j=0,1$) of $X$ 
and level $k$ $(k = 0, 1)$ of $Z$ is assumed to be binomially distributed
$$D_{jk} \sim \text{Binomial}(N_{jk} ; \pi_{jk} ) , $$



# Model-based adjustment of risk difference (cont'd)

*Systematic part*: 
$$ \pi_{jk} = \alpha + \beta X_j +
  \gamma Z_k  , $$
where $X_j$ is 0/1-indicator as before, and

- $Z_k =$ 1 for level $k=1$ of $Z$, otherwise $Z_k =$ 0,
- $\alpha= \pi_{00}$ = baseline (''corner cell'') risk, 
- $\gamma= \pi_{01}- \pi_{00} = \pi_{11} - \pi_{10}$,
- $\beta= \pi_{10} - \pi_{00} = \theta_0$
                = $\pi_{11} - \pi_{01} = \theta_1$,


*How do we read this?*

# Model-based adjustment of risk difference (cont'd)

Implications of model definition

- the model assumes *homogeneity* of true risk difference $\theta$ associated with factor $X$ (exposed vs. unexposed) across levels of $Z$: $\theta_1 = \theta_0 = \beta$,

- inclusion of $Z$ in the model leads to adjustment of it when estimating the ``true'' effect $\theta$ of 
      $X$,
- $\gamma$ = risk difference between levels $1$ and 0 of $Z$; this
      is same in both exposure groups ($j=0,1$)  \newline
      $\Rightarrow$ homogeneity of the effect of $Z$ is assumed, too.

# Example. Treatment of renal calculi (cont'd)


- Define new variable \newline 
`size` = initial stone size (0 for `small`, 1 for `large`)
- Extended data matrix comprises four observational units (rows) and four variables (columns):
\small
\begin{verbatim}
     size   trt   fails   npats
       0      1       6      87   
       0      0      35     270
       1      1      71     263 
       1      0      25      80  
\end{verbatim}
\normalsize

- These may be read in as before, *e.g.*
\small
```{r}
library(Epi)
size <- c( 0, 0, 1, 1) ; trt <- c( 1, 0, 1, 0)
fails <- c( 6, 35, 71, 25)
npats <- c( 87, 270, 263, 80) 
props <- fails/npats
```


# Fitting model for adjusted risk difference
 
As before, but model formula supplemented by `+ size` 

```{r, comment=NA}
RDmod2 <- glm( props ~ trt + size, 
      fam = binomial(link='identity'), w = npats)
round( ci.lin(RDmod2), 3)
```
\normalsize

Reading the results:

- $\widehat\alpha\ $ = **0.128** = $\widehat\pi_{00}$, fitted baseline risk,
- $\widehat\gamma\ $ = **0.195**, RD between large and small stones,
- $\widehat\beta\ $ = $-$ **0.0561** [$-$ **0.114**,
 **0.002** ], estimated \newline $\qquad$ common
  treatment effect $\widehat \theta$ for OS *vs.* PN.\newline
$\quad$ = Weighted average of $\widehat \theta_0 = -0.061$ and $\widehat \theta_1 = -0.043$.



# Effect modification

Homogeneity assumption -- 
true differences were put equal:
$$ \theta_k = \pi_{1k} - \pi_{0k} = \theta $$ 
across all levels $k$ of covariate $Z$. *Is this realistic?*

*Example.* Is the true risk difference for treatment failure 
between OS
and PN similar for small and big stones?

Empirical differences of failure proportions were


small stone: $\widehat\theta_0$ =  0.069 - 0.130 = **-0.061** \newline
large stones: $\widehat\theta_1$  =  0.270 - 0.313 = **-0.043**


Is the contrast $-0.043 - (-0.061) = 0.018$
between these differences due to chance only, or
is there essential \newline 
*effect modification* present?

# Modelling modification of risk difference

The random part is the same, but the systematic part is
\begin{center}
$ \pi_{jk} = \alpha + \beta X_j +
  \gamma Z_k + \tau U_{jk} . $ 
  \end{center}
 
- $U_{jk}= X_j \times Z_k$, product of values of $X$ and $Z$,
- $\alpha=  \pi_{00}$ = baseline (''corner cell'') risk, 
- $\beta= \pi_{10} - \pi_{00} = \theta_0, \qquad \gamma = \pi_{01}- \pi_{00}$,
- $\tau= \theta_1 - \theta_0$ 
        = $(\pi_{11} - \pi_{01}) - ( \pi_{10} - \pi_{00})$,\newline
        $\qquad$ *interaction parameter*

*$\tau$* describes, how much greater is the risk difference 
between levels 1 and 0 of risk factor $X$ among those at
level 1 of factor $Z$ than in those at level 0.

# Fitting model with modification

- Generation of an interaction of *product term*:
- trtsize = size*treat
- Expanded and rearranged data matrix: \small
\begin{verbatim}
 fails   npats       size       trt    trtsize
   35     270          0         0         0
    6      87          0         1         0
   25      80          1         0         0
   71     263          1         1         1
\end{verbatim} 
\normalsize

```{r}
size <- c( 0, 0, 1, 1) ; 
trt <- c( 0, 1, 0, 1);
trtsize <- c(0,0,0,1)
fails <- c( 35, 6, 25, 71)
npats <- c( 270, 87, 80, 263) 
props <- fails/npats

```

# Fitting model with modification (cont'd)

Fitting the model including the product term: 

```{r}
RDmod3 <- glm(props ~ trt + size + trtsize,
               fam = binomial(link='identity'), w = npats)
```

Results and interpretation:

\small
```{r,comment=NA}
round( ci.lin(RDmod3)[ , -(3:4)], 4)
```


\normalsize

- $\widehat \beta = -0.061 = \widehat \theta_0 =$ RD for OS *vs.* PN in small stones,
- $\widehat \gamma = -0.183 =$ RD btw large and small stones for OS.
- estimate [95 \% CI] of the interaction parameter:
$$\widehat \tau = {\bf 0.0181 [-0.115, 0.151]}$$
\item[{ }] $\Rightarrow$ no evidence for essential modification of risk difference.

Interpretation

This model is **saturated**: It has as many 
coefficients as there are observations. Hence

- residual df = 0,
- fitted cell probabilities = observed proportions:
   
- $\widehat\pi_{00}$  = $0.130$, 
- $\widehat\pi_{10}$  = $0.130 - 0.061 = 0.069$,
- $\widehat\pi_{01}$  = $0.130 + 0.183 = 0.313$,
- $\widehat\pi_{11}$  = $0.130 - 0.061 + 0.183 + 0.018$ 
  = 0.027
- residual $X^2$ and deviance are both 0.

# Final comments

- When risk ratio $\phi$ or odds ratio $\psi$ is the parameter of interest, adjustment for confounding and evaluation of modification can be done by fitting an analogous binomial GLM with relevant link function.

- Modelling can easily be extended to cover one or more polytomous and/or continuous covariates.  Flexible functional forms may be specified to describe the effects of the latter type of variables.

- Binomial models are not limited to grouped data but may be fitted on individual data with binary outcomes, too.

- With more complicated models, especially involving continuous variables, the identity  link (sometimes log link, too) violates the basic range restriction:\\
outcome probabilities $\pi$ must remain within 0 and 1. 



