---
title: 'Epidemiologic data analysis using R '
author:
- Faculty of Social Sciences, University of Tampere
- ' -- '
- Janne Pitkaniemi / Karri Seppä
date: 9.3.2020
output:
  word_document: default
  pdf_document: default
subtitle: Practicals 7
editor_options: 
  chunk_output_type: console
---


```{r p7_0, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, comment = NA)
options(warn=-1)
setwd("C:/Users/karri.seppa/Documents/TRE")
```


# Topics of practical 7



Learning objectives of this practical

*   Estimating Kaplan-Meier (``survfit``) and lifetable (``survtab``)
*   Estimating age-standardised survival (``survtab``)
*   Use proportional hazards model (``coxph`` and ``glm``)
*   Test for the assumption of proportional hazards


# Survival analysis: Oral cancer patients


## Description of the data


File ``oralca2.txt`` contains data from 338
patients having an oral squamous cell carcinoma diagnosed and treated
in one tertiary level oncological clinic in Finland since 1985, followed-up
for mortality until 31 December 2008.
The dataset contains the following variables:

variable    | description
----------- | --------------------
sex         |a factor with categories: 1 = "Female", 2 = "Male",    
age         |in years at the date of diagnosing the cancer,  
stage       |TNM stage of the tumour (factor): 1 = "I", ..., 4 = "IV", 5 = "unkn", 
time        |follow-up time (in years) since diagnosis until death or censoring,
event       |status at the end of follow-up (numeric): \newline 0 = censoring alive, 1 = death from oral cancer, 2 = death from other causes. 

            

## Loading the packages and the data


Load the R packages ``Epi``, ``mstate``, and ``survival`` needed in this exercise.
```{r p7_1, eval=TRUE,echo=TRUE,warnings=FALSE}
library(Epi)
library(mstate)
library(survival)
```

Read the datafile ``oralca2.txt`` into an R data frame named orca. 

Look at the head, structure and the summary of the data frame. Using function ``table()`` count the numbers of censorings as well as deaths from oral cancer and other causes, respectively, from the event variable.

```{r p7_2, eval=TRUE,echo=TRUE,warnings=FALSE}
orca <- read.csv("oralca2.txt", sep="")
head(orca)
```

## Total mortality: Kaplan--Meier analyses


We start our analysis of total mortality pooling the two causes of death into
a single outcome.
First, construct a *survival object* ``orca$suob`` from
the event variable and the follow-up time using function ``Surv()``.
Look at the structure and summary of ``orca$suob``.

```{r p7_3, eval=TRUE,echo=TRUE,warnings=FALSE}
# all deaths
orca$suob <- Surv(orca$time, 1*(orca$event > 0) )
str(orca$suob)
summary(orca$suob)
```


Create a survfit object ``s.all``, which does the
default calculations for a Kaplan--Meier
analysis of the overall (marginal) survival curve.

```{r p7_4, eval=TRUE,echo=TRUE,warnings=FALSE}
s.all <- survfit(suob ~ 1, data=orca)
```

See the structure of this object and apply ``print()`` method on it, too. Look at the results; what do you find?

```{r p7_5, eval=TRUE,echo=TRUE,warnings=FALSE}
s.all
str(s.all)
```

The ``summary`` method for a  ``survfit`` object would return a lengthy life table. However, the ``plot`` method with default arguments offers the Kaplan--Meier curve
for a conventional illustration of the survival experience in the whole patient group.

Alternatively, instead of graphing survival proportions,
one can draw a curve describing their complements: the cumulative mortality proportions. This curve is drawn together with the survival curve as the
 result of the second command line below.

```{r p7_6, eval=TRUE,echo=TRUE,warnings=FALSE}
plot(s.all)
lines(s.all, fun = "event", mark.time=F, conf.int=F)
```

The effect of option ``mark.time=F`` is to omit
marking the times when censorings occurred.

Alternatively, use lifetable (actuarial) method in survtab function of popEpi package

```{r p7_61, eval=TRUE,echo=TRUE,warnings=FALSE}
library(popEpi)
library(Epi)

lex <- Lexis(exit = list(FUT = time),
             exit.status = (event>0)*1,
             data = orca[,-6], merge= T)

# lifetable estimated in annual intervals from 0 to 25 years
st <- survtab(Surv(time = FUT, event = lex.Xst) ~ 1, data = lex, 
              surv.method = "lifetable", surv.type = "surv.obs", 
              conf.type = "log", breaks = list(FUT = seq(0, 25, 1)))
```

Compare Kaplan-Meier and lifetable estimates

```{r p7_612, eval=TRUE,echo=TRUE,warnings=FALSE}
plot(s.all)
lines(st, col="red")
```

Lifetable for the first 5 annual intervals:

```{r p7_62, eval=TRUE,echo=TRUE,warnings=FALSE}
summary(st, t=1:5)
```


## Total mortality by stage

Tumour stage is an important prognostic factor in cancer survival studies.

Plot separate cumulative mortality curves for the different stage groups
marking them with different colours, the order which you may define yourself.
Also find the median survival time for each stage.

```{r p7_7, eval=TRUE,echo=TRUE,warnings=FALSE}
s.stg  <- survfit(suob ~ stage, data= orca)
col5 <- c("green", "blue", "black", "red", "gray")
plot(s.stg, col= col5, fun="event", mark.time=F )
s.stg
```



Create now two parallel plots of which the first one describes the cumulative hazards and the second one graphs the log-cumulative hazards against log-time
for the different stages. Compare the two presentations
with each other and with the one in the previous item.

```{r p7_8, eval=TRUE,echo=TRUE,warnings=FALSE,fig.width=7,fig.height=5,dev.args=list(pointsize=15)}
par(mfrow=c(1,2),mar=c(3,3,2,1))
plot(s.stg, col= col5, fun="cumhaz", main="cum. hazards" )
plot(s.stg, col= col5, fun="cloglog", main = "cloglog: log cum.haz"  )
#a straight line with a slope of 1
lines(exp(seq(-4,1,0.1)+3),seq(-4,1,0.1),lty=2)
```


If the survival times follow the exponential distribution
in a given (sub)population,
the corresponding cloglog-curve (plotted against time on the logarithmic scale) should be approximately a straight line with a slope of 1 (a straight line with any slope, if survival times follow the Weibull distribution).
Could this be the case here in the different stages?

Also, if the survival distributions of the different subpopulations would
obey the *proportional hazards* model, the vertical distance between the
cloglog-curves should be approximately constant over the time axis.
Do these curves indicate serious deviation from the proportional hazards assumption?

## Does survival differ between males and females?

```{r p7_101, eval=TRUE,echo=TRUE,warnings=FALSE}
coxph(Surv(lex.dur, lex.Xst) ~ sex, data=lex)
```

No clear effect in survival between males and females was observed based on this simple analysis. 

Next, we shall examine the confounding by age.
First categorize the continuous age variable
into, say, three categories by function ``cut()``
using suitable breakpoints, like 55 and 75 years, and
cross-tabulate sex and age group:

```{r p7_9, eval=TRUE,echo=TRUE,warnings=FALSE}
orca$agegr <- cut(orca$age, br=c(0,55,75,95), right=F)
stat.table( list( sex, agegr), list( count(), percent(agegr) ),margins=T, data = orca )
```

Male patients are clearly younger than females in these data.

There is evidence on the effect of sex after adjusting for age at diagnosis.

```{r p7_102, eval=TRUE,echo=TRUE,warnings=FALSE}
lex$agegr <- cut(lex$age, br=c(0,55,75,95), right=F)
m.adj <- coxph(Surv(lex.dur, lex.Xst) ~ sex+agegr, data=lex)
m.adj
```

Is proportionality of hazards a reasonable assumption?

```{r p7_103, eval=TRUE,echo=TRUE,warnings=FALSE}
cox.zph(m.adj)
```


Now, plot Kaplan--Meier curves jointly classified by sex and age.

```{r p7_10, eval=TRUE,echo=TRUE,warnings=FALSE}
s.agrx <- survfit(suob ~ agegr + sex, data=orca)
par(mfrow=c(1,1))
plot(s.agrx, fun="event", mark.time=F, xlim = c(0,15),
          col=rep(c("red", "blue"),3), lty=c(2,2, 1,1, 5,5))
```

In each ageband the mortality curve for males is on a higher level than that for females.

Now, estimate the age-standardised survival using the distribution of age at diagnosis as the standard:

```{r p7_104, eval=TRUE,echo=TRUE,warnings=FALSE}
lex <- Lexis(exit = list(FUT = time),
             exit.status = (event>0)*1,
             data = orca[,-6], merge= T)
w <- table(orca$agegr)

st <- survtab(Surv(time = FUT, event = lex.Xst) ~ sex + adjust(agegr), 
              data = lex, weights = c(w), 
              surv.method = "lifetable", surv.type = "surv.obs", 
              conf.type = "log", breaks = list(FUT = seq(0, 15, 1/12)))

st.non.stand <- survtab(Surv(time = FUT, event = lex.Xst) ~ sex, 
              data = lex, 
              surv.method = "lifetable", surv.type = "surv.obs", 
              conf.type = "log", breaks = list(FUT = seq(0, 15, 1/12)))

plot(st, col=c("blue","red"),conf.int = F)
lines(st.non.stand, col=c("blue","red"),lty=2,conf.int=F)
legend("topright",c("Men, age-standardised","Women, age-standardised",
                    "Men, non-standardised","Women, non-standardised"),
       col=rep(c("blue","red"),2),lty=c(1,1,2,2))
```


Use piecewise constant hazards model instead of Cox model. Split the follow-up time into annual intervals from 0 to 5 years, and thereafter, into 5-year intervals.

Could hazard be a constant from 0 to 15 years (exponentially distributed survival times)?

```{r p7_105, eval=TRUE,echo=TRUE,warnings=FALSE}
lexF <- splitLexis(lex, c(seq(0, 5, 1),seq(10, 15, 5)), time.scale = "FUT")
#exclude follow-up>15 years
lexF <- subset(lexF, FUT<15)
lexF$FUT <- cut(lexF$FUT, c(seq(0, 5, 1),seq(10, 15, 5)),right=F)

# piecewise constant hazards
m.pc <- glm(lex.Xst ~ FUT + sex + agegr, offset = log(lex.dur),
              data = lexF, family = "poisson")

# constant hazard model
m.c <- glm(lex.Xst ~ sex + agegr, offset = log(lex.dur),
              data = lexF, family = "poisson")

anova(m.c,m.pc,test="LRT")
```

Can we assume proportional hazards between age groups, i.e. no interaction between follow-up time and age at entry?

```{r p7_106, eval=TRUE,echo=TRUE,warnings=FALSE}
# main effects only
m.pc.age <- glm(lex.Xst ~ FUT + sex + agegr, offset = log(lex.dur),
              data = lexF, family = "poisson")

#interaction included
m.pc.age.fu <- glm(lex.Xst ~ FUT + sex + agegr + FUT:agegr, offset = log(lex.dur),
              data = lexF, family = "poisson")

anova(m.pc.age,m.pc.age.fu,test="LRT")
```

The effect of sex adjusted for age based on the piecewise constant hazards model


```{r p7_107, eval=TRUE,echo=TRUE,warnings=FALSE}
summary(m.pc.age)
```


