---
title: 'Epidemiologic data analysis using R '
author:
- Faculty of Social Sciences, University of Tampere
- ' -- '
- Janne Pitkaniemi
date: "`r format(Sys.time(), '%d.%m.%Y')`"
output:
  word_document: default
  pdf_document: default
subtitle: Practicals 7
---


```{r p7_0, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, comment = NA)
setwd("C:/Users/janne.pitkaniemi/Projects/TRE2018")
```


# Topics of practical 7



Learning objectives of this practical

*   Estimating CIF
*   use proportional hazards model when competing risk


#Survival analysis: Oral cancer patients


## Description of the data


File \verb|oralca2.txt|, that you may
access from a url address to be given in the practical,
contains data from 338
patients having an oral squamous cell carcinoma diagnosed and treated
in one tertiary level oncological clinic in Finland since 1985, followed-up
for mortality until 31 December 2008.
The dataset contains the following variables:

\medskip
\begin{tabular}{lcl}
{\tt sex}         &=& sex, a factor with categories
    \verb|1 = "Female", 2 = "Male"|,    \\
{\tt age}         &=& age (years) at the date of diagnosing the cancer,  \\
% loc  &-& location of the tumour (factor):
% \verb|1 = " Other", 2 = " Tung"|, \\
{\tt stage}       &=& TNM stage of the tumour (factor): \verb|1 = "I", ..., 4 = "IV", 5 = "unkn"|, \\
{\tt time}       &=& follow-up time (in years) since diagnosis until death or censoring,\\
{\tt event}      &=& event ending the follow-up (numeric): \\
{ }         & & 0 = censoring alive, 1 = death from oral cancer, 2 = death from other causes. \\
\end{tabular}


## Loading the packages and the data


Load the R packages \texttt{Epi, mstate}, and \texttt{survival} needed in this exercise.
```{r p7_1, eval=TRUE,echo=TRUE,warnings=FALSE}
library(Epi)
library(mstate)
library(survival)
```

Read the datafile {\tt oralca2.txt} from a website, whose precise address  will be given in the practical, into an R data frame named \texttt{orca}. 

Look at the head, structure and the summary of the data frame. Using function \texttt{table()} count the numbers of censorings as well as deaths from oral cancer and other causes, respectively, from the \texttt{event} variable.

```{r p7_2, eval=TRUE,echo=TRUE,warnings=FALSE}
orca <- 
  read.csv("C:/Users/janne.pitkaniemi/Projects/TRE2018/oralca2.txt", 
           sep="")
head(orca)
```

## Total mortality: Kaplan--Meier analyses


We start our analysis of total mortality pooling the two causes of death into
a single outcome.
First, construct a {\it survival object} \verb!orca$suob! from
the event variable and the follow-up time using function \verb+Surv()+.
Look at the structure and summary of \verb!orca$suob! .

```{r p7_3, eval=TRUE,echo=TRUE,warnings=FALSE}
# all deaths
orca$suob <- Surv(orca$time, 1*(orca$event > 0) )
str(orca$suob)
summary(orca$suob)
```


Create a {\tt survfit} object {\tt s.all}, which does the
default calculations for a Kaplan--Meier
analysis of the overall (marginal) survival curve.

```{r p7_4, eval=TRUE,echo=TRUE,warnings=FALSE}
s.all <- survfit(suob ~ 1, data=orca)
```

See the structure of this object and apply {\tt print()} method on it, too. Look at the results; what do you find?

```{r p7_5, eval=TRUE,echo=TRUE,warnings=FALSE}
s.all
str(s.all)
```

The \texttt{summary} method for a  \texttt{survfit} object would return a lengthy life table. However, the \texttt{plot} method with default arguments offers the Kaplan--Meier curve
for a conventional illustration of the survival experience in the whole patient group.

Alternatively, instead of graphing survival proportions,
one can draw a curve describing their complements: the cumulative mortality proportions. This curve is drawn together with the survival curve as the
 result of the second command line below.

```{r p7_6, eval=TRUE,echo=TRUE,warnings=FALSE}
plot(s.all)
lines(s.all, fun = "event", mark.time=F, conf.int=F)
```

The effect of option {\tt mark.time=F} is to omit
marking the times when censorings occurred.


## Total mortality by stage

Tumour stage is an important prognostic factor in cancer survival studies.

Plot separate cumulative mortality curves for the different stage groups
marking them with different colours, the order which you may define yourself.
Also find the median survival time for each stage.

```{r p7_7, eval=TRUE,echo=TRUE,warnings=FALSE}
s.stg  <- survfit(suob ~ stage, data= orca)
col5 <- c("green", "blue", "black", "red", "gray")
plot(s.stg, col= col5, fun="event", mark.time=F )
s.stg
```



Create now two parallel plots of which the first one describes the cumulative hazards and the second one graphs the log-cumulative hazards against log-time
for the different stages. Compare the two presentations
with each other and with the one in the previous item.

```{r p7_8, eval=TRUE,echo=TRUE,warnings=FALSE}
par(mfrow=c(1,2))
plot(s.stg, col= col5, fun="cumhaz", main="cum. hazards" )
plot(s.stg, col= col5, fun="cloglog", main = "cloglog: log cum.haz"  )
```


If the survival times were {\it exponentially}
 distributed in a given (sub)population
 the corresponding cloglog-curve should follow an approximately linear pattern.
Could this be the case here in the different stages?

Also, if the survival distributions of the different subpopulations would
obey the {\it proportional hazards} model, the vertical distance between the
cloglog-curves should be approximately constant over the time axis.
Do these curves indicate serious deviation from the proportional hazards assumption?

it has been observed that
the crude contrast between males and females in total mortality appears unclear, but the age-adjustment in the Cox model provided a more expected hazard ratio estimate.
We shall examine the confounding by age somewhat closer.
First categorize the continuous age variable
into, say, three categories by function {\tt cut()}
using suitable breakpoints, like 55 and 75 years, and
cross-tabulate sex and age group:

```{r p7_9, eval=TRUE,echo=TRUE,warnings=FALSE}

orca$agegr <- cut(orca$age, br=c(0,55,75, 95))
stat.table( list( sex, agegr), list( count(), percent(agegr) ),margins=T, data = orca )
```

Male patients are clearly younger than females in these data.

Now, plot Kaplan--Meier curves jointly classified by sex and age.

```{r p7_10, eval=TRUE,echo=TRUE,warnings=FALSE}
s.agrx <- survfit(suob ~ agegr + sex, data=orca)
par(mfrow=c(1,1))
plot(s.agrx, fun="event", mark.time=F, xlim = c(0,15),
          col=rep(c("red", "blue"),3), lty=c(2,2, 1,1, 5,5))
```

In each ageband the mortality curve for males is on a higher level than that for females.

## Lexis object with multi-state set-up

Before entering to analyses of cause-specific mortality it might be instructive to apply some Lexis tools
to illustrate the competing-risks set-up.

Form a \texttt{Lexis} object from the data frame and
print a summary of it. We shall name the main (and only) time axis in this object as {\tt stime}.
```{r p7_11, eval=TRUE,echo=TRUE,warnings=FALSE}
 orca.lex <- Lexis(exit = list(stime = time),                  exit.status = factor(event,
    labels = c("Alive", "Oral ca. death", "Other death")),
                   data = orca)
 summary(orca.lex)
```


Draw a box diagram of the two-state set-up of competing transitions. Run first the following command line

```{r p7_12, eval=TRUE,echo=TRUE,warnings=FALSE}
boxes( orca.lex,boxpos=T )
```

Now, move the cursor to the point in the graphics window, at which you wish to put the box for ``Alive'', and click. Next, move the cursor to the point at which you wish to have the  box for ``Oral ca. death'', and click. Finally, do the same with the box for ``Other death''.
If you are not happy with the outcome, run the command line again and repeat the necessary mouse moves and clicks.

## Event-specific cumulative mortality curves

We move on to analysing cumulative mortalities for the
 two causes of death separately, first overall and then
 by prognostic factors.

 Use function \texttt{Cuminc()} in package \texttt{mstate}
 and view the structure of the thus created object.

```{r p7_13, eval=TRUE,echo=TRUE,warnings=FALSE}
cif1 <- Cuminc( time = "time", status= "event", data = orca)
str(cif1)
```


Function \texttt{Cuminc()} thus creates an ordinary data frame with quite self-explanatory column names. Unfortunately, no handy \texttt{plot} method is provided in the package, but in Epi package there is funciont plotCIF



Draw two parallel plots describing
the overall cumulative incidence curves for both causes of death

```{r p7_14, eval=TRUE,echo=TRUE,warnings=FALSE}
par(mfrow=c(1,2))
cif1 <- survfit( Surv( time, event, type="mstate") ~ 1,
                 data = orca.lex)
plotCIF(cif1,event=1)
plotCIF(cif1,event=2)
```


Compute the estimated cumulative incidences by stage for both causes of death.
Now you have to add argument \texttt{stage} when calling
\texttt{Cuminc()}.
See the structure of the resulting object, in which you should
observe the first column containing the grouping variable.
Plot the pertinent curves in two parallel graphs.
Cut the $y$-axis for a more efficient graphical presentation

```{r p7_15, eval=TRUE,echo=TRUE,warnings=FALSE}
par(mfrow=c(1,2))
cif2 <- survfit( Surv( time, event, type="mstate") ~ stage,
                 data = orca.lex)
 plotCIF(cif2, 1, main = "Cancer death by stage",
    col=col5, ylim = c(0, 0.7) )
 plotCIF(cif2, 2, main= "Other deaths by stage",
    col=col5, ylim = c(0, 0.7) )
```

Compare the two plots. What would you conclude about the
effect of stage on the two causes of death?

```{r p7_16, eval=TRUE,echo=TRUE,warnings=FALSE}
par(mfrow=c(1,2))
cif2 <- survfit( Surv( time, event, type="mstate") ~ stage,
                 data = orca.lex)
 plotCIF(cif2, 1, main = "Cancer death by stage",
    col=col5, ylim = c(0, 0.7) )
 plotCIF(cif2, 2, main= "Other deaths by stage",
    col=col5, ylim = c(0, 0.7) )
```

