---
title: 
- Statistical Methods in Cancer Epidemiology using R
author: 
- \textbf{Janne Pitkäniemi}  \newline \newline
- Faculty of Social Sciences, University of Tampere  \newline
- Finnish Cancer Registry  \newline  \newline
- Lecture 2b
date: "Feb,17 2020 "
institute: "janne.pitkaniemi@cancer.fi"
output:
  beamer_presentation:
    includes:  
         in_header: beamer_header.tex
---

# Basic analysis of rates

- Person-time data, hazard and incidence rates, 
- Comparative parameters of rates and their estimation, 
- Poisson regression models and comparative parameters, 
- Adjustment for confounding and evaluation of modification by Poisson regression,
- Goodness-of-fit evaluation.

Main R functions covered:

- `glm()` 
- tools for extracting results from a `glm` model object

# Person-time data and incidence rates

Summarized data on
outcome from cohort study, in which two exposure groups, as to binary risk factor $X$, have been followed-up over individually variable times.
\begin{center}
\begin{tabular}{ccc}
Exposure to         &     Number of       &    Person-\\ 
 risk factor        &        cases        &        time\\
      yes        &                        $D_{1}$ &        $Y_{1}$  \\
      no         &                        $D_{0}$  &        $Y_{0}$\\
     total        &                        $D_{\+}$  &  $Y_{\+}$\\
\end{tabular}
\end{center}

Empirical \textbf{incidence rates} by exposure group:
$$ I_1 = {D_1}/{Y_1}, \qquad
    I_0 = {D_0}/{Y_0}. $$ 
These provide estimates for the true {\bf hazards} 
(or hazard rates) $\lambda_1$ and $\lambda_0$ assumed constant within exposure categories.

# Hazards and their comparison

Parameters of interest:

- **hazard ratio**
$$ \rho = \frac{\lambda_1}{\lambda_0} = 
\frac{ \mbox{hazard among exposed} } { \mbox{hazard among unexposed} }. $$

- **hazard difference**
$$\delta = \lambda_1 - \lambda_0$$

Null hypothesis $H_0: \rho = 1 \Leftrightarrow \delta = 0$ $\Leftrightarrow$  exposure has no effect.


# Estimation of hazard ratio

Point estimator of true hazard ratio $\rho$:
empirical **incidence rate ratio** (IR)
$$  \widehat\rho = \mbox{IR}  =  \frac{I_1}{I_0}
    = \frac{ D_{1} / Y_{1} }{ D_{0} / Y_{0} } 
    = \frac{ D_{1} / D_{0} }{ Y_{1} / Y_{0} }. $$
**NB.**  The last form is particularly useful = **exposure odds ratio** (EOR).

Standard error of log(IR),  95% {error factor} & 95% CI for $\rho$: \newline

$$ SEL  =  \sqrt{ \frac{1}{D_1} + \frac{1}{D_0} } $$
$$EF  =  \exp \{ 1.96 \times \SEL \} $$

$$CI  =  [ \mbox{IR} / \EF ,\ \mbox{IR} \times \EF ].$$ 


**NB.** Random error depends inversely on numbers of cases.

# Estimation of hazard difference

Point estimator of true hazard difference $\delta$:
empirical **incidence rate difference** (ID)
$$  \widehat\delta = \mbox{ID}  =  I_1 - I_0
    = \frac{ D_{1}}{ Y_{1} } - \frac{ D_{0}}{ Y_{0} } 
$$

Standard error of ID,  95% error margin & 95% CI

$$\SE  =  \sqrt{ \frac{I_1^2}{D_1} + \frac{I_0^2}{D_0} } $$
 
$$ \EM  =  1.96 \times \SE $$


$$ \CI  =  [ \mbox{ID} - \EM ,\ \mbox{ID} + \EM ]$$
NB. Random error again depends inversely on no. of cases.

# Example. British doctors' study (Doll & Hill 1966) 
 
CHD mortality in males
by smoking and age.\\  Cases ($D$),  person-years ($Y$), and mortality rates ($I$ per 
$10^4$ y).
\begin{center}
\begin{tabular}{@{\extracolsep{6pt}}lrrrrrr}
\multicolumn{1}{c}{ } & 
\multicolumn{3}{c} {Smokers}         &        
\multicolumn{3}{c} {Non-smokers}  \\
Age(y) & $D$ & $Y$ &    $I$ & $D$ & $Y$ &  $I$ \\ \hline
35-44  & 32  & 52407  & 6   & 2   & 18790 & 1 \\
45-54  & 104 & 43248  & 24  & 12  & 10673 & 11 \\
55-64  & 206 & 28612  & 72  & 28  & 5710  & 49 \\
65-74  & 186 & 12663  & 147 & 28  & 2585  & 108 \\
75-84  & 102 & 5317   & 192 & 31  & 1462  & 212 \\ \hline
Total  & 630 & 142247 & 44  & 101 & 39220 &  26 \\ \hline \end{tabular}
\end{center}


# Example (cont'd). 

Crude incidence rates: \newline
$\qquad I_1 = 630/142247 \ \mbox{y} = 44.3$ per $10^4$ y, and
$\qquad I_0 = 101/39220 \ \mbox{y}\ = 25.8$ per $10^4$ y. 

Crude estimate of overall hazard ratio $\rho$ with SE, etc.

$$\widehat\rho   =  \mbox{IR} = \frac{ 44.3 }{ 25.8 } = {\bf 1.72} $$

$$\SEL  =  \sqrt{ \frac{1}{630} +  \frac{1}{101} } = {\bf 0.1072} $$
$$\EF  =  \exp(1.96\times 0.1072) =  {\bf 1.23}$$


95% CI for $\rho$:

$$ [ 1.72/1.23, \ 1.72\times 1.23 ]= [{\bf 1. 39},\ {\bf 2.12}] $$  

Two-tailed P < 0.001 .

# Poisson regression model for rate ratio
- *Random part*: Number of cases in exposure group $j = 0,1$
$$ D_j \sim \mbox{Poisson}(\lambda_j Y_j) , $$
where $\mu_j = \lambda_j Y_j$ = *expected number* of cases.

- *Systematic part & link function*: \newline linear predictor $\alpha + \beta X_j$ with *logarithmic* (log) link 
$$ \log(\lambda_j) = \alpha + \beta X_j , $$
equivalently on the original hazard scale:
$$ \lambda_j = \exp(\alpha + \beta X_j) . $$

# Poisson model for rate ratio (cont'd)

Interpretation, \newline

- $X_j=$ $\begin{cases}
		1  & \text{if exposed } (j=1), \\
            0  & \text{if unexposed } (j=0),\\
		\end{cases}$
		
- $\alpha=$  $\log(\lambda_0)$, log-baseline rate,

- $\beta=$ $\log(\rho) = \log(\lambda_1/\lambda_0)$, 
                logarithm of true hazard ratio,
- $e^{\beta}=$ $\rho$ = true hazard ratio.

Special case of generalized linear models!


# Example.  Crude analysis of CHD mortality in R

A ready data frame contains

- four variables: 
   - `age` = age group -- a factor with 5 levels,
   - `smok` = smoking: 1 = `yes`, 0 = `no`,
   - `d` = number of cases, 
   - `y` = person-years.
      
- 10 observations (one for each age-smoking combination).

# Example.  Analysis of CHD rates (cont'd)

```{r, comment=NA,echo=FALSE}

bd<-read.table(textConnection("35-44    1  32 52407   6.1
 35-44    0   2 18790   1.1
 45-54    1 104 43248  24.0
 45-54    0  12 10673  11.2
 55-64    1 206 28612  72.0
 55-64    0  28  5710  49.0
 65-74    1 186 12663 146.9
 65-74    0  28  2585 108.3
 75-84    1 102  5317 191.8
 75-84    0  31  1462 212.0"))
colnames(bd) <- c("age", "smok", "d", "y", "rate")
bd
```

# Fitting Poisson model for crude rate ratio


Poisson model with log-link (default) for crude rates
\scriptsize
```{r, comment=NA,echo=FALSE,warning=FALSE}
m1 <- glm(d/y ~ smok, family=poisson(), weights=y, data=bd) 
coef(summary(m1))
```
\normalsize

# Fitting crude rate ratio model (cont'd)

Main results: 

- $\widehat\alpha =  -5.96 = \log(25.8/10^4 \mbox{ y}), \quad$ (SE = 0.10),

- $\widehat\beta = 0.54 = \log(1.72), \quad$ (SE = 0.11)

Function `ci.lin()` transforms results to ratio scale

```{r, comment=NA,echo=FALSE,warning=FALSE}
library(Epi)
round(ci.lin(m1, Exp=T)[, -(3:4)], 4 )
```
Compare the results with those obtained above using simple
estimation & SE formulas.

# Fitting crude rate ratio model (cont'd)

The Poisson model above can also be fitted as follows:
  
 glm(d ~ smok, fam =poisson(), offset=log(y))

Here `offset` refers to the logarithm of person-years `y` in formula for expected 
numbers of cases $\mu_j = \lambda_j \times Y_j$:
$$ 
\log(\mu_j)  =  \log(\lambda_j Y_j) 
             =   \log(Y_j) + \log(\lambda_j)  
             =  \log(Y_j) + \alpha + \beta X_j , 
$$

$\log(Y_j)$ is an **offset** term in the linear predictor, meaning that it has a fixed value 1 for the regression coefficient. 

# Stratified analysis

Stratification of cohort data with person-time \newline -- 
at each level $k$ of covariate $Z$ results are summarized:

\begin{center}
\begin{tabular}{ccc}
Exposure to & Number of         & Person- \\
risk factor & cases       & time\\ \hline
 yes        & $D_{1k}$    & $Y_{1k}$ \\
 no         & $D_{0k}$    & $Y_{0k}$ \\ \hline
Total       & $D_{\+ k}$  & $Y_{\+ k}$ \\
\end{tabular}
\end{center}

Stratum-specific rates by exposure group: 
$$ I_{1k}  = \frac{ D_{1k} }{ Y_{1k} } , \qquad
   I_{0k}  = \frac{ D_{0k} }{ Y_{0k} } .$$


# Stratum-specific comparisons

Let $\lambda_{jk}$ be true rate for exposure group $j$ ($j=0,1)$\newline 
and stratum $k$ ($k=0, \dots, K$). 
Let also
$$ \rho_k = \frac{ \lambda_{1k} } { \lambda_{0k} }, \qquad \delta_k = \lambda_{1k} - \lambda_{0k}$$

be the rate ratios and rate differences between the exposure groups in stratum $k$.

Two simple models assuming homogeneity: 

- common rate ratio: $\rho_k = \rho$ for all $k$,
- common rate difference: $\delta_k = \delta$ for all $k$.

Only one of these can in principle hold.  However, almost always neither homogeneity assumption is exactly true.   

# Example. British male doctors (cont'd)


CHD mortality rates (per $10^4$ y) and numbers of cases ($D$)
by age and cigarette smoking.

Mortality rate differences (ID) and ratios (IR) in age strata.

\scriptsize
```{r, comment=NA,echo=F,warning=FALSE}
dbs<-bd
dbs$ID<-0.0
dbs$ID[dbs$smok==1]<-dbs$rate[dbs$smok==1]-dbs$rate[dbs$smok==0]
dbs$IR<-1.0
dbs$IR[dbs$smok==1]<-round(dbs$rate[dbs$smok==1]/dbs$rate[dbs$smok==0],1)
tot1<-c("35-84",1, sum(dbs$d[dbs$smok==1]), sum(dbs$y[dbs$smok==1]),round(10000*(sum(dbs$d[dbs$smok==1])/sum(dbs$y[dbs$smok==1]))),0.0,1.0)
tot0<-c("35-84",0, sum(dbs$d[dbs$smok==0]), sum(dbs$y[dbs$smok==0]),round(10000*(sum(dbs$d[dbs$smok==0])/sum(dbs$y[dbs$smok==0]))),0.0,1.0)
dbs<-rbind(dbs,tot1,tot0)
dbs[order(dbs$age,dbs$smok),]
```
\normalsize

# Example (cont'd).

-Both types of comparative parameter, rate ratios $\rho_k$ and rate differences $\delta_k$ appear heterogenous, because

- ID increases by age -- at least up to 75 y,
-  IR decreases by age.

- Part of this observed heterogeneity may be due to random variation. 
- Yet, any single-parameter comparison by common rate ratio or rate difference  

may not adequately capture the
joint pattern of true rates.


$\quad \Rightarrow$ Effect modification must be evaluated.

# Rate ratio adjustment by Poisson model

Define Poisson regression model for

- one binary exposure variable $X$ and 
- one categorical (polytomous)  factor $Z$.
   - *Random part*: No. of cases in exposure group $j$ $(j=0,1)$
and covariate level $k$ $(k = 1, \dots, K)$ is 
$D_{jk} \sim Poisson( \lambda_{jk} Y_{jk} )$ 
   - *Systematic part*: $log(\lambda_{jk}) = \alpha + \beta X_j + \gamma_k,$ where $X_j$ is (0/1)


- $\alpha=\log(\lambda_{01})$ = log-baseline rate, 
- $\gamma_k=$ $\log(\lambda_{jk}/\lambda_{j1})$,
- $\beta=$ $\log(\rho)=log(\lambda_{1k}/\lambda_{0k})$,
- $e^{\beta}=$ $\rho$ = true rate ratio for the effect of exposure to $X$.

*How do we read this?*

# Implications of model definition


- homogeneity of true rate ratio $\rho_k = \rho$ for $X$ \newline
 across levels of $Z$ is 	assumed,
-  inclusion of $Z$ leads to adjustment for $Z$ in estimating the common effect of $X$,
-  $e^{\gamma_k}$ = rate ratio for level $k$ of $Z$ *vs.* level 1
      is the same in both exposure groups ($j=0,1$) \newline
      $\Rightarrow$ homogeneity of
      the effect of $Z$ is assumed, too.
     
- level $k=1$ is chosen as the *reference* level for $Z$\newline (like 
	``unexposed'' is reference for $X$), 
- before model fitting, binary *indicator* variables $Z_k$ 
	for levels $k=1, \dots, K$ of $Z$ must be defined:
	$$ Z_k  =\begin{cases}
		   1, & \mbox{if observation belongs to level } k, \\
               0, & \mbox{ otherwise}.
		   \end{cases}
	$$

# Example.  CHD in British doctors (cont'd)

Factor `age` has $5$ levels. \newline
Indicator variables for each age level are generated in R when defining the model, and the following model matrix is returned.

\small

```{r, comment=NA,echo=T,warning=FALSE}
 m2 <- glm( d/y ~ age + smok,family=poisson(link=log),
            weights=y,data=bd)
cbind(data.frame(bd$age), model.matrix(m2))
```
\normalsize

# Summary of the adjustment model

\scriptsize
```{r, comment=NA,echo=T,warning=FALSE}
 coef(summary(m2))
```


\normalsize

# Fitting adjustment model (cont'd)

Results on the ratio scale

```{r, comment=NA,echo=T,warning=FALSE}
round(ci.lin(m2, Exp=T)[, 5:7], 4 )
```

$\Rightarrow$ Age-adjusted rate ratio [95\% CI] for smoking:
$$\widehat \rho = {\bf 1.43} \ [{\bf 1.16}, {\bf 1.76}]$$

# Fitted values & residuals

From the estimated coefficients we can calculate \newline
 **fitted**
 linear predictors $\widehat \eta_{jk}$, hazards $\widehat \lambda_{jk}$ 
and numbers $\widehat \mu_{jk}$:
$$ \widehat \eta_{jk}  =  \widehat\alpha + \widehat\beta x + \widehat\gamma_k $$
$$ \widehat \lambda_{jk}  =  \mbox{exp}(\widehat \eta_{jk}), \quad \widehat\mu_{jk} = \widehat\lambda_{jk} Y_{jk} $$

In R the two first can be extracted directly from the\\ fitted model object `m2`:
\small

```{r, comment=NA,echo=T,warning=FALSE}
  bd$fit.eta <- m2$linear.predictor
  bd$fit.rate <- fitted(m2); 
  bd$fit.mu <- bd$y*bd$fit.rate
```

\normalsize

**NB**.  If count `d` is declared as response with ` log(y)` as offset, then `fitted()`
returns the fitted numbers of cases $\widehat \mu_{jk}$.

# Fitted values \& residuals (cont'd)

**Deviance residual** for cell $jk$ (`resid(m2)` in R):
$$ d_{jk}  =  \mbox{sign}(Y_{jk} - \widehat\mu_{jk}) 
                    \times 
                   \sqrt{ 2 \left\lbrace
                         Y_{jk} \log \left( \frac{ Y_{jk} }{ \widehat\mu_{jk} } \right)
                         - (Y_{jk} - \widehat\mu_{jk}) \right\rbrace } 
$$

**Pearson residual** (`resid(m2, type="pearson")`):
$$ r_{jk} = \frac{ Y_{jk} -  \widehat\mu_{jk} }
          {\sqrt{  \widehat\mu_{jk} } } . $$
Small value of either residual\newline  
$\rightarrow$ consistency of observation with model.

``Large'' (in absolute value) residual\newline
$\rightarrow$ lack of fit for that cell.

# Example.  Fitted values \& residuals
 
\scriptsize

```{r, comment=NA,echo=T,warning=FALSE}
bd$fit.rate<-round(10000*bd$fit.rate,1)
bd$fit.mu<-round(bd$fit.mu,1)
bd$res.D<-round(resid(m2,"dev"),2)
bd$res.P<-round(resid(m2,"pear"),2)
data.frame(bd$age,bd$smok,bd$d,bd$rate,bd$fit.rate,bd$fit.mu,bd$res.D,bd$res.P)
```
NB! Fitted rate ratios between smokers and non-smokers:
$$ \frac{5.2}{3.6} = \dots = \frac{209.7}{147.1} = 1.43 \ \text{at each age level}$$

# Summary goodness-of-fit statistics

- **Deviance** (Dev)  
         = sum of squares of deviance residuals,
-  **Pearson's** $X^2$ (chi-square) 
          = s.s. of Pearson residuals  

Assume that 

- the model is "true", and 
- the data are not *sparse*, i.e. numbers of cases in each line of the (grouped) data matrix are sufficiently big.

Under these assumptions 

- residuals $d_{jk}$ and $r_{jk}$ all approximately Normal(0,1),
- Both Dev and $X^2$ are approximately 
       $\chi^2$-distributed with 
       *degrees of freedom* (df) being equal to the \newline
       *residual* df = no. of obs -- no. of coeff's 
- Expected value of these statistics 
      = residual df.

# Goodness of fit (cont'd)

In our example 

-  "large" residuals in extreme age groups,
- $X^2$ and Dev 2.8 to 3 times higher than df.

$\rightarrow$ Some lack of fit with assumed model.

($P$-values: P($X^2$ $>$ 11.2) = 0.025, and \newline
  P(Dev $>$ 12.1) = 0.017)
  
#  Possible causes for lack of fit

- Unrealistic random part (Poisson)?
- Misspecified systematic part?
      - Wrong functional form for effects? \medskip
      - Missing important risk factors? \medskip
      - Failure to take into account *effect modification* or "interaction".
    
In this example the last possibility is very likely:  \newline 
the rate ratios are clearly heterogenous. 

Yet, inclusion of 4 interaction parameters
would lead to a saturated model.

# Continuous covariates
 
Stratified analysis is adequate with

- qualitative exposure factor with few levels,
- few qualitative covariates with few levels.

Effects of continous factors can often be 
described by smooth functions with small number of parameters.

Stratification requires categorization of 
continuous variables and level specific parameters 

$\Rightarrow$ loss of efficiency.

# Continuous covariates (cont'd)

Regression models can accommodate functions of 
continuous variables 

- (+) more parsimonious models, \medskip 
- (+) greater flexibility in modelling, \medskip
- (-) more opportunities for misspecification,
*i.e.* creation of "wrong" 
models.

# Example (cont'd)

Treat age as continuous variable taking midpoint of each ageband
(40, 50, $\dots$, 80)
as quantitative age value. 

Include linear and quadratic term of age applying 

- centering at 60 years, and 
- scaling, *i.e.* division by 10 years

This is achieved by

\small
```{r, comment=NA,echo=T,warning=FALSE}
bd$A.L <- as.numeric(bd$age)-3 
bd$A.Q <- bd$A.L*bd$A.L - 2
```
\normalsize   

# Example (cont'd)}
 
Model with linear and quadratic terms for age, and
interaction between linear term of age and smoking:
\small

```{r, comment=NA,echo=T,warning=FALSE}
 m3 <- glm( d/y ~ A.L + A.Q + smok + A.L:smok,
 family=poisson( ), weights=y,data=bd )
round(ci.lin(m3, Exp=T)[, -(3:4)], 4 )
```
\normalsize
Interpretation of parameters?

# Example (cont'd)

Observed and fitted rates \& fitted 
rate ratios between smokers and non-smokers
in each ageband
\scriptsize
```{r, comment=NA,echo=F,warning=FALSE}
  bd$fit.eta <- m3$linear.predictor
  bd$fit.rate <- fitted(m3); 
  bd$fit.mu <- bd$y*bd$fit.rate

bd$fit.rate<-round(10000*bd$fit.rate,1)
bd$fit.mu<-round(bd$fit.mu,1)
bd$res.D<-round(resid(m3,"dev"),2)
bd$res.P<-round(resid(m3,"pear"),2)
data.frame(bd$age,bd$smok,bd$d,bd$rate,bd$fit.rate,bd$fit.mu,bd$res.D,bd$res.P)
```
\normalsize
Residual Dev = 1.64  df = 5; The fit is excellent! 

# Some closing remarks

- **Hazard differences** can also be estimated by
Poisson-modelling, because `link='identity'`
can be coupled with `family=poisson`  in `glm()`.
- Poisson models can also be fitted on **ungrouped data**. Units of observations may even be shorter intervals of total individual follow-up times: 
- Each member of the study population can contribute several **separate time-slots** as observational units. 
- More about **time-splitting** in the next lectures.
- With continuous covariates it may be difficult to keep
hazards positive when the *additive hazards*
model with `link='identity'` is attempted to be fitted.
