---
title: 
- Statistical Methods in Cancer Epidemiology \newline using R
author: 
- \textbf{Karri Seppä}  \newline \newline
- Finnish Cancer Registry  \newline  \newline
- Lecture 6
date: "Mar 9, 2020"
institute: "karri.seppa@cancer.fi"
output:
  beamer_presentation:
    includes:  
         in_header: beamer_header.tex
tables: true
editor_options: 
  chunk_output_type: console
---



## Contents

1. Case-control designs 
\medskip
2. Exposure odds ratio (EOR)
  and its interpretation
\medskip
3. Estimation of EOR by logistic regression
\medskip
4. Matched case-control studies
\bigskip

Main R functions to be covered

- \texttt{glm()}

- \texttt{clogit()} and \texttt{cch()} in practical 6


## Case-control design

- From given study  population (base pop'n) are selected 
all or a random sample of

    - $D$ \textbf{cases}, or individuals with the \newline
  disease being diagnosed during certain period
\medskip
    - $C$ \textbf{controls}, or ``healthy'' individuals at risk.


\medskip
- Exposure to risk factor $X$ and other covariates \newline
  assessed in cases and chosen controls.
 \medskip
- To increase efficiency and remove confounding,
the sampling of controls is often \textit{stratified} or individually \textit{matched} for age, gender,
place of residence, \textit{etc}.


## Exposure odds ratio (EOR)

With binary risk factor $X$ the results are summarized:

\begin{center}
\begin{tabular}{lccc}
\toprule
Exposure         &        Cases        &    Controls & Total\\
\midrule
yes $(X=1)$        &        $D_{1}$ &        $C_{1}$  & $T_1$  \\
no $(X=0)$         &        $D_{0}$  &        $C_{0}$  & $T_0$  \\
\midrule  
Total        &                $D $  &  $C$  & $T $ \\
\bottomrule
\end{tabular}
\end{center}

Common effect measure:

- \textbf{exposure odds ratio}
$$  
      \text{EOR}  
          =  \frac{ D_{1} / D_{0} } { C_{1} / C_{0} }  
          = \frac{ D_{1} C_{0} } { D_{0} C_{1} }        
$$


## Precision in EOR

Standard error of log(EOR),  95% error factor (EF) & \newline
 95% confidence interval (CI) for 
the associated parameter:
\begin{align*}
 \text{SEL}  = & \sqrt{ \frac{1}{D_1} + \frac{1}{D_0} 
    + \frac{1}{C_1} + \frac{1}{C_0} } \\ & & \\
 \text{EF}  = & \exp \{ 1.96 \times \text{SEL} \} \\ & & \\
 \text{CI}  = & [ \text{EOR} / \text{EF} ,\ \text{EOR} \times \text{EF} ] . 
\end{align*}
NB. Random error depends inversely on numbers of cases and controls.



## What parameter is estimated by EOR?

The answer depends on 

- type of base population, from which cases emerge

    - closed population or \textbf{cohort}, or 
    - open or \textbf{dynamic population},

- time dimensionality  \newline
-- longitudinal or cross-sectional
\medskip
- sampling principle of controls:

    (A) case-noncase sampling
 (epidemic ca-co study)
\medskip 
    (B) case-cohort sampling
\medskip
    (C) density sampling (incl. nested case-control study)



## Sampling controls from a longitudinal base

\pause
Simplified ideal situation:\newline
 Complete follow-up 
of a cohort of initially healthy subjects with 
no losses during a fixed risk period.

\pause
\begin{center}
{\normalsize
\setlength{\unitlength}{0.08cm}
\begin{picture}(135,50)
\thicklines
\linethickness{0.3mm}

%mittlere Box

\put(40,45){\line(6,-1){55}}
\put(40,5){\framebox(55,40){}}

%Achsenbeschriftung
\put(45,-1.5){\makebox(25,5){\small Time ($t$)}}
\put(25,-1){\makebox(25,5){\small Start}}
\put(85,-1){\makebox(25,5){\small End}}
\put(70,2.5){\vector(1,0){20}}

\pause
%linker Text
\put(0,5){\makebox(40,40){\parbox{7cm}{\begin{center}
						   {\small{\bf B}:  Initially at risk}\\
						%	(disease-free)\\ (C)\\
							{\small $(N)$}
							\end{center} }}}
\pause
\put(40,15){\makebox(55,10){\parbox{11cm}{ \begin{center}
						{\small {\bf C}: Currently at risk $(N_t)$}
							\end{center} }}}
\put(67.5,25){\vector(0,1){15}}
\put(67.5,15){\vector(0,-1){9}}							
%rechter Text
\pause
\put(100,30){\makebox(40,15){\parbox{7cm}{\begin{center}
					{\small		New cases\\of disease\\ $(D)$}
							\end{center} }}}
\pause							
\put(100,5){\makebox(40,20){\shortstack{\strut\small
       {\bf A}: Still at risk \\ \strut
						%	(disease-free)\\ \strut 	(A)\\
						 \strut $(N - D)$ }}}


\end{picture}
}
\end{center}

Possible sampling frames: \textbf{A}, \textbf{B} and \textbf{C}


## Sampling schemes or designs for controls

\pause
\textbf{A:  Case-noncase sampling}
\pause

- Controls chosen from those $N-D$ 
subjects % cohort members
still at risk (healthy) 
\underline{\textit{at the end}}
of the follow-up. 

\pause   	     
\textbf{B:  Case-cohort sampling}: 
\pause

- The control group or \textbf{subcohort}   
           is a random sample of the whole cohort $(N)$  
				    \underline{\textit{at the start}} of the follow-up.

\pause           	     
\textbf{C:  Density sampling}: 
\pause 
         
- Controls drawn 
            \underline{\textit{during the follow-up}}
              from those \newline currently at risk. 
           \medskip
- \textbf{Nested case-control design (NCC)}\newline
         A set of controls is sampled from
         the \textit{risk set} 
			  \underline{\textit{at each time $t$ of diagnosis}}
                           of a new case.


## EOR in case-noncase sampling design

- In the traditional or epidemic case-control study
the controls are selected from
those still healthy at the end of the
risk period, during which cases are collected.
\medskip
\pause
- In this design EOR estimates the \textbf{risk odds ratio} 
$$ 
\psi = \frac{ \text{odds of dis. in the exp'd} }
                     {   \text{odds of dis. in the unexp'd} } 
     = \frac{ \pi_1/(1-\pi_1) }{ \pi_0/(1-\pi_0) }, 
$$
  where $\pi_1$ and $\pi_0$ are the risks of disease in the exposed and unexposed groups, estimable from a corresponding cohort study by
  incidence proportions $R_1$ and $R_0$. 
\pause
\medskip 
- \textbf{NB.}  $\psi \approx \pi_1 / \pi_0 = \phi$, \textit{i.e.} close to \textbf{risk ratio}, when risks $\pi_1$ and $\pi_0$ are low % ($< 0.1$) 
\pause
= the ``rare disease assumption''.              


## Density sampling

- New incident cases occurring during given study period 
      are identified from the base population.
\medskip      
- controls are randomly chosen from the population  
      at risk at various times in the period 
      (sometimes only once).
\medskip      
- For chronic disease studies this design is the most popular,
- Logically the only possibility in open populations,

\textbf{Nested case-control study}: \textit{time-matched} selection:

- one or more (rarely over $5$)
 controls chosen from the \newline population at risk 
 at each time $t_d$ when a new case is diagnosed. 
 
 
## EOR in density sampling

- In a full cohort
study the true hazard ratio
$\rho = \lambda_1 / \lambda_0$ 
is estimated by the incidence rate ratio 
$$  
\text{IR} = \frac{I_1} {I_0} = \frac{D_1 / D_0 }{ Y_1 / Y_0} . 
$$
- In a case-control study with density sampling 
the \textbf{exposure odds} among controls
$C_1/C_0$ estimates the exposure odds $Y_1/Y_0$, i.e.
the distribution of  person-years 
 in the base population.
\medskip
- Thus, the exposure odds ratio EOR
$$ 
\text{EOR} = \frac{D_1 / D_0 }{ C_1 / C_0} \approx 
  \frac{D_1 / D_0 }{ Y_1 / Y_0} = \text{IR} 
$$
is a consistent estimator of the true hazard rate ratio
$\rho$ without any rare-disease assumption.


## Example. Alcohol use and oesophageal cancer

 (Tuyns et al 1977, see \textbf{B\&D} 1980).

- 205 new cases of cancer identified in a French province 
during two years, 
and 770 randomly sampled population
controls \newline $\Rightarrow$ Density sampling
- \textbf{NB.} No stratification or matching for age in design \newline
$\Rightarrow$
Too many young controls in relation to few cases \newline
$\Rightarrow$  inefficient! 
\medskip
- Exposure of interest: Daily consumption of alcohol.
\medskip
- In the following table the data are summarized by dichotomized 
exposure and stratified by age group.
\medskip
- In R the data are found: \texttt{data(esoph)}


## Example: Results stratified by age
\small
\begin{center}
\begin{tabular}{lcccc}
\toprule
Age & Exposure $\geq 80$ g/d & Cases & Ctrls & EOR \\
\midrule
25-34 & yes         &  1 & 9 & $\infty$  \\
 { } & no                & 0 & 106 & { } \\
\cmidrule{2-5} 
35-44 & yes         &  4 & 26 &  5.05 \\
 { } & no                & 5 & 164 & { } \\
\cmidrule{2-5} 
45-54 & yes         &  25 & 29 & 5.67 \\
 { } & no                & 21 & 138 & { } \\
\cmidrule{2-5} 
55-64 & yes         &  42 & 27 & 6.36 \\
 { } & no                & 34 & 139 & { } \\
\cmidrule{2-5} 
65-74 & yes         &  19 & 18 & 2.58 \\
 { } & no                &  36 & 88 & { } \\
\cmidrule{2-5}
75-84 & yes         &  5 &  0 & $\infty$ \\
 { } & no                & 8 & 31 & { } \\
\bottomrule 
Total & yes   & 96 & 109 & 5.64 \\
{ }  & no    & 104 & 666 & {(crude)} \\
\bottomrule
\end{tabular}
\end{center}


## Example: (cont'd) 

Modification?

- Stratum-specific EOR$_k$s somewhat variable.
\medskip
- Random error in some of them apparently great (especially in the youngest and 
the oldest age groups).

Confounding?

- Is exposure associated with age in the study population?
\medskip 
- Look at variation in the age-specific prevalences of exposure among controls.
\medskip
- Adjustment for age is generally reasonable.


## Model for stratified data

\textit{Random part}: 

Conditional on total number of subjects 
$$
T_{jk} = D_{jk} + C_{jk}
$$ 
in each level $j$ $(j=1, 2)$ of
exposure variable $X$ 
and level $k$ $(k=1,\dots K)$ of
covariate $Z$ we assume 
$$ 
D_{jk} \sim  \mbox{Binomial}(T_{jk};\ p_{jk}) , 
$$
where $p_{jk}$ is the "probability of being a case"
in a group of cases & controls defined by $X$ and $Z$.


## Model for stratified data (cont'd)

\textit{Systematic part \& logit link}:
$$
\text{logit}( p_{jk} )
 = \log \left( \frac{\displaystyle p_{jk}}
  { \displaystyle 1 - p_{jk} } 
   \right ) = \alpha + \beta X + \gamma_k , 
$$

\begin{enumerate}[ ] 
\item $X$ = exposure, 1: 'exposed'; 0: 'unexposed',
\item $\alpha = \text{logit}(p_{11}) = $  
                  log of "pseudo baseline odds",
\item $\beta$ = logarithm of exposure odds ratio,  
\item   $\quad  =    \log(\rho)$, 
     logarithm of true rate ratio $\rho$
\item    \quad\quad\ with density sampling,
\item   $\gamma_k$ = logarithm of 
   rate ratio btw levels $k$ and  1  of  $Z$.
\end{enumerate}

Hence $e^{\beta} = \rho$ is the common rate ratio 
for the exposure effect assumed constant over the levels of $Z$. 
 

 
## Example. Estimation by \texttt{glm()}

Input of data

\footnotesize
```{r, comment=NA,echo=T}
D <- c(0,1, 5,4, 21,25, 34,42, 36,19, 8,5) # no. of cases
C <- c(106,9, 164,26, 138,29, 139,27, 88,18, 31,0) # controls
Tot <- D + C         # cell totals       
```
\normalsize
Generation and naming of the levels for factors describing 
age group and alcohol exposure
\small
```{r, comment=NA,echo=T}
agr <- gl(6,2,12)  # 6 levels for age factor
levels(agr) <- c("25-34", "35-44", "45-54", 
                "55-64", "65-74", "75-84")
alc <- gl(2,1,12)  # 2 levels for alcohol factor
levels(alc) <- c("0-79g/d", "80+g/d")
```
\normalsize

## Example. Estimation by \texttt{glm()}

```{r, comment=NA,echo=T}
data.frame( agrn = as.numeric(agr), agr, 
            alcn = as.numeric(alc), alc, D, C, Tot)
```


## Example. Estimation by \texttt{glm()} (cont'd)

Crude estimation
\small
```{r, comment=NA,echo=T}
library(Epi)
mod1 <- glm( D/Tot ~ alc, fam = binomial, w = Tot)
round(ci.lin(mod1, Exp=T)[ , -(3:4)], 4)
```
\normalsize

Estimation adjusted for age
\small
```{r, comment=NA,echo=T}
mod2 <- update(mod1, . ~ . + agr)
```
\normalsize

Goodness-of-fit evaluation
\small
```{r, comment=NA,echo=T}
c( mod2$deviance, mod2$df.res )
```
\normalsize



## Example. Estimation by \texttt{glm()} (cont'd)

Estimation results after adjusting for age
\small
```{r, comment=NA,echo=T}
round(ci.lin(mod2, Exp=T)[ , -(3:4)], 4)
```



## Matched case-control study

Matching

- For each case choose 
    1 or more (rarely over $4$) controls with
      same age (\textit{eg.} within 1 year, 
      or in the same 5-year ageband), 
      sex, place of living, \textit{etc.} 
      \medskip
- Implies stratification in design: 
each matched case-control set forms one stratum.
    \medskip
- Improves efficiency of the study & estimation of 
effect parameters, if matching factors are 
   strong determinants of outcome.


## Matched case-control study (cont'd)

Some principles

- Impractical to match on many other covariates than those mentioned,
\medskip
- Matching on a correlate $Z$ of risk factor $X$ of interest, which is not 
	causal determinant of outcome \newline
	 $\Rightarrow$ overmatching, loss of efficiency.
	\medskip
- \textit{Counter-matching}: Choose controls which are different from
	case w.r.t. $Z$, close correlate of $X$ \newline
	 $\Rightarrow$ increases efficiency.



## Matched case-control study (cont'd)

- \textit{Matched design $\Rightarrow$ matched analysis!}
\medskip
- Ignoring matching in analysis my lead to biased results.
\medskip
- Matching factors must always be accounted 
for in estimating the rate ratios of interest. 
\medskip
- With very close matching (based \texttt{e.g.} on sibship, neighbourhood)
 use \textit{conditional logistic regression} \newline modelling \newline
 -- function \texttt{clogit()} in package \texttt{survival}

## Full cohort design

Each member of the cohort provides exposure data for all
cases, as long as this member is at risk, i.e. alive, not censored
& free from outcome.

\begin{center}
\includegraphics[height=2in]{lect6_full_cohort}
\end{center}

Times of new cases define the \textbf{risk-sets}. 

- Use e.g. function \texttt{coxph()}


## Nested case-control design

Whenever a new case occurs, a set of controls
(here 2/case) are sampled from its risk set. Implies time-matching at least.


\begin{center}
\includegraphics[height=2in]{lect6_ncc}
\end{center}



\textbf{NB.} A control once selected for some case can be selected as a
control for another case, and can later on become a case, too.

- Analyse using function \texttt{clogit()}

## Case-cohort design

\textbf{Subcohort:} Sample of the whole cohort randomly selected at
the outset. Serves as reference group for all cases.


\begin{center}
\includegraphics[height=2in]{lect6_cch}
\end{center}



\textbf{NB.} A subcohort member can become a case, too.

- Analyse using function \texttt{cch()} 


## Concluding remarks

- Analysis using \texttt{glm()} on individual data records from 
an unmatched study proceeds similarly as for grouped data.
\medskip
- Matched design $\to$ matched analysis by \texttt{clogit()}.
\medskip
- More complicated designs, like counter-matched and two-phase studies,
require specialized methods and progamming.
\medskip
- Case-cohort design: Use function \texttt{cch()} in package \texttt{survival}
that adjusts standard errors \textit{etc.} appropriately.  



